### 2019-07-11 notebook/ghost-fit.Rmd
and so it begins... :camping:

---
title: "greta.Rmd"
author: "Carl Boettiger"
date: "5/30/2019"
output: github_document
---

```{r message = FALSE, warning=FALSE}
library(tidyverse)
library(greta) # remotes::install_github("greta-dev/greta")
set.seed(123456)
```

```{r}
data <- read_csv("../data/reps.csv")
data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.1)
```

```{r}
wide <- data %>% spread(reps, x) %>% select(-t) %>% 
  select(2) %>% # which / how many reps we use
  as.matrix()
n <- dim(wide)[1]

```

```{r}
x_t1 <- wide[-1,] # data$x[-1]
x_t <- wide[-n,] # data$x[-100]
dim(x_t)
dim(x_t1)
#tibble(x_t, x_t1) %>% ggplot(aes(x_t, x_t1)) + geom_point()
```

```{r}
#r <- 0.05 
Q <- 5
#sigma <- 0.02
#a <- 0.0233
K <- uniform(0, 10)
H <- uniform(0,  2)
a <- uniform(0, 1)
r <- uniform(0, 1)
#Q <- uniform(0, 10)
sigma <- uniform(0, 1)

# Model
mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
distribution(x_t1) <- normal(mean, sigma)
m <- model(K, a, H, sigma, r)
```

```{r }
system.time({
  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
})
```

```{r}
summary(draws)
```

```{r}
#bayesplot::mcmc_trace(draws)
```

```{r}
samples <-  
  map_dfr(draws, 
          function(x) data.frame(x, t = 1:dim(x)[1]), 
          .id = "chain") %>% 
  gather(variable, value, -t, -chain)

samples %>%  
  ggplot(aes(t,value, col=chain, group=chain)) + 
  geom_line() +
  facet_wrap(~variable, scales = "free") + 
  scale_color_viridis_d()
```

```{r}
#Q = 5
true <- data.frame(a = 0.023, r = .05, K = 2, H = .38, sigma = .02) %>%
  gather(variable, value)
```

```{r}
samples %>% ggplot() + 
  geom_histogram(aes(value), bins = 30)  +
  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
  facet_wrap(~variable, scales = "free")
```

### 2019-07-11 notebook/ghost-sims.Rmd
and so it begins... :camping:

---
title: "simulation & inference of transients"
author: "Carl Boettiger"
date: "5/30/2019"
output: github_document
---

```{r}
library(tidyverse)
library(nimble)
```

```{r}
#a = 0.023
r = 0.05
p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 1e3)
growth <- function(x, p) x * p$r * (1 - x / p$K)
consumption <- function(x,p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
```

```{r}
theory <- 
  tibble(x= seq(0,2, length.out = 100)) %>%
  mutate(g = growth(x,p), 
         c = consumption(x,p)) %>%
  mutate(potential = - cumsum(g - c)) %>%
  gather(curve, y, -x, -potential) 
```

```{r noisy_switch}
theory %>%
  ggplot(aes(x, y, col=curve)) +
  geom_line(lwd=1)
```

```{r}
theory %>%
  ggplot(aes(x, potential)) + 
  geom_line(lwd=1)
```

```{r}
library(nimble)
library(tidyverse)
# Define stochastic model in BUGS notation
may  <- nimble::nimbleCode({

  x[1] <- x0
  for(t in 1:(N-1)){
    # Determinstic mean looks like standard R
    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ Q / (x[t] ^ Q + H ^ Q)
    # Note the use of ~ in BUGS to show 'distributed as normal' 
    y[t+1] ~ dnorm(mu[t], sd = sigma)
    x[t+1] <- max(y[t+1],0)
  }

})
model <- nimbleModel(may,constants = p, inits = list(x0 = 0.2))
cmodel <- model #compileNimble(model)
set.seed(123456)
simulate(cmodel)
df <- tibble(t = seq_along(cmodel$x), x = cmodel$x)
```

```{r}
df %>% ggplot(aes(t,x)) + geom_point()
df %>% write_csv("../data/ex.csv")
```

Let's do 100 replicates

```{r}
set.seed(123)

df <- map_dfr(1:100, 
  function(i){
    simulate(cmodel)
    tibble(t = seq_along(cmodel$x), x = cmodel$x, reps = i)
   })
```

```{r}
df %>% 
  #filter(t %in% seq(1000, 2000, by = 1)) %>% 
  ggplot(aes(t,x, group=reps)) + geom_line(alpha=.2)

df %>% write_csv("../data/reps.csv")
```

### 2019-07-11 notebook/ghost-fit.Rmd
larger simulation set

@@
 
+Divide this data into several groups based on
+
 ```{r}
-wide <- data %>% spread(reps, x) %>% select(-t) %>% 
-  select(2) %>% # which / how many reps we use
+means <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
+
+lows <- means %>% dplyr::slice(1:10) %>% pull(reps)
+highs <- means %>% dplyr::slice((n()-10):n()) %>% pull(reps)
+coding <- function(reps) case_when(
+  reps %in% lows ~ "low",
+  reps %in% highs ~ "high"
+)
+df <- data %>% mutate(coding = coding(reps)) %>% na.omit()
+df %>% ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)
+
+#means %>% ggplot(aes(ave)) + geom_histogram(binwidth = 0.02)
+
+```
+
+```{r}
+wide <- data %>% 
+  filter(reps %in% highs) %>%
+  spread(reps, x) %>%
+  select(-t) %>% 
   as.matrix()
@@
 ```{r}
-#bayesplot::mcmc_trace(draws)
+bayesplot::mcmc_trace(draws)
 ```

### 2019-07-11 notebook/ghost-sims.Rmd
larger simulation set

@@
 
-```{r}
+```{r message = FALSE}
 library(tidyverse)
+library(furrr)
 library(nimble)
@@
 ```{r}
-#a = 0.023
-r = 0.05
-p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 1e3)
+p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2)
 growth <- function(x, p) x * p$r * (1 - x / p$K)
-consumption <- function(x,p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+consumption <- function(x, p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
 ```
@@
 ```{r}
-library(nimble)
-library(tidyverse)
-# Define stochastic model in BUGS notation
-may  <- nimble::nimbleCode({
+may <- function(x, p) x + growth(x,p) - consumption(x,p)
+#x + x * p$r * (1 - x / p$K)  - p$a * x ^ p$Q / (x ^ p$Q + p$H ^ p$Q)
 
-  x[1] <- x0
-  for(t in 1:(N-1)){
-    # Determinstic mean looks like standard R
-    mu[t] <- x[t] + x[t] * r * (1 - x[t] / K)  - a * x[t] ^ Q / (x[t] ^ Q + H ^ Q)
-    # Note the use of ~ in BUGS to show 'distributed as normal' 
-    y[t+1] ~ dnorm(mu[t], sd = sigma)
-    x[t+1] <- max(y[t+1],0)
+stochastic_sim <- function(f, p){
+  x <- numeric(p$N)
+  x[1] <- p$x0
+  dBt <- rnorm(p$N, 0, p$sigma)
+  for(t in 1:(p$N-1)){
+    x[t+1] <- max( 
+                  f(x[t], p) + dBt[t],
+                  0)
+  }
+  data.frame(t = 1:p$N, x)
 }
 
-})
-model <- nimbleModel(may,constants = p, inits = list(x0 = 0.2))
-cmodel <- model #compileNimble(model)
-set.seed(123456)
-simulate(cmodel)
-df <- tibble(t = seq_along(cmodel$x), x = cmodel$x)
+det_sim <- function(f, p){
+  x <- numeric(p$N)
+  x[1] <- p$x0
+  for(t in 1:(p$N-1)){
+    x[t+1] <- f(x[t], p)
+  }
+  data.frame(t = 1:p$N, x)
+}
 ```
@@
 ```{r}
-df %>% ggplot(aes(t,x)) + geom_point()
-df %>% write_csv("../data/ex.csv")
-```
+plan("multisession")
+system.time({
+  df <- future_map_dfr(1:500, function(reps) stochastic_sim(may, p), .id = "reps")
+})
 
-Let's do 100 replicates
-
-```{r}
-set.seed(123)
-
-df <- map_dfr(1:100, 
-  function(i){
-    simulate(cmodel)
-    tibble(t = seq_along(cmodel$x), x = cmodel$x, reps = i)
-   })
 ```
@@
 ```{r}
-df %>% 
-  #filter(t %in% seq(1000, 2000, by = 1)) %>% 
-  ggplot(aes(t,x, group=reps)) + geom_line(alpha=.2)
+det <- det_sim(may,p) %>% mutate(reps=1)
+mean <- df %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
+```
 
+```{r}
+df %>% ggplot(aes(t, x, group=reps)) + 
+  geom_line(alpha=0.1) +
+  geom_line(data = det, color = "blue") + 
+  geom_line(data = mean, color = "red")
+```
+
+```{r}
 df %>% write_csv("../data/reps.csv")

### 2019-07-11 notebook/ghost-fit.Rmd
high vs low inference

@@
 ```{r}
-data <- read_csv("../data/reps.csv")
-data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.1)
+data <- read_csv("../data/reps.csv") # %>% filter(t %in% seq(0,4000, by=4))
+data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.02)
 ```
@@
 
+## highs
+
 ```{r}
@@
 ```{r}
-x_t1 <- wide[-1,] # data$x[-1]
-x_t <- wide[-n,] # data$x[-100]
-dim(x_t)
-dim(x_t1)
-#tibble(x_t, x_t1) %>% ggplot(aes(x_t, x_t1)) + geom_point()
+x_t1 <- wide[-1,]
+x_t <- wide[-n,] 
 ```
@@
 ```{r}
-#r <- 0.05 
+r <- 0.05 
 Q <- 5
-#sigma <- 0.02
-#a <- 0.0233
-K <- uniform(0, 10)
-H <- uniform(0,  2)
+K <- 2
+# a <- 0.0233
+H <- 0.38
+sigma <- 0.02
+
+# r <- uniform(0, 1)
+# Q <- uniform(0, 10)
+#K <- uniform(0, 10)
 a <- uniform(0, 1)
-r <- uniform(0, 1)
-#Q <- uniform(0, 10)
-sigma <- uniform(0, 1)
+# H <- uniform(0,  2)
+#sigma <- uniform(0, 1)
 
@@
 distribution(x_t1) <- normal(mean, sigma)
-m <- model(K, a, H, sigma, r)
+m <- model(a)
 ```
@@
 #Q = 5
-true <- data.frame(a = 0.023, r = .05, K = 2, H = .38, sigma = .02) %>%
+true <- 
+  data.frame(a = 0.023 #, 
+             #r = .05, 
+             #K = 2, 
+             #H = .38, 
+             #sigma = .02
+             ) %>%
   gather(variable, value)
@@
 ```
+
+## lows
+
+```{r}
+wide <- data %>% 
+  filter(reps %in% lows) %>%
+  spread(reps, x) %>%
+  select(-t) %>% 
+  as.matrix()
+n <- dim(wide)[1]
+
+```
+
+```{r}
+x_t1 <- wide[-1,]
+x_t <- wide[-n,] 
+```
+
+```{r}
+mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
+distribution(x_t1) <- normal(mean, sigma)
+m <- model(a)
+```
+
+```{r }
+system.time({
+  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
+```
+
+```{r}
+samples %>% ggplot() + 
+  geom_histogram(aes(value), bins = 30)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+```

### 2019-07-12 notebook/ghost-fit.Rmd
notes

@@
   gather(variable, value, -t, -chain)
+```
 
+```{r include=FALSE}
+## manual version of bayesplot::mcmc_trace()
 samples %>%  

### 2019-07-12 notebook/ghost-sims.Rmd
add may outbreak model

@@
 library(furrr)
-library(nimble)
 ```
@@
 ```{r}
-p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2)
+p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2, N = 1e4)
 growth <- function(x, p) x * p$r * (1 - x / p$K)
@@
 theory <- 
+  function(p){
     tibble(x= seq(0,2, length.out = 100)) %>%
@@
     gather(curve, y, -x, -potential) 
+}
 ```
@@
 ```{r noisy_switch}
-theory %>%
+theory(p) %>%
   ggplot(aes(x, y, col=curve)) +
@@
 ```{r}
-theory %>%
+theory(p) %>%
   ggplot(aes(x, potential)) + 
@@
     x[t+1] <- max( 
-                  f(x[t], p) + dBt[t],
+                  f(x[t], p) + x[t] * dBt[t],  # arguably x[t] * dB[t] would be a better model!
                   0)
@@
 }
+```
+
+```{r}
+a <- c(0.0225, 0.0230, 0.02305, 0.0231)
+names(a) <- a
+
+df <- map_dfr(a, function(a){
+  p$a <- a
+  p$N <- 1e4
+  det_sim(may, p)
+},
+.id = "a")
+
+df %>% ggplot(aes(t,x, col = a)) + geom_line() + coord_cartesian(xlim=c(0,1000))
+
+theory(p) %>% ggplot(aes(x, y, col=curve)) + geom_line(lwd=1)
+
 ```

### 2019-07-12 notebook/may-outbreak.Rmd
add may outbreak model

---
output: github_document
---

```{r}
knitr::opts_chunk$set(fig.width = 7)
```

```{r message=FALSE}
library(MDPtoolbox)
library(sarsop) # remotes::install_github("boettiger-lab/sarsop")
library(tidyverse) # for plotting
library(mdplearning)
```

# Outbreak model

- 'harvest' term corresponds to removal of pest, with associated cost
- also experience damage costs proportional to pest abundance

```{r}
damage <- 0.05
control <- 1
reward_fn <- function(x,h) - damage * x ^ 2 - control * h
discount <- 0.95

states <- seq(0,2, length=100)
actions <- states
observations <- states
sigma_g <- 0.05
sigma_m <- 0.0

r <- 0.8
K <- 1.53
q <- 2
b <- .2
eps <- states[2]
Tmax <- 100

may <- function(a){  
  function(x, h){ # May
    x <- pmax(x - h, 0)   # harvest then recruit   
    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps
  }
}

```

```{r}
a <- 0.29
m <- fisheries_matrices(states, actions, observations, reward_fn, 
                        may(a), sigma_g, sigma_m, noise = "lognormal")
```

```{r, results="hide"}
soln <- mdp_value_iteration(m$transition, m$reward, discount)
```

```{r}
df <- tibble(state = states,
             action = actions[soln$policy],
             value = soln$V)
## tipping point at maximum action:
df %>% ggplot(aes(state, action)) + geom_point() 
```

```{r}
x0 <- which.min(abs(states - 0.1))
#x0 <- which.min(abs(states - 2))
df <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
                   policy = soln$policy, x0 = x0, Tmax = 100)
df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) + geom_point()+geom_path() + 
  geom_line(aes(time, action), col="blue")
```

Range of possible a that covers tipping in both directions:

```{r}
possible_a <- seq(.26, .33, by = 0.01)

true_a <- 0.29
true_i <- which.min(abs(possible_a - true_a))

```

```{r}

models <- map(possible_a, function(a){
  fisheries_matrices(states, actions, observations, reward_fn, 
                     may(a), sigma_g, sigma_m, noise = "lognormal")
})

```

```{r}
transition <- lapply(models, `[[`, "transition")
reward <- models[[1]][["reward"]]
```

## Small a assumption

```{r}
mu <- possible_a[true_i - 2]
prior <- dnorm(possible_a, mu, 0.01)
prior <- prior / sum(prior)
```

```{r}
barplot(prior)
f <- may(mu)
tibble(x = states[1:80],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + geom_line() +
  geom_point() +  geom_hline(aes(yintercept = 0))
```

```{r}
sim <- mdp_learning(transition, reward, discount, 
                    x0 = x0, 
                    Tmax = Tmax, 
                    true_transition = transition[[true_i]],
                    model_prior = prior,
                    type = "policy iteration")
```

```{r}
 sim$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

```{r}
 sim$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% seq(1,Tmax, by = 5)) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = as.numeric(param)) %>% 
  ggplot(aes(param, probability, group = time, alpha = time)) + 
  geom_line()
```

## large a assumption

```{r}
mu <- possible_a[true_i + 2]
prior <- dnorm(possible_a, mu, 0.01)
prior <- prior / sum(prior)
```

```{r}
barplot(prior)
f <- may(mu)
tibble(x = states,
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + geom_line() +
  geom_point() +  geom_hline(aes(yintercept = 0))
```

```{r}
sim <- mdp_learning(transition, reward, discount, 
                    x0 = x0, 
                    Tmax = Tmax, 
                    true_transition = transition[[true_i]],
                    model_prior = prior,
                    type = "policy iteration")
```

```{r}
 sim$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

```{r}
 sim$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% seq(1,Tmax, by = 5)) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = as.numeric(param)) %>% 
  ggplot(aes(param, probability, group = time, alpha = time)) + 
  geom_line()
```

## Perfect knowledge

```{r}
mu <- possible_a[true_i]
#prior <- dnorm(possible_a, mu, 0.01)
#prior <- prior / sum(prior)
prior <- numeric(length(possible_a))
prior[true_i] <- 1
```

```{r}
barplot(prior)
f <- may(mu)
tibble(x = states,
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + geom_line() +
  geom_point() +  geom_hline(aes(yintercept = 0))
```

```{r}
sim <- mdp_learning(transition, reward, discount, 
                    x0 = x0, 
                    Tmax = Tmax, 
                    true_transition = transition[[true_i]],
                    model_prior = prior,
                    type = "policy iteration")
```

```{r}
 sim$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

```{r}
 sim$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% seq(1,Tmax, by = 5)) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = as.numeric(param)) %>% 
  ggplot(aes(param, probability, group = time, alpha = time)) + 
  geom_line()
```

### 2019-07-13 slides/boettiger.Rmd
WIP

---
title: "Conservation decision-making under long transients"
author: "Carl Boettiger"
institute: "UC Berkeley"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

background-image: url(img/royalsociety/penguin-swim.jpg)
background-position: center
background-size: 120%
class: center, top

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Decision-making under transients: what could go wrong?

???

Image credit: [Royal Society Photo Competition](https://royalsociety.org/journals/publishing-activities/photo-competition/)

---

class: center, middle

# Ghost Attractors

---
class: inverse, center, middle

# Acknowledgements

### 2019-07-15 notebook/may-outbreak.Rmd
larger learning space

@@
 ```{r}
-possible_a <- seq(.26, .33, by = 0.01)
+possible_a <- seq(.25, .34, by = 0.005)
 
@@
 ```{r}
-mu <- possible_a[true_i - 2]
+mu <- possible_a[true_i - 8]
 prior <- dnorm(possible_a, mu, 0.01)
@@
                     model_prior = prior,
-                    type = "policy iteration")
+                    type = "value iteration", 
+                    epsilon = 1e-2)
 ```
@@
 ```{r}
-mu <- possible_a[true_i + 2]
+mu <- possible_a[true_i + 8]
 prior <- dnorm(possible_a, mu, 0.01)
@@
                     model_prior = prior,
-                    type = "policy iteration")
+                    type = "value iteration",
+                    epsilon = 1e-2)
 ```
@@
 ```{r}
- sim$df %>% 
-  select(-value) %>% 
-  gather(series, state, -time) %>% 
-  ggplot(aes(time, states[state], color = series)) + geom_line()
+sim$df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
 ```

### 2019-07-17 notebook/may-outbreak.Rmd
exploring

@@
 library(mdplearning)
+
+#if(interactive()) ggplot2::theme_set(ggdark::dark_theme_light())
+#ggplot2::theme_set(theme_grey())
+#ggdark::invert_geom_defaults()
 ```
@@
 
-states <- seq(0,2, length=100)
+states <- seq(0,2, length=200)
 actions <- states
@@
 b <- .2
-eps <- states[2]
+eps <- states[2]/10
 Tmax <- 100
@@
   function(x, h){ # May
-    x <- pmax(x - h, 0)   # harvest then recruit   
-    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps
+    # x <- pmax(x - h, 0)   # harvest then recruit   
+    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps - h
   }
@@
 
-```
-
-```{r}
-a <- 0.29
-m <- fisheries_matrices(states, actions, observations, reward_fn, 
-                        may(a), sigma_g, sigma_m, noise = "lognormal")
-```
-
-```{r, results="hide"}
-soln <- mdp_value_iteration(m$transition, m$reward, discount)
-```
-
-```{r}
-df <- tibble(state = states,
-             action = actions[soln$policy],
-             value = soln$V)
-## tipping point at maximum action:
-df %>% ggplot(aes(state, action)) + geom_point() 
-```
-
-```{r}
-x0 <- which.min(abs(states - 0.1))
-#x0 <- which.min(abs(states - 2))
-df <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
-                   policy = soln$policy, x0 = x0, Tmax = 100)
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
 ```
@@
 possible_a <- seq(.25, .34, by = 0.005)
+true_a <- 0.26   ## reality has just a transient.  use  0.277 for stronger ghost
+believe_a <- 0.31 ## believe there's an attractor
+true_i <- which.min(abs(possible_a - true_a))
+```
 
-true_a <- 0.29
-true_i <- which.min(abs(possible_a - true_a))
+```{r}
+f <- may(true_a)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + geom_line() +
+  geom_point() +  geom_hline(aes(yintercept = 0))
+```
 
+```{r}
+f <- may(believe_a)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + geom_line() +
+  geom_point() +  geom_hline(aes(yintercept = 0))
+```
+
+```{r}
+m_true <- fisheries_matrices(states, actions, observations, reward_fn, 
+                        may(true_a), sigma_g, sigma_m, noise = "lognormal")
+
+m <- fisheries_matrices(states, actions, observations, reward_fn, 
+                        may(believe_a), sigma_g, sigma_m, noise = "lognormal")
+```
+
+```{r, results="hide"}
+soln <- mdp_value_iteration(m$transition, m$reward, discount)
+
+opt_soln <- mdp_value_iteration(m_true$transition, m_true$reward, discount)
+
+```
+
+Policy based on the belief (i.e. that system is bi-stable)
+
+```{r}
+df <- tibble(state = states,
+             action = actions[soln$policy],
+             value = soln$V)
+df %>% ggplot(aes(state, action)) + geom_point() 
+```
+
+Optimal policy (knowing it is really just a transient)
+```{r}
+tibble(state = states,
+       action = actions[opt_soln$policy]) %>% 
+  ggplot(aes(state, action)) + geom_point() 
+```
+
+```{r}
+no_policy <- numeric(length(states)) + 1
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = no_policy, x0 = x0, Tmax = 100)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Result experienced by incorrect belief:  initial in-action followed by need for continued maintenance to prevent high-level outbreak:
+
+```{r}
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = soln$policy, x0 = x0, Tmax = 100)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Expected result based on the belief: stable low level is acceptable, so no action is required:
+
+```{r}
+
+df <- mdp_planning(m$transition, m$reward, discount, model_prior = c(1), 
+                   policy = soln$policy, x0 = x0, Tmax = 100)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Optimal strategy knowing this is just a transient:
+
+```{r}
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = opt_soln$policy, x0 = x0, Tmax = 400)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
 ```
@@
 
-## Small a assumption
+## a near ghost
 
 ```{r}
-mu <- possible_a[true_i - 8]
-prior <- dnorm(possible_a, mu, 0.01)
+prior <- dnorm(possible_a, believe_a, 0.03)
 prior <- prior / sum(prior)
@@
 ```{r}
-barplot(prior)
-f <- may(mu)
-tibble(x = states[1:80],
-       f = f(x,0) - x) %>%
-  ggplot(aes(x, f)) + geom_line() +
-  geom_point() +  geom_hline(aes(yintercept = 0))
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
+  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
+  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
 ```
@@
 ```
-
-## large a assumption
-
-```{r}
-mu <- possible_a[true_i + 8]
-prior <- dnorm(possible_a, mu, 0.01)
-prior <- prior / sum(prior)
-```
-
-```{r}
-barplot(prior)
-f <- may(mu)
-tibble(x = states,
-       f = f(x,0) - x) %>%
-  ggplot(aes(x, f)) + geom_line() +
-  geom_point() +  geom_hline(aes(yintercept = 0))
-```
-
-```{r}
-sim <- mdp_learning(transition, reward, discount, 
-                    x0 = x0, 
-                    Tmax = Tmax, 
-                    true_transition = transition[[true_i]],
-                    model_prior = prior,
-                    type = "value iteration", 
-                    epsilon = 1e-2)
-```
-
-```{r}
- sim$df %>% 
-  select(-value) %>% 
-  gather(series, state, -time) %>% 
-  ggplot(aes(time, states[state], color = series)) + geom_line()
-```
-
-```{r}
- sim$posterior %>% 
-  data.frame(time = 1:Tmax) %>%
-  filter(time %in% seq(1,Tmax, by = 5)) %>%
-  gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = as.numeric(param)) %>% 
-  ggplot(aes(param, probability, group = time, alpha = time)) + 
-  geom_line()
-```
-
-## Perfect knowledge
-
-```{r}
-mu <- possible_a[true_i]
-#prior <- dnorm(possible_a, mu, 0.01)
-#prior <- prior / sum(prior)
-prior <- numeric(length(possible_a))
-prior[true_i] <- 1
-```
-
-```{r}
-barplot(prior)
-f <- may(mu)
-tibble(x = states,
-       f = f(x,0) - x) %>%
-  ggplot(aes(x, f)) + geom_line() +
-  geom_point() +  geom_hline(aes(yintercept = 0))
-```
-
-```{r}
-sim <- mdp_learning(transition, reward, discount, 
-                    x0 = x0, 
-                    Tmax = Tmax, 
-                    true_transition = transition[[true_i]],
-                    model_prior = prior,
-                    type = "value iteration",
-                    epsilon = 1e-2)
-```
-
-```{r}
-sim$df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-```{r}
- sim$posterior %>% 
-  data.frame(time = 1:Tmax) %>%
-  filter(time %in% seq(1,Tmax, by = 5)) %>%
-  gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = as.numeric(param)) %>% 
-  ggplot(aes(param, probability, group = time, alpha = time)) + 
-  geom_line()
-```

### 2019-07-19 notebook/may-outbreak-ghost.Rmd
Examples of learning under different priors.
With a wide prior, the decision-maker acts optimally to squash the outbreak early on. The decision-maker never really learns for sure if this was the right decision (i.e. if this was really necessary or if the outbreak would have stopped at low level), it is basically a precautionary reaction -- the posterior belief remains broad since learning opportunity is lost by crushing the outbreak.

In contrast, the decision-maker with a prior more confident in a low-level stable state (i.e. mistaking past evidence of the "ghost" for a stable state), fails t

--- 2019-07-17 notebook/may-outbreak.Rmd
@@
 reward_fn <- function(x,h) - damage * x ^ 2 - control * h
-discount <- 0.95
+discount <- 0.98
 
@@
 possible_a <- seq(.25, .34, by = 0.005)
-true_a <- 0.26   ## reality has just a transient.  use  0.277 for stronger ghost
-believe_a <- 0.31 ## believe there's an attractor
+true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
+believe_a <- 0.29 ## believe there's an attractor
 true_i <- which.min(abs(possible_a - true_a))
@@
 soln <- mdp_value_iteration(m$transition, m$reward, discount)
-
 opt_soln <- mdp_value_iteration(m_true$transition, m_true$reward, discount)
@@
   ggplot(aes(state, action)) + geom_point() 
+```
+
+```{r}
+x0 <- which.min(abs(states - 0.1))
 ```
@@
 
-Optimal strategy knowing this is just a transient:
+Optimal strategy knowing this is just a transient (will depend on discount rate):
 
@@
 df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
-                   policy = opt_soln$policy, x0 = x0, Tmax = 400)
+                   policy = opt_soln$policy, x0 = x0, Tmax = 100)
 
@@
 ```{r}
-prior <- dnorm(possible_a, believe_a, 0.03)
+prior <- dnorm(possible_a, believe_a, 0.01)
 prior <- prior / sum(prior)
@@
 ```
+
+```{r}
+```

### 2019-07-19 slides/boettiger.Rmd
WIP

@@
   xaringan::moon_reader:
+    css: ["default", "custom.css"]
     lib_dir: libs
@@
 
-# Decision-making under transients: what could go wrong?
+# Decisions vs transient dynamics: what could possibly go wrong?
 
@@
 ---
+background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .9)), url(img/royalsociety/pensive-polarbear.jpg)
+background-size: 120%
+class: center, middle, inverse
 
-class: center, middle
+# in an era of predictive black-box models...
+
+---
+background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .9)), url(img/royalsociety/pensive-polarbear.jpg)
+background-size: 120%
+class: center, middle, inverse
+
+# ... a need for theory & process-based models
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
+background-size: 120%
+class: center, middle, inverse
 
@@
 ---
-class: inverse, center, middle
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
+background-size: 120%
+class: center, top, inverse
+
+# Ghost Attractors
+
+![](img/ghost-invert.png)
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
+background-size: 120%
+class: center, top, inverse
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+background-size: 120%
+class: center, top, inverse
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/pensive-polarbear.jpg)
+background-size: 120%
+class: center, top, inverse
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 120%
+class: center, top, inverse
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/eagle-ray.jpg)
+background-size: 120%
+class: center, top, inverse
+
+---
+class: center, middle
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/bohemian-waxwing.jpg)
+background-size: 120%
 
 # Acknowledgements
+
+![](img/hastings.jpg)
+
+![](img/nimbios.jpg)
+
+<!--
+- Lizzie Wolkovich
+- Henry Scharf
+-->

### 2019-07-26 notebook/ghost-fit.Rmd
add may outbreak model
slides WIP

@@
 )
-df <- data %>% mutate(coding = coding(reps)) %>% na.omit()
-df %>% ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)
+extremes <- data %>% mutate(coding = coding(reps)) %>% na.omit()
+
+write_csv(extremes, "../data/extremes.csv")
+extremes %>% ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)
 
@@
 
+----
+
 ## lows
@@
   filter(reps %in% lows) %>%
+  spread(reps, x) %>%
+  select(-t) %>% 
+  as.matrix()
+n <- dim(wide)[1]
+
+```
+
+```{r}
+x_t1 <- wide[-1,]
+x_t <- wide[-n,] 
+```
+
+```{r}
+mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
+distribution(x_t1) <- normal(mean, sigma * x_t)
+m <- model(a)
+```
+
+```{r }
+system.time({
+  draws2 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws2)
+```
+
+```{r}
+samples2 <-  
+  map_dfr(draws2, 
+          function(x) data.frame(x, t = 1:dim(x)[1]), 
+          .id = "chain") %>% 
+  gather(variable, value, -t, -chain)
+
+samples2 %>% ggplot() + 
+  geom_histogram(aes(value), bins = 30)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+```
+
+## Random sample
+
+```{r}
+
+random <- sample(unique(data$reps), 1)
+wide <- data %>% 
+  filter(reps %in% random) %>%
   spread(reps, x) %>%
@@
 system.time({
-  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+  draws3 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
 })
@@
 ```{r}
-bayesplot::mcmc_trace(draws)
+bayesplot::mcmc_trace(draws3)
 ```
@@
 ```{r}
-samples %>% ggplot() + 
+samples3 <-  
+  map_dfr(draws3, 
+          function(x) data.frame(x, t = 1:dim(x)[1]), 
+          .id = "chain") %>% 
+  gather(variable, value, -t, -chain)
+
+samples3 %>% ggplot() + 
   geom_histogram(aes(value), bins = 30)  +

### 2019-07-26 notebook/ghost-sims.Rmd
slides WIP

@@
 library(furrr)
+set.seed(12345)
 ```
@@
     tibble(x= seq(0,2, length.out = 100)) %>%
-    mutate(g = growth(x,p), 
-           c = consumption(x,p)) %>%
+    mutate(growth = growth(x,p), 
+           consumption = consumption(x,p)) %>%
     mutate(potential = - cumsum(g - c)) %>%
@@
     x[t+1] <- max( 
-                  f(x[t], p) + x[t] * dBt[t],  # arguably x[t] * dB[t] would be a better model!
+                  f(x[t], p) + x[t] * dBt[t],
                   0)
@@
 }
+```
+
+```{r}
+det <- det_sim(may,p) 
+det %>% ggplot(aes(t, x)) + geom_line() + theme_bw()
 ```

### 2019-07-26 slides/boettiger.Rmd
slides WIP

@@
   xaringan::moon_reader:
-    css: ["default", "custom.css"]
+    css: ["default"]
     lib_dir: libs
@@
 ```{r setup, include=FALSE}
-options(htmltools.dir.version = FALSE)
+knitr::opts_chunk$set(dev.args=list(bg="transparent"), 
+                      echo = FALSE, message=FALSE, warning=FALSE,
+                      fig.width=11, fig.height=6.5, cache = TRUE)
+
+library(tidyverse)
+library(ggthemes)
+
+theme_set(
+ theme_bw(base_size=26) + 
+  theme(line = element_line(colour = "white", size = 1.5),
+        rect = element_rect(fill = "transparent", colour = "white"),
+        text = element_text(colour = "white"),
+        axis.text = element_text(colour = "white"),
+        axis.ticks = element_line(colour = "white", size = 1.5),
+        panel.border = element_rect(colour = "white", size = 1.5),
+        panel.background = element_rect(fill = "transparent"),
+        plot.background = element_blank(), 
+        legend.background = element_blank(), 
+        legend.key = element_rect(fill = "transparent", colour = "transparent"), 
+        strip.background = element_blank(),
+        strip.text = element_text(colour = "white"),
+        panel.grid = element_blank()))
+
+source("../R/ghost.R")
 ```
@@
 ---
-background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .9)), url(img/royalsociety/pensive-polarbear.jpg)
-background-size: 120%
-class: center, middle, inverse
+background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url(img/city-mit-license.jpg)
+background-size: 140%
+class: center, top, inverse
 
@@
 ---
-background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .9)), url(img/royalsociety/pensive-polarbear.jpg)
+background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url(img/royalsociety/pensive-polarbear.jpg)
 background-size: 120%
-class: center, middle, inverse
+class: center, top, inverse
 
@@
 background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
+background-position: center
 background-size: 120%
-class: center, middle, inverse
+class: center, top, inverse
 
@@
 background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
+background-position: center
 background-size: 120%
@@
 
----
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
-background-size: 120%
-class: center, top, inverse
+<br/>
+Hastings et al. *Science* (2018)
 
@@
 background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+background-size: 120%
+background-position: center
+class: left, top, inverse
+
+# Examples
+
+Outbreak: disease, pest, invasive species
+
+- Low endemic level
+- outbreak
+- saturation
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+background-size: 120%
+background-position: center
+class: center, top, inverse
+
+# A look at the model
+
+$$X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{logistic growth}}  - \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{saturating consumption}} + \underbrace{\xi_t X_t}_{\textrm{environmental noise}}$$
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
+background-size: 120%
+background-position: center
+class: center, top, inverse
+
+## Potential well diagram
+
+```{r echo=FALSE}
+p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, 
+          N = 4e3, x0 = 0.2, N = 1e4)
+theory(p) %>%
+  ggplot(aes(x, y, col=curve)) +
+  geom_line(lwd = 2)
+
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
+background-size: 120%
+background-position: center
+class: center, top, inverse
+
+## Potential well diagram
+
+```{r echo=FALSE}
+theory(p) %>%
+  ggplot(aes(x, potential)) + 
+  geom_line(lwd = 2, col="white")
+
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
+background-size: 120%
+background-position: center
+class: center, top, inverse
+
+## Dynamics from the deterministic skeleton
+
+```{r echo=FALSE}
+det <- det_sim(may,p) %>% mutate(reps=1) %>% filter(t < 3000)
+det  %>%
+  ggplot(aes(t, x)) + 
+  geom_line(lwd = 2, col="white")
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
+background-size: 120%
+class: center, top, inverse
+background-position: center
+
+# Under the stochastic model
+
+```{r}
+extremes <- read_csv("../data/extremes.csv")
+#extremes %>%  filter(coding == "high", t < 3000) %>% pull(reps) %>% unique()
+
+extremes %>% 
+  filter(coding == "high", reps == 14, t < 3000) %>%  # %>% pull(reps) %>% unique()
+  ggplot(aes(t, x)) + 
+  geom_line(lwd = 2, col="white")
+
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
+background-size: 120%
+class: center, top, inverse
+background-position: center
+
+# Under the stochastic model
+
+```{r echo=FALSE}
+#extremes %>%  filter(coding == "low", t < 3000) %>% pull(reps) %>% unique()
+extremes %>% 
+  filter(coding == "low", reps == 47, t < 3000) %>%
+  ggplot(aes(t, x)) + 
+  geom_line(lwd = 2, col="white")
+
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 120%
+class: center, top, inverse
+background-position: center
+
+# Under the stochastic model
+
+```{r}
+data <- vroom::vroom("../data/reps.csv") %>% filter(t < 3000)
+mean <- data %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
+```
+
+```{r}
+data %>% ggplot(aes(t,x, group=reps)) + 
+  geom_line(alpha=0.02, color = "white")
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 120%
+class: center, top, inverse
+background-position: center
+
+# Under the stochastic model
+
+```{r echo=FALSE}
+data %>% ggplot(aes(t,x, group=reps)) + 
+  geom_line(alpha=0.02, color = "white") + 
+  geom_line(data = det, lwd=2, color = "white")
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 120%
+class: center, top, inverse
+background-position: center
+
+# Under the stochastic model
+
+```{r echo=FALSE}
+data %>% 
+  ggplot(aes(t,x, group=reps)) + 
+  geom_line(alpha=0.02, color = "white") + 
+  geom_line(data = mean, lwd=2, color = rgb(.6, .6, 1))
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+background-size: 120%
+class: center, top, inverse
+background-position: center
+
+# Challenges to Model Inference
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
 background-size: 120%

### 2019-07-26 R/ghost.R

@@ -1,0 +1,39 @@
+growth <- function(x, p) x * p$r * (1 - x / p$K)
+consumption <- function(x, p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+
+theory <-
+  function(p){
+    tibble(x= seq(0,2, length.out = 100)) %>%
+      mutate(growth = growth(x,p),
+             consumption = consumption(x,p)) %>%
+      mutate(potential = - cumsum(growth - consumption)) %>%
+      gather(curve, y, -x, -potential)
+  }
+
+
+
+
+may <- function(x, p) x + growth(x,p) - consumption(x,p)
+#x + x * p$r * (1 - x / p$K)  - p$a * x ^ p$Q / (x ^ p$Q + p$H ^ p$Q)
+
+stochastic_sim <- function(f, p){
+  x <- numeric(p$N)
+  x[1] <- p$x0
+  dBt <- rnorm(p$N, 0, p$sigma)
+  for(t in 1:(p$N-1)){
+    x[t+1] <- max(
+      f(x[t], p) + x[t] * dBt[t],
+      0)
+  }
+  data.frame(t = 1:p$N, x)
+}
+
+det_sim <- function(f, p){
+  x <- numeric(p$N)
+  x[1] <- p$x0
+  for(t in 1:(p$N-1)){
+    x[t+1] <- f(x[t], p)
+  }
+  data.frame(t = 1:p$N, x)
+}
+

### 2019-08-02 slides/boettiger.Rmd
slides material

@@
       countIncrementalSlides: false
+      slideNumberFormat: "%current%"
+
 ---
@@
   geom_line(alpha=0.02, color = "white") + 
-  geom_line(data = det, lwd=2, color = "white")
+  geom_line(data = mean, lwd=2, color = "white")
 ```
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin-wtf.jpg)
 background-size: 120%
@@
   geom_line(alpha=0.02, color = "white") + 
-  geom_line(data = mean, lwd=2, color = rgb(.6, .6, 1))
+  geom_line(data = mean, lwd = 2, color = "white") +
+  geom_line(data = det, lwd = 2, color = rgb(.6, .6, 1)) 
 ```
@@
 
----
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/gentoo-penguin.jpg)
-background-size: 120%
-class: center, top, inverse
+# Descision Theory
 

### 2019-08-06 slides/boettiger.Rmd
WIP

@@
 
----
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 120%
-background-position: center
-class: center, top, inverse
-
-## Potential well diagram
-
 ```{r echo=FALSE}
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
 background-size: 120%
@@
 
-## Dynamics from the deterministic skeleton
+## Time-series of the dynamics from the deterministic skeleton
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
-background-size: 120%
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 110%
 class: center, top, inverse
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin-wtf.jpg)
-background-size: 120%
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin-wtf.jpg)
+background-size: 110%
 class: center, top, inverse
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 110%
+class: center, top, inverse
+background-position: center
+
+#  First-passage times to the stable attractor
+
+```{r echo=FALSE}
+## det equib
+# data  %>% filter(t == max(t)) %>% pull(x) %>% mean() ## ~ 1.25
+
+## First-passage time, tau
+passage <- data %>% group_by(reps) %>% filter(x > 1.25) %>% summarise(tau = min(t))
+det_tau <- det %>% group_by(reps) %>% filter(x > 1.25) %>% summarise(tau = min(t)) %>% pull(tau)
+
+passage %>% ggplot(aes(tau)) + geom_histogram(binwidth = 200, fill=rgb(.95,.95,.95)) + geom_vline(aes(xintercept=det_tau), col="blue", lwd=2) #+ theme_bw(42)
+
+```
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 110%
+class: left, top, inverse
+background-position: center
+
+# Features that drive long transient dynamics also create stochastic phenomena
+
+- Emergent ensemble dynamics not predicted by the deterministic model
+- ... But intuition is simple: noise helps 'jump' the ghost
+- .... But also creates dynamics that are deeply challenging for both inference & decision-making
+
+---
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/reef-shark.jpg)
 background-size: 120%
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/reef-shark.jpg)
 background-size: 120%
-class: center, top, inverse
+class: left, top, inverse
+
+# Challenges to Model Inference
+
+Given time-series observations, can
 

### 2019-08-06 notebook/discrete-sims.Rmd
nice simple example of discrete simulation

---
output: github_document
---

```{r}
knitr::opts_chunk$set(fig.width = 7)
```

```{r message=FALSE}
library(MDPtoolbox)
library(mdplearning)
library(expm)
library(truncnorm)
library(tidyverse)

```

# Outbreak model

- 'harvest' term corresponds to removal of pest, with associated cost
- also experience damage costs proportional to pest abundance

```{r}
damage <- 0.05
control <- 1
reward_fn <- function(x,h) - damage * x ^ 2 - control * h
discount <- 0.98

n_s <- 200
states <- seq(0,2, length=n_s)
actions <- states
observations <- states
sigma_g <- 0.05

r <- 0.8
K <- 1.53
q <- 2
b <- .2
eps <- states[2]/10
Tmax <- 100

may <- function(a){  
  function(x, h = 0){ # May
    # x <- pmax(x - h, 0)   # harvest then recruit   
    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps - h
  }
}

```

Range of possible a that covers tipping in both directions:

```{r}
possible_a <- seq(.25, .34, by = 0.005)
true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
## 0.28: technically 'ghost' is an attractor but effectively not
believe_a <- 0.28 ## believe there's an attractor
true_i <- which.min(abs(possible_a - true_a))
```

```{r}
f <- may(true_a)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + geom_line() +
  geom_point() +  geom_hline(aes(yintercept = 0))
```

```{r}
f <- may(believe_a)
tibble(x = states[1:120],
       f = f(x,0) - x) %>%
  ggplot(aes(x, f)) + geom_line() +
  geom_point() +  geom_hline(aes(yintercept = 0))
```

```{r}
transition_matrix <- function(states, f, sigma){
    n_s <- length(states)
    transition <- array(0, dim = c(n_s, n_s))
    for (k in 1:n_s) {

            nextpop <- f(states[k])
            if (nextpop <= 0) {
                transition[, k] <- c(1, rep(0, n_s - 1))
            }
            else {
              x <- truncnorm::dtruncnorm(states, 0, max(states), nextpop, sigma * nextpop)
              transition[, k] <- x / sum(x)

            }

    }
    transition
}
```

```{r}
m_true <- transition_matrix(states, may(true_a), sigma_g)
m <- transition_matrix(states, may(believe_a), sigma_g)
```

"simulate" without any actions

```{r}
x0 <- which.min(abs(states - 0.1))
Tmax <- 300
```

```{r}
X <- numeric(length(states))
X[x0] <- 1

prob_dynamics <- function(M, X, Tmax){
  probability <- M %^% Tmax %*% X 
  data.frame(state = states, probability)
}

prob_dynamics(m, X, Tmax) %>% 
  ggplot(aes(state,probability))  + geom_point() # + geom_bar(stat="identity")
```

```{r}

sim <- function (transition,  x0, Tmax ){
    n_states <- dim(transition)[2]
    state <- numeric(Tmax + 1)
    state[1] <- x0
    time <- 1:Tmax
    for (t in time) {
        state[t + 1] <- base::sample(1:n_states, 
                               1, 
                               prob = transition[, state[t]])
    }
    data.frame(time = 1:Tmax, state = state[time])
}

sim(m, x0, Tmax) %>% 
  mutate(state = states[state]) %>% 
  ggplot(aes(time, state)) + geom_point() + geom_path() 
```

```{r}
```

### 2019-08-06 notebook/ghost-fit.Rmd
update sims

@@
 ```{r}
-data <- read_csv("../data/reps.csv") # %>% filter(t %in% seq(0,4000, by=4))
+data <- read_csv("../data/reps.csv.xz") # %>% filter(t %in% seq(0,4000, by=4))
 data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.02)
@@
 
-write_csv(extremes, "../data/extremes.csv")
+write_csv(extremes, "../data/extremes.csv.xz")
 extremes %>% ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)

### 2019-08-06 notebook/ghost-sims.Rmd
sims
update sims

@@
            consumption = consumption(x,p)) %>%
-    mutate(potential = - cumsum(g - c)) %>%
+    mutate(potential = - cumsum(growth - consumption)) %>%
     gather(curve, y, -x, -potential) 
@@
 ```{r}
-df %>% write_csv("../data/reps.csv")
+df %>% write_csv("../data/reps.csv.xz")
 ```

### 2019-08-06 notebook/make.R

@@ -1,0 +1,18 @@
+library(callr)
+
+
+p1 <- callr::r_bg(function() rmarkdown::render("notebook/ghost-sims.Rmd") )
+
+#  needs to wait for p1 to complete
+p2 <- callr::r_bg(function() rmarkdown::render("notebook/ghost-fit.Rmd") )
+
+# greta tasks probably cannot run in parallel given current memory settings
+p3 <- callr::r_bg(function() rmarkdown::render("notebook/discrete-sims.Rmd") )
+
+
+callr::poll(list(p1, p2), -1)
+
+
+p1$read_output_lines()
+p1$get_result()
+p1$get_exit_status()

### 2019-08-06 notebook/ghost-fit.Rmd
full ensemble

@@
 mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
-distribution(x_t1) <- normal(mean, sigma)
+distribution(x_t1) <- normal(mean, sigma * x_t)
 m <- model(a)
@@
 ```{r}
+a <- uniform(0, 1)
 mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
@@
 
-## Random sample
+## Full ensemble
 
 ```{r}
-
-random <- sample(unique(data$reps), 1)
 wide <- data %>% 
-  filter(reps %in% random) %>%
   spread(reps, x) %>%
@@
 ```{r}
+a <- uniform(0, 1)
 mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
-distribution(x_t1) <- normal(mean, sigma)
+distribution(x_t1) <- normal(mean, sigma * x_t)
 m <- model(a)
@@
 system.time({
-  draws3 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+  draws3 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 6, verbose = FALSE)
 })
@@
 
+```{r }
+p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2, N = 1e4)
+source("../R/ghost.R")
+p$a <- 0.021
+tf <- theory(p) %>% mutate(a = 0.021)
+
+p$a <- 0.023
+tf <- theory(p) %>% mutate(a = 0.023) %>% bind_rows(tf)
+
+p$a <- 0.024
+tf <- theory(p) %>% mutate(a = 0.024) %>% bind_rows(tf)
+
+tf <- tf %>% mutate(a = as.factor(a))
+
+tf %>%
+  ggplot(aes(x, potential, col=a)) +
+  geom_line(lwd=1)
+
+```
+
 ```{r}

### 2019-08-07 notebook/ghost-fit.Rmd
store indices only

@@
 lows <- means %>% dplyr::slice(1:10) %>% pull(reps)
-highs <- means %>% dplyr::slice((n()-10):n()) %>% pull(reps)
+highs <- means %>% dplyr::slice((n()-9):n()) %>% pull(reps)
 coding <- function(reps) case_when(
@@
 
-write_csv(extremes, "../data/extremes.csv.xz")
+write_csv(data.frame(lows=lows,highs=highs), "../data/extremes.csv.xz")
 extremes %>% ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)

### 2019-08-08 notebook/discrete-sims.Rmd
model inference & sims

@@
 control <- 1
-reward_fn <- function(x,h) - damage * x ^ 2 - control * h
-discount <- 0.98
+#reward_fn <- function(x,h) - damage * x ^ 2 - control * h
+#discount <- 0.98
 
@@
 states <- seq(0,2, length=n_s)
-actions <- states
-observations <- states
+
 sigma_g <- 0.05
-
 r <- 0.8
@@
 eps <- states[2]/10
-Tmax <- 100
 
@@
 m_true <- transition_matrix(states, may(true_a), sigma_g)
-m <- transition_matrix(states, may(believe_a), sigma_g)
+m_belief <- transition_matrix(states, may(believe_a), sigma_g)
 ```
@@
 
-prob_dynamics(m, X, Tmax) %>% 
+prob_dynamics(m_belief, X, Tmax) %>% 
   ggplot(aes(state,probability))  + geom_point() # + geom_bar(stat="identity")
@@
 }
+```
 
-sim(m, x0, Tmax) %>% 
-  mutate(state = states[state]) %>% 
+```{r}
+set.seed(12345)
+no_switches <- sim(m_belief, x0, Tmax)  %>% 
+  mutate(state = states[state])
+
+no_switches %>% 
   ggplot(aes(time, state)) + geom_point() + geom_path() 
@@
 ```{r}
+set.seed(11)
+switches <- sim(m_belief, x0, Tmax) %>% mutate(state = states[state])
+
+switches  %>% 
+  ggplot(aes(time, state)) + geom_point() + geom_path() 
 ```
+
+-----
+
+# Greta
+
+Vectorize data
+
+```{r}
+X <- switches$state
+n <- length(X)
+x_t1 <- X[-1]
+x_t <- X[-n] 
+```
+
+```{r}
+library(greta)
+#a <- uniform(0, 1)
+a <- uniform(.25, .34)
+wd <- 0.0005 # binwidth for plots
+```
+
+```{r}
+mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ q / (x_t ^ q + b ^ q)
+distribution(x_t1) <- normal(mean, sigma_g * x_t)
+m <- model(a)
+```
+
+```{r }
+system.time({
+  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
+```
+
+```{r}
+samples <-  
+  map_dfr(draws, function(x) data.frame(x, t = 1:dim(x)[1]), .id = "chain") %>% 
+  gather(variable, value, -t, -chain)
+
+true <- data.frame(a = believe_a)  %>% gather(variable, value)
+
+samples %>% ggplot() + 
+  geom_histogram(aes(value), binwidth = wd)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+
+write_csv(samples, "../data/switches.csv.xz")
+```
+
+-------
+
+```{r}
+
+X <- no_switches$state
+n <- length(X)
+x_t1 <- X[-1]
+x_t <- X[-n] 
+```
+
+```{r}
+a <- uniform(.25, .34)
+mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ q / (x_t ^ q + b ^ q)
+distribution(x_t1) <- normal(mean, sigma_g * x_t)
+m <- model(a)
+```
+
+```{r }
+system.time({
+  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
+```
+
+```{r}
+samples <-  
+  map_dfr(draws, 
+          function(x) data.frame(x, t = 1:dim(x)[1]), 
+          .id = "chain") %>% 
+  gather(variable, value, -t, -chain)
+
+samples %>% ggplot() + 
+  geom_histogram(aes(value), binwidth = wd)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+
+write_csv(samples, "../data/no_switches.csv.xz")
+
+```
+
+```{r}
+set.seed(12345)
+weak_ghost <- sim(m_true, x0, Tmax)  %>% 
+  mutate(state = states[state])
+
+weak_ghost %>% 
+  ggplot(aes(time, state)) + geom_point() + geom_path() 
+```
+
+Vectorize data
+
+```{r}
+X <- weak_ghost$state
+n <- length(X)
+x_t1 <- X[-1]
+x_t <- X[-n] 
+```
+
+```{r}
+#a <- uniform(.25, .34)
+a <- uniform(0, 1)
+```
+
+```{r}
+mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ q / (x_t ^ q + b ^ q)
+distribution(x_t1) <- normal(mean, sigma_g * x_t)
+m <- model(a)
+```
+
+```{r }
+system.time({
+  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
+})
+```
+
+```{r}
+bayesplot::mcmc_trace(draws)
+```
+
+```{r}
+samples <-  
+  map_dfr(draws, function(x) data.frame(x, t = 1:dim(x)[1]), .id = "chain") %>% 
+  gather(variable, value, -t, -chain)
+
+true <- data.frame(a = true_a)  %>% gather(variable, value)
+
+samples %>% ggplot() + 
+  geom_histogram(aes(value), binwidth = wd)  +
+  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
+  facet_wrap(~variable, scales = "free")
+
+write_csv(samples, "../data/weak_ghost.csv.xz")
+
+```
+
+```{r}
+df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
+  map_dfr(read_csv, .id = "timeseries")
+```

### 2019-08-08 notebook/make.R

@@ -10,9 +10,13 @@ p2 <- callr::r_bg(function() rmarkdown::render("notebook/ghost-fit.Rmd") )
 p3 <- callr::r_bg(function() rmarkdown::render("notebook/discrete-sims.Rmd") )
 
 
-callr::poll(list(p1, p2), -1)
+
+p4 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost-2.Rmd") )
 
 
-p1$read_output_lines()
-p1$get_result()
+callr::poll(list(p4), -1)
+
+
+p4$read_output_lines()
+p4$get_result()
 p1$get_exit_status()

### 2019-08-08 notebook/may-outbreak-ghost-2.Rmd
with worse initial belief

--- 2019-07-19 notebook/may-outbreak-ghost.Rmd
@@
 true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
-believe_a <- 0.29 ## believe there's an attractor
+believe_a <- 0.31 ## believe there's a strong attractor
 true_i <- which.min(abs(possible_a - true_a))
@@
 ```{r}
-prior <- dnorm(possible_a, believe_a, 0.01)
+prior <- dnorm(possible_a, believe_a, 0.05)
 prior <- prior / sum(prior)

### 2019-08-08 slides/boettiger.Rmd
WIP slides

--- 2019-08-02 slides/boettiger.Rmd
@@
 
----
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
-background-position: center
-background-size: 120%
-class: center, top, inverse
-
-# Ghost Attractors
+--
 
@@
 ---
+layout: true
 background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
@@
 
-# Examples
-
-Outbreak: disease, pest, invasive species
-
-- Low endemic level
-- outbreak
-- saturation
-
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
-background-size: 120%
-background-position: center
-class: center, top, inverse
 
@@
 
----
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 120%
-background-position: center
-class: center, top, inverse
-
-## Potential well diagram
-
-```{r echo=FALSE}
+```{r fig.height=5.5}
 p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 120%
-background-position: center
-class: center, top, inverse
 
-## Potential well diagram
+## potential well diagram
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 120%
-background-position: center
-class: center, top, inverse
 
-## Dynamics from the deterministic skeleton
+## deterministic skeleton timeseries
 
@@
 ```
+---
+layout: true
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-size: 150%
+class: center, top, inverse
+background-position: left
+
+# Stochastic simulations
 
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 120%
-class: center, top, inverse
-background-position: center
-
-# Under the stochastic model
 
 ```{r}
-extremes <- read_csv("../data/extremes.csv")
-#extremes %>%  filter(coding == "high", t < 3000) %>% pull(reps) %>% unique()
+data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
+## get tail cases for comparison
+ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
+lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
+highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
+```
 
-extremes %>% 
-  filter(coding == "high", reps == 14, t < 3000) %>%  # %>% pull(reps) %>% unique()
-  ggplot(aes(t, x)) + 
-  geom_line(lwd = 2, col="white")
+```{r}
+data %>% filter(reps == lows[[6]]) %>%
+  ggplot(aes(t, x, group=reps)) + 
+  geom_line(lwd=2, col="white") + 
+  ylim(0,1.6)
+```
+
+---
+
+```{r}
+
+data %>% filter(reps == highs[[6]]) %>%
+  ggplot(aes(t, x, group=reps)) + 
+  geom_line(lwd=2, col="white") + 
+  ylim(0,1.6)
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 120%
-class: center, top, inverse
-background-position: center
-
-# Under the stochastic model
-
-```{r echo=FALSE}
-#extremes %>%  filter(coding == "low", t < 3000) %>% pull(reps) %>% unique()
-extremes %>% 
-  filter(coding == "low", reps == 47, t < 3000) %>%
-  ggplot(aes(t, x)) + 
-  geom_line(lwd = 2, col="white")
-
-```
-
----
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
-background-size: 120%
-class: center, top, inverse
-background-position: center
-
-# Under the stochastic model
-
-```{r}
-data <- vroom::vroom("../data/reps.csv") %>% filter(t < 3000)
-mean <- data %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
-```
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
-background-size: 120%
-class: center, top, inverse
-background-position: center
-
-# Under the stochastic model
 
 ```{r echo=FALSE}
+mean <- data %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
 data %>% ggplot(aes(t,x, group=reps)) + 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin-wtf.jpg)
-background-size: 120%
-class: center, top, inverse
-background-position: center
-
-# Under the stochastic model
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin-wtf.jpg)
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+
+**First-passage times**
+
+```{r echo=FALSE}
+## det equib
+# data  %>% filter(t == max(t)) %>% pull(x) %>% mean() ## ~ 1.25
+
+## First-passage time, tau
+passage <- data %>% group_by(reps) %>% filter(x > 1.25) %>% summarise(tau = min(t))
+det_tau <- det %>% group_by(reps) %>% filter(x > 1.25) %>% summarise(tau = min(t)) %>% pull(tau)
+```
+
+```{r}
+passage %>% ggplot(aes(tau)) + 
+  geom_histogram(binwidth = 200, fill=rgb(.9, .9, .9, .7)) + 
+  geom_vline(aes(xintercept=det_tau), col="blue", lwd=2)
+```
+
+---
+layout: false
+background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/green-crested-lizard.jpg)
+background-size: 110%
+class: left, top, inverse
+background-position: center
+
+## Long transient dynamics can
+## also create stochastic phenomena
+
+--
+
+- Emergent ensemble dynamics not predicted by the deterministic model
+
+--
+
+- But intuition is simple: noise helps 'jump' the ghost
+
+--
+
+- Noise creates dynamics that are deeply challenging for both inference & decision-making...
+
+---
+background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .85)), url(img/royalsociety/reef-shark.jpg)
 background-size: 120%
-class: center, top, inverse
+class: top, inverse
 background-position: center
+layout: true
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
-background-size: 120%
-class: center, top, inverse
 
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/pensive-polarbear.jpg)
+
+- Given time-series observations, can we distinguish ghost attractors from real attractors?
+
+```{r}
+discrete_sims <- read_csv("../data/discrete-sims.csv.xz")
+discrete_sims %>% filter(series %in% c("1", "2")) %>% ggplot(aes(time,state)) + geom_line(col="white", lwd=1.5) + facet_wrap(~ series)
+```
+
+---
+class: left
+
+Hierarchical Bayesian estimation of stability using TensorFlow(R)
+
+```r
+library("greta")
+
+mean <- x_t + r * x_t * (1 - x_t / K) - 
+        a * x_t ^ q / (x_t ^ q + b ^ q)
+distribution(x_t1) <- normal(mean, sigma_g * x_t)
+
+a <- uniform(.25, .34) # Prior
+
+draws <- mcmc(model(a), n_samples = 1000, warmup = 3000, chains = 4)
+```
+
+---
+class: left
+
+```{r fig.width=12.5, fig.height=7}
+label <- c("no switch", "switch", "no ghost")
+
+df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
+  map_dfr(read_csv, .id = "timeseries") %>% 
+  mutate(timeseries = label[as.integer(timeseries)])
+
+inset <- discrete_sims %>% mutate(series = label[series]) %>% 
+  ggplot(aes(time,state, col=series)) + 
+  geom_line(lwd=1.5, show.legend = FALSE) + 
+  ggplot2::scale_color_viridis_d()
+
+p1 <- df %>% 
+  ggplot((aes(value, fill=timeseries))) + 
+  geom_histogram(alpha=0.9) + 
+  geom_vline(aes(xintercept=0.27, col = "no switch"), col= viridis::viridis_pal()(3)[1], lwd=2) + 
+  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = viridis::viridis_pal()(3)[2], lwd=2) + 
+  ggplot2::scale_fill_viridis_d()
+
+vp <- grid::viewport(width = 4, height = 0.8, x = 0.3, y = 0.7)
+
+disp <- function(){
+print(p1)
+print(inset, vp = vp)
+}
+
+disp()
+```
+
+---
+layout: true
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/forest-fire.jpg)
 background-size: 120%
@@
 # Descision Theory
+
+---
+
+### Management under uncertainty
+
+---
+
+---
+layout: true
+background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/pensive-polarbear.jpg)
+background-size: 120%
+class: center, top, inverse
 

### 2019-08-09 slides/boettiger.Rmd
arima example

@@
 
+```{r}
+data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
+## get tail cases for comparison
+ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
+lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
+highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
+```
+
 # Decisions vs transient dynamics: what could possibly go wrong?
@@
 ---
+name: polarbear
 background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url(img/royalsociety/pensive-polarbear.jpg)
@@
 
+---
+template: polarbear
+
 # ... a need for theory & process-based models
@@
 ---
+layout: true
 background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
@@
 class: center, top, inverse
+
+---
 
@@
 ---
+
+# Ghost Attractor
+
+![](img/ghost.png)
+
+<br/>
+Hastings et al. *Science* (2018)
+
+---
+
+# A Ghost Attractor
+
+```{r}
+data %>% filter(reps == lows[[6]]) %>%
+  ggplot(aes(t, x, group=reps)) + 
+  geom_line(lwd=2, col="white") + 
+  ylim(0,1.6)
+```
+
+---
+
+# Simple ARIMA forecast:
+
+```{r}
+library(stats)
+x <- data %>% filter(reps == lows[[6]], t < 2000) %>% pull(x) %>% as.ts() 
+fit <- arima(x)
+y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
+y <- bind_rows(data.frame(pred = x, se = 0), y)
+arima_forecast <- data %>% filter(reps == lows[[6]]) %>% select(t,x) %>% bind_cols(y)
+```
+
+```{r arima_forecast}
+arima_forecast %>% ggplot(aes(t)) + 
+  geom_line(aes(y=x), col="white") +
+  geom_line(aes(y=pred), col="grey", lwd = 1) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5)
+
+```
+
+---
 layout: true
@@
 
-# A look at the model
+# A look at a candidate model
 
@@
 
-```{r fig.height=5.5}
+```{r theory_f, fig.height=5.5}
 p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, 
@@
 
-```{r echo=FALSE}
+```{r theory_U}
 theory(p) %>%
@@
 
-```{r echo=FALSE}
+```{r det}
 det <- det_sim(may,p) %>% mutate(reps=1) %>% filter(t < 3000)
@@
 
-```{r}
-data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
-## get tail cases for comparison
-ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
-lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
-highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
-```
-
-```{r}
+```{r stoch1}
 data %>% filter(reps == lows[[6]]) %>%
@@
 
-```{r}
+```{r stoch2}
 
@@
 
-```{r}
+```{r stoch_all}
 data %>% ggplot(aes(t,x, group=reps)) + 
@@
 
-```{r echo=FALSE}
+```{r ensemble}
 mean <- data %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
@@
 
-```{r echo=FALSE}
+```{r wtf}
 data %>% 
@@
 
-```{r echo=FALSE}
+```{r}
 ## det equib
@@
 
-```{r}
+```{r fpt}
 passage %>% ggplot(aes(tau)) + 
@@
 
-```{r}
+```{r disc_sims}
 discrete_sims <- read_csv("../data/discrete-sims.csv.xz")
-discrete_sims %>% filter(series %in% c("1", "2")) %>% ggplot(aes(time,state)) + geom_line(col="white", lwd=1.5) + facet_wrap(~ series)
+discrete_sims %>% filter(series %in% c("1", "2")) %>%
+  ggplot(aes(time,state)) + geom_line(col="white", lwd=1.5) + facet_wrap(~ series)
 ```
@@
 
-```{r fig.width=12.5, fig.height=7}
+```{r posteriors, fig.width=12.5, fig.height=7}
 label <- c("no switch", "switch", "no ghost")

### 2019-08-09 slides/boettiger.Rmd
WIP

@@
 
+```{r arima_data}
+arima_forecast %>% filter(t < 2000) %>%
+  ggplot(aes(t)) + 
+  geom_line(aes(y=x), col="white") +
+  geom_line(aes(y=pred), col="grey", lwd = 1) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5) + 
+  xlim(0,3000) + ylim(0, 1.6)+ ylab("x")
+
+```
+
+---
+
+# Simple ARIMA forecast:
+
+```{r arima_forecast1}
+arima_forecast %>% ggplot(aes(t)) + 
+  geom_line(aes(y=pred), col="grey", lwd = 1) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5) + 
+  xlim(0,3000) + ylim(0, 1.6) + ylab("x")
+```
+
+---
+
+# Simple ARIMA forecast:
+
 ```{r arima_forecast}
@@
   geom_line(aes(y=pred), col="grey", lwd = 1) + 
-  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5)
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5)  + 
+  xlim(0,3000) + ylim(0, 1.6)+ ylab("x")
 
 ```
+
+---
+
+# Statistical Issues
+
+---
+
+# Theoretical Issues
 
@@
 
-# A look at a candidate model
+### A look at a candidate model
 
@@
 
-```{r theory_f, fig.height=5.5}
+```{r theory_f, fig.height=5.2}
 p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, 
@@
 
-```{r posteriors, fig.width=12.5, fig.height=7}
+```{r posteriors, fig.width=13, fig.height=6}
 label <- c("no switch", "switch", "no ghost")
@@
 
-vp <- grid::viewport(width = 4, height = 0.8, x = 0.3, y = 0.7)
+vp <- grid::viewport(width = .5, height = 0.55, x = 0.35, y = 0.7)
 

### 2019-08-09 slides/pdf-slides.R

@@ -1,0 +1,10 @@
+#pagedown::chrome_print(here::here("slides/boettiger.Rmd"), timeout=3000)
+
+
+
+library(webshot)
+install_phantomjs()
+file_name <- paste0("file://", here::here("slides/boettiger.Rmd"))
+webshot(file_name, "slides.pdf")
+
+

### 2019-08-12 notebook/may-outbreak-ghost.Rmd
saveRDS

@@
 
-states <- seq(0,2, length=200)
+states <- seq(0,2, length=140)
 actions <- states
@@
 true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
-believe_a <- 0.29 ## believe there's an attractor
+believe_a <- 0.295 ## believe there's an attractor
 true_i <- which.min(abs(possible_a - true_a))
@@
 ```{r}
+set.seed(12345)
 sim <- mdp_learning(transition, reward, discount, 
@@
 ```{r}
+saveRDS(sim, "sim-ghost-learning.Rds")
 ```

### 2019-08-12 notebook/make.R
async

@@ -11,7 +11,8 @@ p3 <- callr::r_bg(function() rmarkdown::render("notebook/discrete-sims.Rmd") )
 
 
 
-p4 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost-2.Rmd") )
+p4 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost.Rmd") )
+p5 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost-2.Rmd") )
 
 
 callr::poll(list(p4), -1)

### 2019-08-12 slides/boettiger.Rmd

@@
 layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
 background-size: 120%
@@
 layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
 background-size: 150%
@@
 ---
-background-image: linear-gradient(to right, rgba(150, 150, 150, .6), rgba(150, 150, 150, .85)), url(img/royalsociety/reef-shark.jpg)
+background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
 background-size: 120%
@@
 label <- c("no switch", "switch", "no ghost")
-
+pal <- viridis::viridis_pal()(3)
 df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
@@
   geom_histogram(alpha=0.9) + 
-  geom_vline(aes(xintercept=0.27, col = "no switch"), col= viridis::viridis_pal()(3)[1], lwd=2) + 
-  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = viridis::viridis_pal()(3)[2], lwd=2) + 
+  geom_vline(aes(xintercept=0.27, col = "no switch"), col= pal[1], lwd=2) + 
+  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = pal[2], lwd=2) + 
   ggplot2::scale_fill_viridis_d()
@@
 layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/forest-fire.jpg)
+background-image: linear-gradient(to right, rgba(30, 30, 30, .8), rgba(30, 30, 30, .9)), url(img/royalsociety/forest-fire.jpg)
 background-size: 120%
@@
 
+---
+
 # Descision Theory
+### Management under uncertainty
+.pull-left[
+
+## Social
+
+### Utility fn
+
+]
+.pull-right[
+
+## Ecological
+
+### Transition fn
+
+]
 
@@
 
-### Management under uncertainty
+# Example systems
+
+.pull-left[
+## Outbreak
+
+![](img/flickr-vsmoothe-pine-beetle.jpg)
+
+]
+.pull-right[
+
+- manage at acceptable/endemic level?
+- supress entirely?
+
+]
 
 ---
+
+```{r}
+library(MDPtoolbox)
+library(sarsop) # remotes::install_github("boettiger-lab/sarsop")
+library(tidyverse) # for plotting
+library(mdplearning)
+
+damage <- 0.05
+control <- 1
+reward_fn <- function(x,h) - damage * x ^ 2 - control * h
+discount <- 0.98
+
+states <- seq(0,2, length=200)
+actions <- states
+observations <- states
+sigma_g <- 0.05
+sigma_m <- 0.0
+
+r <- 0.8
+K <- 1.53
+q <- 2
+b <- .2
+eps <- states[2]/10
+Tmax <- 100
+
+may <- function(a){  
+  function(x, h){ # May
+    # x <- pmax(x - h, 0)   # harvest then recruit   
+    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps - h
+  }
+}
+
+possible_a <- seq(.25, .34, by = 0.005)
+true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
+believe_a <- 0.31 ## believe there's a strong attractor
+true_i <- which.min(abs(possible_a - true_a))
+```
+
+**True model: weak ghost attractor**
+
+```{r true_a}
+f <- may(true_a)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + 
+  geom_line(col="white") +
+  geom_point(col="white") + 
+  geom_hline(aes(yintercept = 0),col="white")
+```
+
+---
+
+**Incorrect belief of bistable system**
+
+```{r a29}
+f <- may(0.29)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + 
+  geom_line(col="white") +
+  geom_point(col="white") + 
+  geom_hline(aes(yintercept = 0),col="white")
+```
+
+---
+
+**Incorrect belief of stronger bistable system**
+
+```{r a31}
+f <- may(believe_a)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + 
+  geom_line(col="white") +
+  geom_point(col="white") + 
+  geom_hline(aes(yintercept = 0), col="white")
+```
+
+---
+class: center, middle
+
+# Example management scenarios
+
+---
+
+```{r}
+m_true <- fisheries_matrices(states, actions, observations, reward_fn, 
+                        may(true_a), sigma_g, sigma_m, noise = "lognormal")
+
+m <- fisheries_matrices(states, actions, observations, reward_fn, 
+                        may(believe_a), sigma_g, sigma_m, noise = "lognormal")
+soln <- mdp_value_iteration(m$transition, m$reward, discount)
+opt_soln <- mdp_value_iteration(m_true$transition, m_true$reward, discount)
+
+```
+
+**no Intervention**
+
+```{r no_intervention}
+x0 <- which.min(abs(states - 0.1))
+no_policy <- numeric(length(states)) + 1
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = no_policy, x0 = x0, Tmax = 100)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) +
+  geom_point(col="white")+
+  geom_line(aes(time, action), col="grey", lwd=2) + ylim(0,1.4)
+```
+
+---
+
+**Mistaken ghost for alternate stable state**:
+
+```{r mistaken_ghost}
+x0 <- which.min(abs(states - 0.1))
+no_policy <- numeric(length(states)) + 1
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = soln$policy, x0 = x0, Tmax = 100)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) +
+  geom_point(col="white")+
+  geom_line(aes(time, action), col="grey", lwd=2) + ylim(0,1.4)
+```
+
+---
+
+## Adaptive managment: revising beliefs about the process
+
+---
+
+### Prior beliefs
+
+```{r priors}
+prior <- dnorm(possible_a, 0.29, 0.01)
+prior <- prior / sum(prior)
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill="white") +
+  geom_vline(aes(xintercept = true_a), col="white", lwd=1, lty=2) 
+```
+
+---
+
+### Learning too late
+
+---
+
+### More agnostic prior beliefs
+
+```{r priors_wide}
+prior <- dnorm(possible_a, 0.29, 0.01)
+prior <- prior / sum(prior)
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill="white") +
+  geom_vline(aes(xintercept = true_a), col="white", lwd=1, lty=2) 
+```
+
+---
+
+### Less learning, more caution
 
@@
 
+.pull-left[
+
+Image Credits:
+
+![](img/royalsociety.jpg) Photo Competition 
+]
+.pull-right[
+
 ![](img/hastings.jpg)
@@
 
-<!--
 - Lizzie Wolkovich
 - Henry Scharf
--->
+
+]

### 2019-08-12 slides/solarized.Rmd
solarized theme

--- 2019-08-12 slides/boettiger.Rmd
@@
   xaringan::moon_reader:
-    css: ["default"]
+    css: ["default","solarized-light.css"]
     lib_dir: libs
+    seal: false
     nature:
-      highlightStyle: github
+      highlightStyle: "solarized-light"
       highlightLines: true
@@
 
+# Conservation decision-making under long transients
+
+## Carl Boettiger
+
+## UC Berkeley
+---
+layout: true
+background-color: #fdf6e3
+class: center, top
+
+---
+
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(dev.args=list(bg="transparent"), 
-                      echo = FALSE, message=FALSE, warning=FALSE,
-                      fig.width=11, fig.height=6.5, cache = TRUE)
+knitr::opts_chunk$set(#dev.args=list(bg="transparent"), 
+                      echo = FALSE, 
+                      message=FALSE, 
+                      warning=FALSE,
+                      fig.width=11, 
+                      fig.height=6.5, 
+                      cache = TRUE)
 
@@
 library(ggthemes)
+theme_set(theme_solarized(base_size=26))
+scale_colour_discrete <- function(...) scale_colour_solarized()
+scale_fill_discrete <- function(...) scale_fill_solarized()
+pal <- solarized_pal()(3)
 
-theme_set(
- theme_bw(base_size=26) + 
-  theme(line = element_line(colour = "white", size = 1.5),
-        rect = element_rect(fill = "transparent", colour = "white"),
-        text = element_text(colour = "white"),
-        axis.text = element_text(colour = "white"),
-        axis.ticks = element_line(colour = "white", size = 1.5),
-        panel.border = element_rect(colour = "white", size = 1.5),
-        panel.background = element_rect(fill = "transparent"),
-        plot.background = element_blank(), 
-        legend.background = element_blank(), 
-        legend.key = element_rect(fill = "transparent", colour = "transparent"), 
-        strip.background = element_blank(),
-        strip.text = element_text(colour = "white"),
-        panel.grid = element_blank()))
+txtcolor <- "#586e75"
+```
 
+```{r}
 source("../R/ghost.R")
@@
 
+![](img/royalsociety/penguin-swim.jpg)
+
 ???
@@
 ---
-background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url(img/city-mit-license.jpg)
-background-size: 140%
-class: center, top, inverse
+
+<div style='float:left;width:50%'>
+  <img src='img/city-mit-license.jpg' style='transform:translate(-20%, -10%);height:100%;object-fit:cover;overflow:hidden'>
+</div>
 
@@
 ---
-name: polarbear
-background-image: linear-gradient(to right, rgba(150, 150, 150, .1), rgba(150, 150, 150, .4)), url(img/royalsociety/pensive-polarbear.jpg)
-background-size: 120%
-class: center, top, inverse
 
----
-template: polarbear
+<div style='float:left;width:50%'>
+  <img src='img/royalsociety/pensive-polarbear.jpg' style='transform:translate(-20%, -10%);height:900px;overflow:hidden'>
+</div>
 
 # ... a need for theory & process-based models
-
----
-layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/whale-shark.jpg)
-background-position: center
-background-size: 120%
-class: center, top, inverse
 
@@
 
---
-
-![](img/ghost-invert.png)
-
-<br/>
-Hastings et al. *Science* (2018)
-
 ---
 
-# Ghost Attractor
+# Ghost Attractors
 
@@
 data %>% filter(reps == lows[[6]]) %>%
-  ggplot(aes(t, x, group=reps)) + 
-  geom_line(lwd=2, col="white") + 
-  ylim(0,1.6)
+  ggplot(aes(t, x, group=reps, col=as.character(reps))) + 
+  geom_line(lwd=2, show.legend = FALSE) + 
+  ylim(0,1.6) + scale_color_solarized(accent="blue")
 ```
@@
   ggplot(aes(t)) + 
-  geom_line(aes(y=x), col="white") +
-  geom_line(aes(y=pred), col="grey", lwd = 1) + 
-  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5) + 
+  geom_line(aes(y=x),  lwd = 1, col=pal[1]) +
   xlim(0,3000) + ylim(0, 1.6)+ ylab("x")
@@
 arima_forecast %>% ggplot(aes(t)) + 
-  geom_line(aes(y=pred), col="grey", lwd = 1) + 
-  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5) + 
+  geom_line(aes(y=pred),  lwd = 1.5, col=pal[1]) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se),  alpha = 0.5, fill=pal[1]) + 
   xlim(0,3000) + ylim(0, 1.6) + ylab("x")
@@
 arima_forecast %>% ggplot(aes(t)) + 
-  geom_line(aes(y=x), col="white") +
-  geom_line(aes(y=pred), col="grey", lwd = 1) + 
-  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), fill="white", alpha = 0.5)  + 
+  geom_line(aes(y=x), col=pal[2], lwd=1) +
+  geom_line(aes(y=pred),  lwd = 1.5, col=pal[1]) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se),  alpha = 0.5, fill=pal[1])  + 
   xlim(0,3000) + ylim(0, 1.6)+ ylab("x")
@@
 # Theoretical Issues
-
----
-layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .9)), url(img/royalsociety/snake.jpg)
-background-size: 120%
-background-position: center
-class: left, top, inverse
 
@@
   ggplot(aes(x, potential)) + 
-  geom_line(lwd = 2, col="white")
+  geom_line(lwd = 2,col=txtcolor)
 
@@
   ggplot(aes(t, x)) + 
-  geom_line(lwd = 2, col="white")
+  geom_line(lwd = 2,col=txtcolor)
 ```
+
 ---
-layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin.jpg)
-background-size: 150%
-class: center, top, inverse
-background-position: left
 
 # Stochastic simulations
-
----
 
@@
   ggplot(aes(t, x, group=reps)) + 
-  geom_line(lwd=2, col="white") + 
+  geom_line(lwd=2,col=txtcolor) + 
   ylim(0,1.6)
@@
 
+# Stochastic simulations
+
 ```{r stoch2}
@@
   ggplot(aes(t, x, group=reps)) + 
-  geom_line(lwd=2, col="white") + 
+  geom_line(lwd=2,,col=txtcolor) + 
   ylim(0,1.6)
@@
 
+# Stochastic simulations
+
 ```{r stoch_all}
 data %>% ggplot(aes(t,x, group=reps)) + 
-  geom_line(alpha=0.02, color = "white")
+  geom_line(alpha=0.02,col=txtcolor)
 ```
@@
 
+# Stochastic simulations
+
 ```{r ensemble}
@@
 data %>% ggplot(aes(t,x, group=reps)) + 
-  geom_line(alpha=0.02, color = "white") + 
-  geom_line(data = mean, lwd=2, color = "white")
+  geom_line(alpha=0.02,col=txtcolor) + 
+  geom_line(data = mean, lwd=2,col=pal[1])
 ```
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/gentoo-penguin-wtf.jpg)
+
+# Stochastic simulations
 
@@
   ggplot(aes(t,x, group=reps)) + 
-  geom_line(alpha=0.02, color = "white") + 
-  geom_line(data = mean, lwd = 2, color = "white") +
-  geom_line(data = det, lwd = 2, color = rgb(.6, .6, 1)) 
+  geom_line(alpha=0.02,col=txtcolor) + 
+  geom_line(data = mean, lwd = 2, col=pal[1]) +
+  geom_line(data = det, lwd = 2, col=pal[2]) 
 ```
@@
 ---
+
+# Stochastic simulations
 
@@
 passage %>% ggplot(aes(tau)) + 
-  geom_histogram(binwidth = 200, fill=rgb(.9, .9, .9, .7)) + 
-  geom_vline(aes(xintercept=det_tau), col="blue", lwd=2)
+  geom_histogram(binwidth = 200, col=txtcolor) + 
+  geom_vline(aes(xintercept=det_tau), col=pal[1], lwd=2)
 ```
@@
 ---
-layout: false
-background-image: linear-gradient(to right, rgba(50, 50, 50, .7), rgba(50, 50, 50, .95)), url(img/royalsociety/green-crested-lizard.jpg)
-background-size: 110%
-class: left, top, inverse
-background-position: center
+class: left
 
@@
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .8), rgba(50, 50, 50, .9)), url(img/royalsociety/reef-shark.jpg)
-background-size: 120%
-class: top, inverse
-background-position: center
-layout: true
 
@@
 
----
+![](img/royalsociety/reef-shark.jpg)
 
@@
 discrete_sims %>% filter(series %in% c("1", "2")) %>%
-  ggplot(aes(time,state)) + geom_line(col="white", lwd=1.5) + facet_wrap(~ series)
+  ggplot(aes(time,state)) + 
+  geom_line(lwd=1.5, col=pal[1]) + 
+  facet_wrap(~ series)
 ```
@@
 class: left
+
+# Model Inference
 
@@
 
+# Model Inference
+
 ```{r posteriors, fig.width=13, fig.height=6}
 label <- c("no switch", "switch", "no ghost")
-pal <- viridis::viridis_pal()(3)
 df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
@@
   ggplot(aes(time,state, col=series)) + 
-  geom_line(lwd=1.5, show.legend = FALSE) + 
-  ggplot2::scale_color_viridis_d()
+  geom_line(lwd=1.5, show.legend = FALSE)
 
@@
   ggplot((aes(value, fill=timeseries))) + 
-  geom_histogram(alpha=0.9) + 
+  geom_histogram(alpha=0.9, show.legend = FALSE) + 
   geom_vline(aes(xintercept=0.27, col = "no switch"), col= pal[1], lwd=2) + 
-  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = pal[2], lwd=2) + 
-  ggplot2::scale_fill_viridis_d()
+  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = pal[2], lwd=2)
 
-vp <- grid::viewport(width = .5, height = 0.55, x = 0.35, y = 0.7)
+vp <- grid::viewport(width = .4, height = 0.45, x = 0.35, y = 0.7)
 
@@
 ```
-
----
-layout: true
-background-image: linear-gradient(to right, rgba(30, 30, 30, .8), rgba(30, 30, 30, .9)), url(img/royalsociety/forest-fire.jpg)
-background-size: 120%
-class: center, top, inverse
 
@@
 
-```{r}
+```{r results="hide"}
 library(MDPtoolbox)
@@
 true_i <- which.min(abs(possible_a - true_a))
+
 ```
 
-**True model: weak ghost attractor**
+## True model: weak ghost attractor
 
@@
   ggplot(aes(x, f)) + 
-  geom_line(col="white") +
-  geom_point(col="white") + 
-  geom_hline(aes(yintercept = 0),col="white")
+  geom_line(col=pal[1], lwd=2) +
+  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
+  ylim(-0.06, 0.06)
 ```
@@
   ggplot(aes(x, f)) + 
-  geom_line(col="white") +
-  geom_point(col="white") + 
-  geom_hline(aes(yintercept = 0),col="white")
+  geom_line(col=pal[1], lwd=2) +
+  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
+  ylim(c(-0.04, 0.04))
 ```
@@
   ggplot(aes(x, f)) + 
-  geom_line(col="white") +
-  geom_point(col="white") + 
-  geom_hline(aes(yintercept = 0), col="white")
+  geom_line(col=pal[1], lwd=2) +
+  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
+  ylim(-0.03, 0.03)
 ```
@@
 
-```{r}
+```{r results="hide"}
 m_true <- fisheries_matrices(states, actions, observations, reward_fn, 
@@
 
-**no Intervention**
+## Simulation under *no intervention*
 
@@
 df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) +
-  geom_point(col="white")+
-  geom_line(aes(time, action), col="grey", lwd=2) + ylim(0,1.4)
+  select(state, action, time) %>% 
+  gather(series, state, -time) %>%
+  ggplot(aes(time, state, col=series)) +
+  geom_line(lwd=2) + 
+  ylim(0,1.4)
 ```
@@
 df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) +
-  geom_point(col="white")+
-  geom_line(aes(time, action), col="grey", lwd=2) + ylim(0,1.4)
+select(state, action, time) %>% 
+  gather(series, state, -time) %>%
+  ggplot(aes(time, state, col=series)) +
+  geom_line(lwd=2) + 
+  ylim(0,1.4)
 ```
@@
 data.frame(a = possible_a, probability = prior) %>%
-  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill="white") +
-  geom_vline(aes(xintercept = true_a), col="white", lwd=1, lty=2) 
+  ggplot(aes(a,prior)) + geom_bar(stat="identity", col=txtcolor) +
+  geom_vline(aes(xintercept = true_a), lwd=1, lty=2, col=pal[1]) 
 ```
@@
 ```{r priors_wide}
-prior <- dnorm(possible_a, 0.29, 0.01)
+prior <- dnorm(possible_a, 0.3, 0.01)
 prior <- prior / sum(prior)
 data.frame(a = possible_a, probability = prior) %>%
-  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill="white") +
-  geom_vline(aes(xintercept = true_a), col="white", lwd=1, lty=2) 
+  ggplot(aes(a,prior)) + geom_bar(stat="identity", col=txtcolor) +
+  geom_vline(aes(xintercept = true_a), lwd=1, lty=2, col=pal[1]) 
 ```
@@
 
-### Less learning, more caution
+## Less learning, more caution
+
+```{r}
+```
 
 ---
-layout: true
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/pensive-polarbear.jpg)
-background-size: 120%
-class: center, top, inverse
+class: center, top
+background-image: url(img/royalsociety/bohemian-waxwing.jpg)
+background-size: 140%
+
+# Thanks!
 
 ---
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/eagle-ray.jpg)
-background-size: 120%
-class: center, top, inverse
-
----
-class: center, middle
-background-image: linear-gradient(to right, rgba(50, 50, 50, .6), rgba(50, 50, 50, .9)), url(img/royalsociety/bohemian-waxwing.jpg)
-background-size: 120%
 
@@
 
-![](img/royalsociety.jpg) Photo Competition 
+![](img/royalsociety.jpg)
+
+**Publishing Photo Competition**
+
+--
+
+- Lizzie Wolkovich
+- Henry Scharf
+
 ]
@@
 
-- Lizzie Wolkovich
-- Henry Scharf
-
 ]

### 2019-08-12 notebook/make.R
rerun

@@ -15,9 +15,9 @@ p4 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost.Rmd"
 p5 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost-2.Rmd") )
 
 
-callr::poll(list(p4), -1)
+callr::poll(list(p4,p5), -1)
 
 
 p4$read_output_lines()
 p4$get_result()
-p1$get_exit_status()
+p4$get_exit_status()

### 2019-08-12 slides/solarized.Rmd
tweaks

@@
 
-# Decisions vs transient dynamics: what could possibly go wrong?
+# What could possibly go wrong?
 
@@
 <div style='float:left;width:50%'>
-  <img src='img/city-mit-license.jpg' style='transform:translate(-20%, -10%);height:100%;object-fit:cover;overflow:hidden'>
+  <img src='img/city-slice.jpeg' style='transform:translate(-22%, -10%);height:100%;object-fit:cover;overflow:hidden'>
 </div>
@@
 <div style='float:left;width:50%'>
-  <img src='img/royalsociety/pensive-polarbear.jpg' style='transform:translate(-20%, -10%);height:900px;overflow:hidden'>
+  <img src='img/royalsociety/pensive-polarbear-slice.jpg' style='transform:translate(-22%, -10%);height:990px; width:1329; overflow:hidden'>
 </div>
@@
 ---
+class: left
 
@@
 ---
+class: left
 
 # Theoretical Issues
+
+--
+
+<img src="img/royalsociety/pensive-polarbear.jpg" width="400", style="float:center">
 
@@
 
-- Given time-series observations, can we distinguish ghost attractors from real attractors?
+# Model Inference
+
+Given time-series observations, *can we distinguish* **ghost** attractors from **real** ones?
 
@@
 
-Hierarchical Bayesian estimation of stability using TensorFlow(R)
+Hierarchical Bayesian estimation of stability
 
@@
 ```
+
+Using TensorFlow(R) via `greta` R package, <https://greta-stats.org>
 
@@
 
+# Model Inference
+
+(Full ensemble 500 length-3000 replicate timeseries)
+
+```{r}
+samples3 <- read_csv("../data/posterior_ensemble.csv.xz")
+samples3 %>%  
+  ggplot() + 
+  geom_histogram(aes(value), bins = 30, col=txtcolor)  +
+  geom_vline(aes(xintercept = 0.023), col = pal[1], lwd = 2)
+
+```
+
+---
+background-image: url(img/royalsociety/gentoo-penguin.jpg)
+background-position: left
+background-size: 140%
+class: center, top
+
 # Descision Theory
-### Management under uncertainty
+
+---
+
+# Descision Theory
+
 .pull-left[
@@
 ---
+class: middle
 
@@
 
-.pull-left[
-## Outbreak
+---
+background-image: url(img/royalsociety/forest-fire.jpg)
+background-position: center
+background-size: 120%
+class: center, top
 
-![](img/flickr-vsmoothe-pine-beetle.jpg)
+---
+background-image: url(img/royalsociety/methane_bubbles.jpg)
+background-position: center
+background-size: 100%
+class: center, top
 
-]
-.pull-right[
-
-- manage at acceptable/endemic level?
-- supress entirely?
-
-]
+---
+background-image: url(img/flickr-vsmoothe-pine-beetle.jpg)
+background-position: center
+background-size: 100%
+class: center, top
 
@@
 
-**Incorrect belief of bistable system**
+## Incorrect belief of bistable system
 
@@
 
-**Incorrect belief of stronger bistable system**
+## Incorrect belief of stronger bistable system
 
@@
 
-**Mistaken ghost for alternate stable state**:
+## Mistaken ghost for alternate stable state
 
@@
 
-```{r}
+```{r eval=FALSE}
+sim <- readRDS("sim-ghost-learning-2.Rds")
+
+ sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% seq(1,Tmax, by = 5)) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(param, probability, group = time, alpha = time)) + 
+  geom_bar()
+
+library(patchwork)
+p1 + p2
 ```
@@
 
---
+![](img/noaa.png)
 
-- Lizzie Wolkovich
-- Henry Scharf
+Lizzie Wolkovich
+
+Henry Scharf
 

### 2019-08-13 slides/solarized.Rmd
WIP

@@
 
+background-image: url(img/royalsociety/snake.jpg)
+background-position: center
+background-size: 120%
+class: center, top
+
 # Ghost Attractors
@@
 
+.pull-left[
+
+]
+.pull-right[
+<img src="img/royalsociety/ballons.jpg" width="400", style="float:center">
+]
+
 ---
@@
 
+.pull-left[
+
+... Issues run deeper: transients hard to infer  *even under idealized conditions*
+
 --
 
+#### *Stochastic* dynamics *qualitatively* alter the picture
+
+]
+.pull-right[
 <img src="img/royalsociety/pensive-polarbear.jpg" width="400", style="float:center">
+]
 
@@
 passage %>% ggplot(aes(tau)) + 
-  geom_histogram(binwidth = 200, col=txtcolor) + 
-  geom_vline(aes(xintercept=det_tau), col=pal[1], lwd=2)
+  geom_histogram(binwidth = 200, fill=txtcolor) + 
+  geom_vline(aes(xintercept=det_tau), col=pal[1], lwd=3)
 ```
@@
   geom_histogram(alpha=0.9, show.legend = FALSE) + 
-  geom_vline(aes(xintercept=0.27, col = "no switch"), col= pal[1], lwd=2) + 
-  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = pal[2], lwd=2)
+  geom_vline(aes(xintercept=0.27, col = "no switch"), col= pal[1], lwd=3) + 
+  geom_vline(aes(xintercept=0.28, col = "no ghost"), col = pal[2], lwd=3)
 
@@
   ggplot() + 
-  geom_histogram(aes(value), bins = 30, col=txtcolor)  +
-  geom_vline(aes(xintercept = 0.023), col = pal[1], lwd = 2)
+  geom_histogram(aes(value), bins = 30, fill=txtcolor)  +
+  geom_vline(aes(xintercept = 0.023), col = pal[1], lwd = 3)
 
@@
 
-# Descision Theory
+# Decision Theory
 
@@
 
-# Example management scenarios
+# Example Contexts
 
@@
                    policy = no_policy, x0 = x0, Tmax = 100)
+```
 
-df %>% mutate(state = states[state], action = actions[action]) %>% 
+```{r}
+df %>% mutate(state = states[state], action = 5*actions[action]) %>% 
   select(state, action, time) %>% 
@@
   geom_line(lwd=2) + 
-  ylim(0,1.4)
+  ylim(0,1.3)
 ```
@@
 
-df %>% mutate(state = states[state], action = actions[action]) %>% 
+df %>%
+  mutate(state = states[state], action = 5*actions[action]) %>% 
   select(state, action, time) %>% 
@@
 
-## Adaptive managment: revising beliefs about the process
+# Adaptive Management:
+
+### Updating beliefs
+
+<img src="img/royalsociety/waiting-in-the-shallows.jpg" width="500">
 
@@
 
-### Prior beliefs
+## Prior beliefs
 
 ```{r priors}
-prior <- dnorm(possible_a, 0.29, 0.01)
+prior <- dnorm(possible_a, 0.295, 0.005)
 prior <- prior / sum(prior)
 data.frame(a = possible_a, probability = prior) %>%
-  ggplot(aes(a,prior)) + geom_bar(stat="identity", col=txtcolor) +
-  geom_vline(aes(xintercept = true_a), lwd=1, lty=2, col=pal[1]) 
+  ggplot(aes(a,prior)) + 
+  geom_bar(stat="identity", fill=pal[1], alpha=0.6) +
+  geom_vline(aes(xintercept = true_a), lwd=3, lty=2, col=pal[1]) + 
+  ylim(0, 0.6)
 ```
@@
 
-### Learning too late
+## Learning late
+
+```{r}
+sim <- readRDS("../notebook/sim-ghost-learning.Rds")
+states <- seq(0,2, length=140)
+actions <- states
+Tmax <- 100
+
+ sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% seq(1,Tmax, by = 10), time < 50) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) 
+```
 
@@
 
-### More agnostic prior beliefs
+## Outcome
+
+```{r}
+ sim$df %>% 
+   mutate(state = states[state], action = 5*actions[action]) %>% 
+  select(-value, -obs) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, state, color = series)) +
+  geom_line(lwd=2) +
+  ylim(0,1.3) +
+  ylab("state / action")
+```
+
+---
+
+## More agnostic prior beliefs
 
 ```{r priors_wide}
-prior <- dnorm(possible_a, 0.3, 0.01)
+prior <- dnorm(possible_a, 0.3, 0.05)
 prior <- prior / sum(prior)
 data.frame(a = possible_a, probability = prior) %>%
-  ggplot(aes(a,prior)) + geom_bar(stat="identity", col=txtcolor) +
-  geom_vline(aes(xintercept = true_a), lwd=1, lty=2, col=pal[1]) 
+  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill=pal[1], alpha=0.5) +
+  geom_vline(aes(xintercept = true_a), lwd=3, lty=2, col=pal[2]) +
+  ylim(0,.3)
 ```
@@
 
-```{r eval=FALSE}
-sim <- readRDS("sim-ghost-learning-2.Rds")
+```{r}
+sim2 <- readRDS("../notebook/sim-ghost-learning-2.Rds")
+states <- seq(0,2, length=200)
+actions <- states
+Tmax <- 100
 
- sim$posterior %>% 
+ sim2$posterior %>% 
   data.frame(time = 1:Tmax) %>%
-  filter(time %in% seq(1,Tmax, by = 5)) %>%
+  filter(time %in% seq(1,Tmax, by = 10), time < 50) %>%
   gather(param, probability, -time, factor_key =TRUE) %>% 
   mutate(param = as.numeric(param)) %>% 
-  ggplot(aes(param, probability, group = time, alpha = time)) + 
-  geom_bar()
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+   ylim(0,0.3)
+```
 
-library(patchwork)
-p1 + p2
+---
+
+## Less learning, more caution
+
+```{r}
+ sim2$df %>% 
+   mutate(state = states[state], action = 5*actions[action]) %>% 
+  select(-value, -obs) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, state, color = series)) +
+  geom_line(lwd=2) + 
+  ylim(0,1.3) +
+  ylab("state / action")
 ```
+
+---
+
+# An emergent precautionary principle
+
+<!-- Better to be ignorant than wrong -->
 
@@
 
-**Publishing Photo Competition**
-
 ![](img/noaa.png)

### 2019-08-13 slides/solarized.Rmd
still going
slides and symlinks

@@
   xaringan::moon_reader:
-    css: ["default","solarized-light.css"]
+    css: ["default", "solarized-light.css"]
     lib_dir: libs
@@
 background-color: #fdf6e3
-class: center, top
+class: top
 
@@
 ---
+class: center
 
@@
   ggplot(aes(t)) + 
-  geom_line(aes(y=x),  lwd = 1, col=pal[1]) +
+  geom_line(aes(y=x),  lwd = 2, col=pal[1]) +
   xlim(0,3000) + ylim(0, 1.6)+ ylab("x")
@@
 
+- sufficient data
+
+- accurate measurements
+
+- appropriate model
+
+- numerical/computational issues
+
+- ...
+
 ]
+
 .pull-right[
@@
 
---
-
 #### *Stochastic* dynamics *qualitatively* alter the picture
@@
 ]
+
+--
+
+Knowing the correct model is not enough
 
@@
 ---
-class: left
+class: top, left
 
@@
 
+background-image: url(img/royalsociety/reef-shark.jpg)
+background-position: center
+background-size: 120%
+class: center, top
+
 # Challenges to Model Inference
-
-![](img/royalsociety/reef-shark.jpg)
 
@@
 label <- c("no switch", "switch", "no ghost")
-df <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
+discrete_posteriors <- c("../data/no_switches.csv.xz", "../data/switches.csv.xz", "../data/weak_ghost.csv.xz") %>%
   map_dfr(read_csv, .id = "timeseries") %>% 
@@
 
-p1 <- df %>% 
+p1 <- discrete_posteriors %>% 
   ggplot((aes(value, fill=timeseries))) + 
@@
 ---
-background-image: url(img/royalsociety/gentoo-penguin.jpg)
+background-image: url(img/chess.jpg)
 background-position: left
-background-size: 140%
-class: center, top
+background-size: 120%
+class: center, bottom
 
@@
 
-## Social
+## Societal Dynamics
 
-### Utility fn
+Utility function $U(x_t,a_t)$
+
+Given state $x_t$ and action $a_t$, balance
+costs and benefits.
+
+<img src="img/royalsociety/balance.jpg" width="300px">
 
 ]
+
+--
+
 .pull-right[
 
-## Ecological
+## Ecological Dynamics
 
-### Transition fn
+Transition function (forecast probabilities) $f(x,a)$
+
+<img src="solarized_files/figure-html/posteriors-1.png" width="400px">
 
@@
 ---
-class: middle
+class: middle, center
 
-# Example systems
+# Potential contexts
+
+--
+
+*Tipping point or ghost?*
 
@@
 ---
-class: center, middle
-
-# Example Contexts
-
----
 
@@
 
-<img src="img/royalsociety/waiting-in-the-shallows.jpg" width="500">
+<img src="img/royalsociety/waiting-in-the-shallows.jpg" width="450">
+
+--
+
+*Reinforcement Learning*
+
+--
+
+(Markov Decision Process over possible models with Bayesian updating)
 
@@
 
-```{r}
+```{r learning}
 sim <- readRDS("../notebook/sim-ghost-learning.Rds")
@@
 
-```{r}
+```{r less-learning}
 sim2 <- readRDS("../notebook/sim-ghost-learning-2.Rds")
@@
 ---
+class: middle
 
-# An emergent precautionary principle
+# An emergent precautionary principle?
 
@@
 ---
+class: top, left
+
+# Conclusions
+
+<img src="solarized_files/figure-html/arima_forecast-1.png" width="100px" height="60px"> Ghost attractors look like real attractor but behave differently
+
+--
+
+<img src="solarized_files/figure-html/wtf-1.png" width="100px" height="60px"> Stochastic ghosts $\neq$ deterministic ones
+
+--
+
+<img src="solarized_files/figure-html/posteriors-1.png" width="100px" height="60px"> This makes model inference hard!
+
+--
+
+<img src="img/royalsociety/forest-fire.jpg" width="100px" height="60px"> .. and can lead to very bad decisions!
+
+--
+
+<img src="solarized_files/figure-html/learning-1.png" width="100px" height="60px"> Reinforcement learning can help deal with uncertainty
+
+--
+
+<img src="solarized_files/figure-html/less-learning-1.png" width="100px" height="60px"> Precaution is sometimes better than knowledge
+
+---
 class: center, top
-background-image: url(img/royalsociety/bohemian-waxwing.jpg)
+background-image: url(img/royalsociety/penguin-swim.jpg)
 background-size: 140%
@@
 ]
+
+Code, slides, etc: <https://github.com/cboettig/decisions-vs-transients>
+
+---
+
+# Traces for single time-series
+
+```{r}
+discrete_posteriors %>% 
+  ggplot(aes(t, value, group=chain, col=chain)) + 
+  geom_line(show.legend = FALSE) +
+  facet_wrap(~timeseries)
+```
+
+---
+
+# Traces for the ensemble fit
+
+```{r}
+samples3 %>% 
+  ggplot(aes(t, value, group=chain, col=chain)) + 
+  geom_line(show.legend = FALSE) 
+```

### 2019-08-13 slides/gh-pages.R

@@ -1,0 +1,11 @@
+library(fs)
+library(here)
+
+fs::dir_create(here("docs"))
+
+slides <- here("slides/solarized.html")
+fs::link_create(slides, here("docs/index.html"))
+fs::link_create(here("slides/img"), here("docs/img"))
+fs::link_create(here("slides/libs"), here("docs/libs"))
+fs::link_create(here("slides/solarized_files"), here("docs/solarized_files"))
+fs::link_create(here("slides/solarized-light.css"), here("docs/solarized-light.css"))

### 2019-08-13 slides/solarized.Rmd
update slides, compress images

@@
 
+background-color: #fdf6e3
 background-image: url(img/royalsociety/penguin-swim.jpg)
@@
 
-## UC Berkeley
+### *UC Berkeley*
+
 ---
@@
 ---
+class: center, top
 
@@
 
-![](img/royalsociety/penguin-swim.jpg)
+<img src="img/royalsociety/penguin-swim.jpg" width="500px">
 

### 2019-08-13 slides/gh-pages.R
gh-pages no like symlink

@@ -4,7 +4,8 @@ library(here)
 fs::dir_create(here("docs"))
 
 slides <- here("slides/solarized.html")
-fs::link_create(slides, here("docs/index.html"))
+#fs::link_create(slides, here("docs/index.html"))
+fs::file_copy(slides, here("docs/index.html"), overwrite = TRUE)
 fs::link_create(here("slides/img"), here("docs/img"))
 fs::link_create(here("slides/libs"), here("docs/libs"))
 fs::link_create(here("slides/solarized_files"), here("docs/solarized_files"))

### 2019-08-13 gh-pages.R
re-knit

@@ -1,0 +1,12 @@
+library(fs)
+library(here)
+
+fs::dir_create(fs::path_rel("docs"))
+
+slides <- fs::path_rel("slides/solarized.html")
+fs::link_create(slides, fs::path_rel("docs/index.html"))
+#fs::file_copy(slides, fs::path_rel("docs/index.html"), overwrite = TRUE)
+fs::link_create(fs::path_rel("slides/img"), fs::path_rel("docs/img"))
+fs::link_create(fs::path_rel("slides/libs"), fs::path_rel("docs/libs"))
+fs::link_create(fs::path_rel("slides/solarized_files"), fs::path_rel("docs/solarized_files"))
+fs::link_create(fs::path_rel("slides/solarized-light.css"), fs::path_rel("docs/solarized-light.css"))

### 2019-08-13 slides/gh-pages.R
mv

@@ -1,12 +1,0 @@
-library(fs)
-library(here)
-
-fs::dir_create(here("docs"))
-
-slides <- here("slides/solarized.html")
-#fs::link_create(slides, here("docs/index.html"))
-fs::file_copy(slides, here("docs/index.html"), overwrite = TRUE)
-fs::link_create(here("slides/img"), here("docs/img"))
-fs::link_create(here("slides/libs"), here("docs/libs"))
-fs::link_create(here("slides/solarized_files"), here("docs/solarized_files"))
-fs::link_create(here("slides/solarized-light.css"), here("docs/solarized-light.css"))

### 2019-08-13 gh-pages.R
try again w/o symlink

@@ -4,8 +4,8 @@ library(here)
 fs::dir_create(fs::path_rel("docs"))
 
 slides <- fs::path_rel("slides/solarized.html")
-fs::link_create(slides, fs::path_rel("docs/index.html"))
-#fs::file_copy(slides, fs::path_rel("docs/index.html"), overwrite = TRUE)
+#fs::link_create(slides, fs::path_rel("docs/index.html"))
+fs::file_copy(slides, fs::path_rel("docs/index.html"), overwrite = TRUE)
 fs::link_create(fs::path_rel("slides/img"), fs::path_rel("docs/img"))
 fs::link_create(fs::path_rel("slides/libs"), fs::path_rel("docs/libs"))
 fs::link_create(fs::path_rel("slides/solarized_files"), fs::path_rel("docs/solarized_files"))

### 2019-08-13 docs/index.Rmd
try renaming for gh-pages hosting

--- 2019-08-13 slides/solarized.Rmd
@@
 
-<img src="solarized_files/figure-html/posteriors-1.png" width="400px">
+<img src="index_files/figure-html/posteriors-1.png" width="400px">
 
@@
 
-<img src="solarized_files/figure-html/arima_forecast-1.png" width="100px" height="60px"> Ghost attractors look like real attractor but behave differently
+<img src="index_files/figure-html/arima_forecast-1.png" width="100px" height="60px"> Ghost attractors look like real attractor but behave differently
 
@@
 
-<img src="solarized_files/figure-html/wtf-1.png" width="100px" height="60px"> Stochastic ghosts $\neq$ deterministic ones
+<img src="index_files/figure-html/wtf-1.png" width="100px" height="60px"> Stochastic ghosts $\neq$ deterministic ones
 
@@
 
-<img src="solarized_files/figure-html/posteriors-1.png" width="100px" height="60px"> This makes model inference hard!
+<img src="index_files/figure-html/posteriors-1.png" width="100px" height="60px"> This makes model inference hard!
 
@@
 
-<img src="solarized_files/figure-html/learning-1.png" width="100px" height="60px"> Reinforcement learning can help deal with uncertainty
+<img src="index_files/figure-html/learning-1.png" width="100px" height="60px"> Reinforcement learning can help deal with uncertainty
 
@@
 
-<img src="solarized_files/figure-html/less-learning-1.png" width="100px" height="60px"> Precaution is sometimes better than knowledge
+<img src="index_files/figure-html/less-learning-1.png" width="100px" height="60px"> Precaution is sometimes better than knowledge
 

### 2019-08-14 notebook/ghost-sims.Rmd
updated ghost sims

@@
 ```{r}
-p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2, N = 1e4)
+p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 3e3, x0 = 0.2)
 growth <- function(x, p) x * p$r * (1 - x / p$K)
@@
 det <- det_sim(may,p) 
-det %>% ggplot(aes(t, x)) + geom_line() + theme_bw()
+det %>% ggplot(aes(t, x)) + geom_line() 
 ```
@@
 
+## Stochastic ensemble mean vs deterministic mean
+
 ```{r}
 df %>% ggplot(aes(t, x, group=reps)) + 
-  geom_line(alpha=0.1) +
+  geom_line(alpha=0.05) +
   geom_line(data = det, color = "blue") + 
@@
 ```
+
+## Deterministic system with initial conditions in neighborhood below ghost
+
+```{r}
+lambda <- function(x0, p){
+                      pars <- p
+                      p$x0 <- x0
+                      det_sim(may, p)
+                    }
+mu <- 0.02
+init <- mu + rnorm(100, 0, .02 * mu)
+df <- future_map_dfr(init, lambda, p = p, .id = "reps")
+
+df %>% ggplot(aes(t, x, group=reps)) + 
+  geom_line(alpha=0.1)
+```
+
+## Deterministic System with initial condition in neighborhood of ghost
+
+```{r}
+mu <- 0.55
+init <- mu + rnorm(100, 0, .02 * mu)
+df <- future_map_dfr(init, lambda, p = p, .id = "reps")
+
+df %>% ggplot(aes(t, x, group=reps)) + 
+  geom_line(alpha=0.1)
+```
+
+```{r}
+```

### 2019-08-15 notebook/ghost-fit-forecast.Rmd
a few additional explorations

--- 2019-08-06 notebook/ghost-fit.Rmd
@@
 ```{r}
-data <- read_csv("../data/reps.csv.xz") # %>% filter(t %in% seq(0,4000, by=4))
+data <- read_csv("../data/reps.csv.xz") %>% filter(t < 2000)
 data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.02)
@@
 lows <- means %>% dplyr::slice(1:10) %>% pull(reps)
-highs <- means %>% dplyr::slice((n()-10):n()) %>% pull(reps)
+highs <- means %>% dplyr::slice((n()-9):n()) %>% pull(reps)
+write_csv(data.frame(lows=lows,highs=highs), "../data/extremes.csv.xz")
+
 coding <- function(reps) case_when(
@@
 )
-extremes <- data %>% mutate(coding = coding(reps)) %>% na.omit()
-
-write_csv(extremes, "../data/extremes.csv.xz")
-extremes %>% ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)
+data %>% 
+  mutate(coding = coding(reps)) %>% 
+  na.omit() %>% 
+ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)
 
@@
   facet_wrap(~variable, scales = "free")
+
+write_csv(samples, "../data/posterior_highs.csv.xz")
+
 ```
@@
   facet_wrap(~variable, scales = "free")
+
+write_csv(samples2, "../data/posterior_lows.csv.xz")
 ```
@@
   facet_wrap(~variable, scales = "free")
+
+write_csv(samples3, "../data/posterior_ensemble.csv.xz")
 ```

### 2019-08-21 manuscript/manuscript.Rmd
placeholder draft

---
title: "Long transients can mislead model inference and decision making -- and what to do about it"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
TBD
journal: TBD
layout: 5p
bibliography: refs.bib
output: rticles::elsevier_article

---

```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = 1,  dev = "cairo_pdf", fig.width=4.5, fig.height=3.5)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
extrafont::loadfonts()
```

```{r message = FALSE}
library(tidyverse)   # plotting and data manipulation
```

Periods of long transient dynamics have long been identified in the ecological systems.   Ecological @Hastings2018

## Long transients when the underlying model is not known

One of the more intuitive implications of the existence of long transients is the inherent risk in mistaking them for stable state dynamics.  For example, Figure \ref{armina} illustrates how a transient

### Stochastic dynamics and transients

Stochastic dynamics can weaken the average influence of a ghost attractor, but long transients are still likely.  Stochastic dynamics can also make weak attractors behave essentially as ghost attractors.

# Methods

Stochastic dynamic programming solutions were computed in R [@R] using value iteration using the package `MDPToolbox` [@Marescot2013; @MDPtoolbox].  Code to reproduce this analysis is available at <https://github.com/cboettig/decisions-vs-transients>

\pagebreak

# References

### 2019-09-23 manuscript/appendix/appendix.Rmd
WIP
Need to revise & consolidate notebooks

---
title: "Appendix"
author: "Carl Boettiger"
output: github_document
---

```{r message = FALSE}
library(tidyverse)
library(patchwork)
library(furrr)
set.seed(12345)
```

## Model Definition

```{r}
p <- list(r = .05, K = 200, Q = 5, H = 38, sigma = .02, a=2.3, N = 3e3, x0 = 20)
```

```{r}
growth <- function(x, p) x * p$r * (1 - x / p$K)
consumption <- function(x, p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
may <- function(x, p) x + growth(x,p) - consumption(x,p)
```

```{r}
theory <- 
  function(p){
    tibble(x= seq(0,p$K, length.out = 100)) %>%
    mutate(growth = growth(x,p), 
           consumption = consumption(x,p)) %>%
    mutate(potential = - cumsum(growth - consumption)) %>%
    gather(curve, y, -x, -potential) 
}
```

Plots of the growth rate functions and the potential well diagram

```{r}
fig_a <- theory(p) %>%
  ggplot(aes(x, y, col=curve)) +
  geom_line(lwd=1) + 
  labs(title = "Population dynamics") +
  theme(legend.position = c(.5, .2), legend.background = element_blank())

fig_b <- theory(p) %>%
  ggplot(aes(x, potential)) + 
  geom_line(lwd=1)  + 
  labs(title = "Potential Well Diagram")

fig_a + fig_b
```

## Model simulations

```{r}
# Generic simulator routine
sim <- function(f, p){
  x <- numeric(p$N)
  x[1] <- p$x0
  dBt <- numeric(p$N)
  if(p$sigma > 0) dBt <- rnorm(p$N, 0, p$sigma)
  for(t in 1:(p$N-1)){
    x[t+1] <- max(f(x[t], p) + x[t] * dBt[t], 0)
  }
  data.frame(t = 1:p$N, x)
}
```

Deterministic simulations:

```{r}
p_det <- p
p_det$sigma <- 0
det <- sim(may,p_det) 
det %>% ggplot(aes(t, x)) + geom_line() 
```

Compare deterministic simulations across a range of `a` values:

```{r}
a <- c(2.25, 2.30, 2.305, 2.31)
names(a) <- a
p_det$N <- 1e4
df <- map_dfr(a, function(a){
  p_det$a <- a
  sim(may, p_det)
},
.id = "a")

df %>% ggplot(aes(t,x, col = a)) + geom_line() + coord_cartesian(xlim=c(0,5000))

```

Stochastic replicate simulations (run in parallel):

```{r}
plan("multisession")
system.time({
  df <- future_map_dfr(1:500, function(reps) sim(may, p), .id = "reps")
})

```

```{r}
p_det$N <- 3e3
det <- sim(may,p_det) %>% mutate(reps=1)
mean <- df %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
```

## Stochastic ensemble mean vs deterministic mean

```{r}
df %>% group_by(reps) %>%
  sample_n(5e2) %>%
  ggplot(aes(t, x, group=reps)) + 
  geom_line(alpha=0.05) +
  geom_line(data = det, color = "blue") + 
  geom_line(data = mean, color = "red")
```

---

# ARIMA Forecast

```{r}
library(tidyverse)
library(stats)
```

Extract a series with relatively long ghost:

```{r}
means <- df %>% group_by(reps) %>% summarise(ave = mean(x), max = max(x))
lows <- means %>% filter(ave < mean(ave) - 3*sd(ave), max > 120) %>% pull(reps)
```

```{r arima_forecast}
x <- df %>% filter(reps == lows[[1]], t <= 2000) %>% pull(x) %>% as.ts() 
fit <- arima(x)
y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
y <- bind_rows(data.frame(pred = x, se = 0), y)

arima_forecast <- df %>% 
  filter(reps == lows[[1]]) %>% 
  select(t,x) %>%
  bind_cols(y)
```

```{r arima_figure, message=FALSE, warning=FALSE}
train <- arima_forecast %>% filter(t < 2000) 
test <- arima_forecast %>% filter(t > 2000) 
train %>%
  ggplot(aes(t)) + 
  geom_line(aes(y=x)) +
  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
  xlim(0,3000) + ylab("x") + theme_minimal()

```

----

# Model-based forecast

```{r}
library(greta)
```

Bayesian estimate on training data only:

```{r}
wide <- select(train, x) %>% as.matrix() 
n <- dim(wide)[1]
x_t1 <- wide[-1,]
x_t <- wide[-n,] 
```

```{r}
r <- p$r
Q <- p$Q
K <- p$K
# a <- 2.3
H <- p$H
sigma <- p$sigma

# r <- uniform(0, 1)
# Q <- uniform(0, 10)
#K <- uniform(0, 10)
a <- uniform(0, 10)
# H <- uniform(0,  2)
#sigma <- uniform(0, 1)

# Model   (mean <-  may(x_t, p))
mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
distribution(x_t1) <- normal(mean, sigma * x_t)
m <- model(a)
```

```{r }
system.time({
  draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
})
```

```{r}
samples <-  
  map_dfr(draws, 
          function(x) data.frame(x, t = 1:dim(x)[1]), 
          .id = "chain") %>% 
  gather(variable, value, -t, -chain)
```

```{r}
#Q = 5
true <- 
  data.frame(a = 2.3) %>%
  gather(variable, value)
```

```{r}
samples %>% ggplot() + 
  geom_histogram(aes(value), bins = 30)  +
  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
  facet_wrap(~variable, scales = "free")

```

Replicate simulations of stochastic model with parameters drawn from posteriors

(would be great to have a `greta` method for this...)

```{r}
a <- sample(unlist(draws, TRUE), 10)
p$N <- 3e3
posterior_sims <- map_dfr(a, function(a){
  p$a <- a
  sim(may, p)
},
.id = "a")

```

```{r}
posterior_sims %>% 
  mutate(a = as.numeric(a)) %>% 
  ggplot(aes(t,x, group=a)) + 
  geom_line(alpha = 0.05) 
```

```{r}
```

----

### 2019-09-23 manuscript/manuscript.Rmd
WIP
Need to revise & consolidate notebooks

@@
 ---
-title: "Long transients can mislead model inference and decision making -- and what to do about it"
+title: "Consequences of long transients in model inference and optimal control"
 date: "`r Sys.Date()`"
@@
 abstract: |
-TBD
-journal: TBD
+  I explore the impact of long transient dynamics for model inference and optimal
+journal: Theoretical ecology
 layout: 5p
@@
 ```{r include = FALSE}
-knitr::opts_chunk$set(echo = FALSE, cache = 1,  dev = "cairo_pdf", fig.width=4.5, fig.height=3.5)
+knitr::opts_chunk$set(echo = FALSE, cache = 1,  
+                      dev = "cairo_pdf", fig.width=4.5, 
+                      fig.height=3.5, message = FALSE,
+                      warning = FALSE)
 library(hrbrthemes)
 library(ggplot2)
+library(ggthemes)
 library(Cairo)
@@
 
-Periods of long transient dynamics have long been identified in the ecological systems.   Ecological @Hastings2018
+```{r}
+source("../R/ghost.R")
+```
+
+```{r}
+data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
+## get tail cases for comparison
+ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
+lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
+highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
+```
+
+```{r arima_forecast}
+library(stats)
+x <- data %>% filter(reps == lows[[6]], t < 2000) %>% pull(x) %>% as.ts() 
+fit <- arima(x)
+y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
+y <- bind_rows(data.frame(pred = x, se = 0), y)
+arima_forecast <- data %>% filter(reps == lows[[6]]) %>% select(t,x) %>% bind_cols(y)
+```
+
+```{r arima_figure, message=FALSE, warning=FALSE}
+train <- arima_forecast %>% filter(t < 2000) 
+test <- arima_forecast %>% filter(t > 2000) 
+train %>%
+  ggplot(aes(t)) + 
+  geom_line(aes(y=x)) +
+  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
+  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
+  xlim(0,3000) + ylim(0, 1.6)+ ylab("x") + theme_minimal()
+
+```
+
+Transient dynamics have long been identified in the ecological systems @Hastings1994.   Ecological @Hastings2018
 

### 2019-09-23 notebook/ghost-fit-forecast.Rmd
WIP
Need to revise & consolidate notebooks

@@
 ```{r}
-data <- read_csv("../data/reps.csv.xz") %>% filter(t < 2000)
-data %>% ggplot(aes(t,x, group=reps)) + geom_line(alpha=0.02)
+source("../R/ghost.R")
 ```
 
-Divide this data into several groups based on
-
 ```{r}
-means <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
-
-lows <- means %>% dplyr::slice(1:10) %>% pull(reps)
-highs <- means %>% dplyr::slice((n()-9):n()) %>% pull(reps)
-write_csv(data.frame(lows=lows,highs=highs), "../data/extremes.csv.xz")
-
-coding <- function(reps) case_when(
-  reps %in% lows ~ "low",
-  reps %in% highs ~ "high"
-)
-data %>% 
-  mutate(coding = coding(reps)) %>% 
-  na.omit() %>% 
-ggplot(aes(t, x, group=reps, color = coding)) + geom_line(alpha=0.5)
-
-#means %>% ggplot(aes(ave)) + geom_histogram(binwidth = 0.02)
-
+data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
+## get tail cases for comparison
+ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
+lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
+highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
 ```
 
-## highs
+```{r arima_forecast}
+library(stats)
+x <- data %>% filter(reps == lows[[6]], t < 2000) %>% pull(x) %>% as.ts() 
+fit <- arima(x)
+y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
+y <- bind_rows(data.frame(pred = x, se = 0), y)
+arima_forecast <- data %>% filter(reps == lows[[6]]) %>% select(t,x) %>% bind_cols(y)
+```
 
-```{r}
-wide <- data %>% 
-  filter(reps %in% highs) %>%
-  spread(reps, x) %>%
-  select(-t) %>% 
-  as.matrix()
-n <- dim(wide)[1]
+```{r arima_figure, message=FALSE, warning=FALSE}
+train <- arima_forecast %>% filter(t < 2000) 
+test <- arima_forecast %>% filter(t > 2000) 
+train %>%
+  ggplot(aes(t)) + 
+  geom_line(aes(y=x)) +
+  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
+  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
+  xlim(0,3000) + ylim(0, 1.6)+ ylab("x") + theme_minimal()
 
@@
 ```{r}
+wide <- select(train, x) %>% as.matrix() 
+n <- dim(wide)[1]
 x_t1 <- wide[-1,]
@@
 
-write_csv(samples, "../data/posterior_highs.csv.xz")
-
 ```
-
-----
-
-## lows
-
-```{r}
-wide <- data %>% 
-  filter(reps %in% lows) %>%
-  spread(reps, x) %>%
-  select(-t) %>% 
-  as.matrix()
-n <- dim(wide)[1]
-
-```
-
-```{r}
-x_t1 <- wide[-1,]
-x_t <- wide[-n,] 
-```
-
-```{r}
-a <- uniform(0, 1)
-mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
-distribution(x_t1) <- normal(mean, sigma * x_t)
-m <- model(a)
-```
-
-```{r }
-system.time({
-  draws2 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
-})
-```
-
-```{r}
-bayesplot::mcmc_trace(draws2)
-```
-
-```{r}
-samples2 <-  
-  map_dfr(draws2, 
-          function(x) data.frame(x, t = 1:dim(x)[1]), 
-          .id = "chain") %>% 
-  gather(variable, value, -t, -chain)
-
-samples2 %>% ggplot() + 
-  geom_histogram(aes(value), bins = 30)  +
-  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
-  facet_wrap(~variable, scales = "free")
-
-write_csv(samples2, "../data/posterior_lows.csv.xz")
-```
-
-## Full ensemble
-
-```{r}
-wide <- data %>% 
-  spread(reps, x) %>%
-  select(-t) %>% 
-  as.matrix()
-n <- dim(wide)[1]
-
-```
-
-```{r}
-x_t1 <- wide[-1,]
-x_t <- wide[-n,] 
-```
-
-```{r}
-a <- uniform(0, 1)
-mean <- x_t + r * x_t * (1 - x_t / K) - a * x_t ^ Q / (x_t ^ Q + H ^ Q)
-distribution(x_t1) <- normal(mean, sigma * x_t)
-m <- model(a)
-```
-
-```{r }
-system.time({
-  draws3 <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 6, verbose = FALSE)
-})
-```
-
-```{r}
-bayesplot::mcmc_trace(draws3)
-```
-
-```{r}
-samples3 <-  
-  map_dfr(draws3, 
-          function(x) data.frame(x, t = 1:dim(x)[1]), 
-          .id = "chain") %>% 
-  gather(variable, value, -t, -chain)
-
-samples3 %>% ggplot() + 
-  geom_histogram(aes(value), bins = 30)  +
-  geom_vline(data = true, aes(xintercept = value), col = "red", lwd = 1) + 
-  facet_wrap(~variable, scales = "free")
-
-write_csv(samples3, "../data/posterior_ensemble.csv.xz")
-```
-
-```{r }
-p <- list(r = .05, K = 2, Q = 5, H = .38, sigma = .02, a=0.023, N = 4e3, x0 = 0.2, N = 1e4)
-source("../R/ghost.R")
-p$a <- 0.021
-tf <- theory(p) %>% mutate(a = 0.021)
-
-p$a <- 0.023
-tf <- theory(p) %>% mutate(a = 0.023) %>% bind_rows(tf)
-
-p$a <- 0.024
-tf <- theory(p) %>% mutate(a = 0.024) %>% bind_rows(tf)
-
-tf <- tf %>% mutate(a = as.factor(a))
-
-tf %>%
-  ggplot(aes(x, potential, col=a)) +
-  geom_line(lwd=1)
-
-```
-
-```{r}
-```

### 2019-09-23 notebook/gillespie.Rmd
WIP
Need to revise & consolidate notebooks

```{r}
library(tidyverse)
library(furrr)
library(adaptivetau)
```

```{r}
set.seed(1234) # set random number generator seed to be reproducible

p <- list(
  r = .05,
  K = 200,
  Q = 5,
  H = 38,
  a = 2.3,
  x0 = 20,
  Tmax = 3000
)

transitions = list(c(n = +1), # birth event
                   c(n = -1)) # death event
growth <- function(x, p)
  x * p$r * (1 - x / p$K)
consumption <- function(x, p)
  p$a * x ^ p$Q / (x ^ p$Q + p$H ^ p$Q)

rates <- function(state, params, t) {
  c(growth(state[["n"]], params),
    consumption(state[["n"]], params)
   )
}
```

Run simulation
```{r}
gillespie <- function(reps) adaptivetau::ssa.exact(
      init.values = c(n = p$x0),
      transitions,
      rates,
      params = p,
      tf = p$Tmax
    ) %>% as_tibble()
```

Test a single replicate:
```{r}
sim <- gillespie(1)
sim %>% ggplot(aes(time, n)) + geom_path()
```

## Stochastic replicates

```{r}
plan("multisession")
df <- future_map_dfr(1:500, gillespie, .id = "reps")
```

Deterministic sim
```{r}
dt <- 0.01
x <- numeric(p$Tmax/dt)
x[1] <- p$x0
for(t in 1:(p$Tmax/dt-1)){
    x[t+1] <- (growth(x[t], p) - consumption(x[t], p)) * dt + x[t]
}
det <- tibble(time = seq(dt,p$Tmax,by=dt), n = x, reps = 1)
```

```{r}
## Stochastic means -- nope, not sampled evenly
## mean <- df %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
```

<!-- Re-sample df at uniform intervals! -->

```{r}
df %>% 
  group_by(reps) %>% sample_n(1000) %>%
  ggplot(aes(time, n, group=reps)) +
  geom_line(alpha=0.03) +
  geom_line(data = sample_n(det,1000), col="blue")

```

```{r}
#write_csv(df, "data/gillespie.csv.xz")
```

### 2019-09-24 manuscript/appendix/appendix.Rmd
plot both prediction and true replicates

@@
 library(patchwork)
+library(ggthemes)
 library(furrr)
@@
 system.time({
-  df <- future_map_dfr(1:500, function(reps) sim(may, p), .id = "reps")
+  sim_data <- future_map_dfr(1:500, function(reps) sim(may, p), .id = "reps")
 })
@@
 det <- sim(may,p_det) %>% mutate(reps=1)
-mean <- df %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
+mean <- sim_data %>% group_by(t) %>% summarise(x = mean(x)) %>% mutate(reps = 1)
 ```
@@
 ```{r}
-df %>% group_by(reps) %>%
+sim_data %>% group_by(reps) %>%
   sample_n(5e2) %>%
@@
 ```{r}
-means <- df %>% group_by(reps) %>% summarise(ave = mean(x), max = max(x))
+means <- sim_data %>% group_by(reps) %>% summarise(ave = mean(x), max = max(x))
 lows <- means %>% filter(ave < mean(ave) - 3*sd(ave), max > 120) %>% pull(reps)
@@
 ```{r arima_forecast}
-x <- df %>% filter(reps == lows[[1]], t <= 2000) %>% pull(x) %>% as.ts() 
+x <- sim_data %>% filter(reps == lows[[1]], t <= 2000) %>% pull(x) %>% as.ts() 
 fit <- arima(x)
@@
 
-arima_forecast <- df %>% 
+arima_forecast <- sim_data %>% 
   filter(reps == lows[[1]]) %>% 
@@
 ```{r}
-a <- sample(unlist(draws, TRUE), 10)
-p$N <- 3e3
+a <- unname(sample(unlist(draws, TRUE), 100))
+p$N <- 1e3
+x <- sim_data %>% filter(reps == lows[[1]], t == 2001) %>% pull(x)
+
+p$x0 <- x[[1]]
 posterior_sims <- map_dfr(a, function(a){
@@
 },
-.id = "a")
+.id = "reps")
 
@@
 ```{r}
+alpha <- 0.1
+training_data <- train %>% select(t,x) %>% mutate(reps="1", set = "training", alpha = 1)
+
+true_forecast <- 
+  future_map_dfr(1:100, 
+                 function(reps) sim(may, p), 
+                 .id = "reps"
+                 ) %>% 
+  mutate(t = t+2000,
+         set = "true",
+         alpha = alpha) %>% 
+  bind_rows(training_data)
+
+predicted_forecast <- posterior_sims %>% 
+    mutate(t = t+2000,
+           set = "predicted",
+           alpha = alpha)
+```
+
+```{r}
+bind_rows(predicted_forecast, true_forecast) %>% 
+  ggplot(aes(t,x)) +
+  geom_line(aes(group = interaction(reps, set), col = set), alpha = .05) + 
+  geom_line(data = training_data) + 
+  scale_color_solarized()
+```
+
+```{r}
+
 posterior_sims %>% 
-  mutate(a = as.numeric(a)) %>% 
-  ggplot(aes(t,x, group=a)) + 
-  geom_line(alpha = 0.05) 
+  mutate(a = as.numeric(a), 
+         t = t+2000) %>% 
+  ggplot(aes(t,x)) + 
+  geom_line(aes(group = reps), col = "blue", alpha = 0.05,
+            data = true_forecast) + 
+  geom_line(aes(group = a), alpha = 0.05, col = "red") + 
+  geom_line(data = training_set) 
 ```

### 2019-09-24 manuscript/appendix/appendix.Rmd
with faceting

@@
 alpha <- 0.1
-training_data <- train %>% select(t,x) %>% mutate(reps="1", set = "training", alpha = 1)
+training_data <- train %>% select(t,x) %>% mutate(reps="1") #, set = "training")
 
@@
                  ) %>% 
-  mutate(t = t+2000,
-         set = "true",
-         alpha = alpha) %>% 
-  bind_rows(training_data)
+  mutate(t = t+2000) %>% 
+  bind_rows(training_data) %>% 
+  mutate(set = "true")
 
-predicted_forecast <- posterior_sims %>% 
-    mutate(t = t+2000,
-           set = "predicted",
-           alpha = alpha)
+predicted_forecast <- 
+  posterior_sims %>% 
+    mutate(t = t+2000) %>%
+    bind_rows(training_data) %>% 
+    mutate(set = "predicted")
 ```
@@
 bind_rows(predicted_forecast, true_forecast) %>% 
+#predicted_forecast %>%
   ggplot(aes(t,x)) +
@@
   geom_line(data = training_data) + 
+  facet_wrap(~set) + 
   scale_color_solarized()

### 2019-09-26 manuscript/appendix/appendix.Rmd
wip

@@
 
-# r <- uniform(0, 1)
-# Q <- uniform(0, 10)
-#K <- uniform(0, 10)
+r <- uniform(0, 2 * p$r)
+Q <- uniform(0, 10)
+K <- uniform(0, 2*p$K)
 a <- uniform(0, 10)
-# H <- uniform(0,  2)
-#sigma <- uniform(0, 1)
+H <- uniform(0,  2*p$H)
+sigma <- uniform(0, 2 * p$sigma)
 
@@
 ```{r}
-a <- unname(sample(unlist(draws, TRUE), 100))
+#a <- unname(sample(unlist(draws, TRUE), 100))
+x0 <- sim_data %>% filter(reps == lows[[1]], t == 2001) %>% pull(x)
+
+posterior_samples <- 
+  bind_rows(map(draws, as_tibble)) %>% 
+  sample_n(100)
+
+posterior_samples %>%
+  mutate(N=1e3, x0 = x0) %>%
+  purrr::transpose() %>% map_dfr()
+
 p$N <- 1e3
-x <- sim_data %>% filter(reps == lows[[1]], t == 2001) %>% pull(x)
 
@@
 ```{r}
-
-posterior_sims %>% 
-  mutate(a = as.numeric(a), 
-         t = t+2000) %>% 
-  ggplot(aes(t,x)) + 
-  geom_line(aes(group = reps), col = "blue", alpha = 0.05,
-            data = true_forecast) + 
-  geom_line(aes(group = a), alpha = 0.05, col = "red") + 
-  geom_line(data = training_set) 
-```
-
-```{r}
 ```

### 2019-09-26 manuscript/manuscript.Rmd
wip

@@
 
-```{r}
-source("../R/ghost.R")
-```
-
-```{r}
-data <- vroom::vroom("../data/reps.csv.xz") %>% filter(t < 3000)
-## get tail cases for comparison
-ex <- data %>% group_by(reps) %>% summarise(ave = mean(x)) %>% arrange(ave)
-lows <- ex %>% dplyr::slice(1:10) %>% pull(reps)
-highs <- ex %>% dplyr::slice((n()-9):n()) %>% pull(reps)
-```
-
-```{r arima_forecast}
-library(stats)
-x <- data %>% filter(reps == lows[[6]], t < 2000) %>% pull(x) %>% as.ts() 
-fit <- arima(x)
-y <- predict(fit, n.ahead = 1000) %>% as.data.frame()
-y <- bind_rows(data.frame(pred = x, se = 0), y)
-arima_forecast <- data %>% filter(reps == lows[[6]]) %>% select(t,x) %>% bind_cols(y)
-```
-
-```{r arima_figure, message=FALSE, warning=FALSE}
-train <- arima_forecast %>% filter(t < 2000) 
-test <- arima_forecast %>% filter(t > 2000) 
-train %>%
-  ggplot(aes(t)) + 
-  geom_line(aes(y=x)) +
-  geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
-  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
-  geom_line(aes(y=x), data = test, lty = 2, col="blue") +
-  xlim(0,3000) + ylim(0, 1.6)+ ylab("x") + theme_minimal()
-
-```
-
 Transient dynamics have long been identified in the ecological systems @Hastings1994.   Ecological @Hastings2018

### 2019-09-26 manuscript/appendix/appendix.Rmd
model estimate over all parameters

@@
 library(furrr)
-set.seed(12345)
+
+library(greta)
+tensorflow::use_session_with_seed(12345)
+
 ```
@@
 ```{r}
+
+# a <- 2.3
 r <- p$r
@@
 K <- p$K
-# a <- 2.3
 H <- p$H
@@
 
+a <- uniform(0, 10)
 r <- uniform(0, 2 * p$r)
-Q <- uniform(0, 10)
+#Q <- uniform(0, 10)
 K <- uniform(0, 2*p$K)
-a <- uniform(0, 10)
 H <- uniform(0,  2*p$H)
@@
 distribution(x_t1) <- normal(mean, sigma * x_t)
-m <- model(a)
+
+m <- model(a, r, K, H, sigma)
 ```
@@
 ```{r }
+
 system.time({
@@
 true <- 
-  data.frame(a = 2.3) %>%
+  as_tibble(p) %>% select(-N, -x0, -Q) %>%
   gather(variable, value)
@@
 
-posterior_samples %>%
-  mutate(N=1e3, x0 = x0) %>%
-  purrr::transpose() %>% map_dfr()
+posterior_sims <- posterior_samples %>%
+  mutate(N=1e3, x0 = x0, Q = p$Q) %>% 
+         ## sigma = p$sigma, H = p$H, K = p$K, r = p$r) %>%
+  purrr::transpose() %>%
+  map_dfr(function(q) sim(may, q) ,.id = "reps")
+```
 
-p$N <- 1e3
+```{r}
+#q <- posterior_samples[1,] %>%  mutate(N=1e3, x0 = x0, Q = p$Q)
+#sim(may, q)
 
-p$x0 <- x[[1]]
-posterior_sims <- map_dfr(a, function(a){
-  p$a <- a
-  sim(may, p)
-},
-.id = "reps")
+#posterior_sims <- map_dfr(a, function(a){
+#  p$a <- a
+#  sim(may, p)
+#},
+#.id = "reps")
 
@@
 
+p$x0 <- x0
+p$N <- 1e3
 true_forecast <- 

### 2019-09-27 manuscript/manuscript.Rmd
WIP

@@
 
+Long transients may be quite common features of the dynamics of ecological systems.
+What implications does this have for our ability to predict and manage them?
+One of the most intuitive implications for the existence of long transients in the risk in mistaking transient behavior for stable state behavior and the consequences of managing a system under this false conclusion.
+The potential for this mistake can is straightforward to demonstrate even under idealized conditions such as access to accurate, high frequency data and knowledge of the correct model structure.
+Decision-theoretic approaches to ecological management provide a more nuianced approach for dealing with uncertainty by explicitly considering the set of all actions available to the manager and the likely consequences of those actions over time.
+In this paper, I explore the ability of both passive and active adaptive management approaches to cope with the challenge of long transient dynamics.
+My results underscore
+
 ## Long transients when the underlying model is not known

### 2019-10-02 manuscript/manuscript.Rmd
WIP

@@
 
-Transient dynamics have long been identified in the ecological systems @Hastings1994.   Ecological @Hastings2018
+Transient dynamics have long been identified in the ecological systems
 
-Long transients may be quite common features of the dynamics of ecological systems.
+Long transients may be quite common features of the dynamics of ecological systems [@Hastings1994; @Hastings2018].
 What implications does this have for our ability to predict and manage them?
@@
 In this paper, I explore the ability of both passive and active adaptive management approaches to cope with the challenge of long transient dynamics.
-My results underscore
+My results underscore the challenge transients create for accurate model inference, but also show a potential way forward for accounting for the resulting uncertainty in a decision-theory framework.
+Examining the impact different prior beliefs have on adaptive management, I show how a precautionary principle emerges from the optimization.
+This analysis also highlights the importance of stochasticity has for transient dynamics.
 

### 2019-10-03 manuscript/manuscript.Rmd
WIP

@@
 extrafont::loadfonts()
+pal <- solarized_pal()(3)
 ```
@@
 ```
-
-Transient dynamics have long been identified in the ecological systems
 
@@
 This analysis also highlights the importance of stochasticity has for transient dynamics.
+
+```{r results="hide"}
+library(MDPtoolbox)
+library(sarsop) # remotes::install_github("boettiger-lab/sarsop")
+library(tidyverse) # for plotting
+library(mdplearning)
+
+damage <- 0.05
+control <- 1
+reward_fn <- function(x,h) - damage * x ^ 2 - control * h
+discount <- 0.98
+
+states <- seq(0,2, length=200)
+actions <- states
+observations <- states
+sigma_g <- 0.05
+sigma_m <- 0.0
+
+r <- 0.8
+K <- 1.53
+q <- 2
+b <- .2
+eps <- states[2]/10
+Tmax <- 100
+
+may <- function(a){  
+  function(x, h){ # May
+    # x <- pmax(x - h, 0)   # harvest then recruit   
+    x + x * r * (1 - x / K)  - a * x ^ q / (x ^ q + b ^ q) + eps - h
+  }
+}
+
+possible_a <- seq(.25, .34, by = 0.005)
+true_a <- 0.27   ## reality has just a transient.  use  0.277 for stronger ghost
+believe_a <- 0.31 ## believe there's a strong attractor
+true_i <- which.min(abs(possible_a - true_a))
+
+```
+
+## True model: weak ghost attractor
+
+```{r true_a}
+f <- may(true_a)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + 
+  geom_line(col=pal[1], lwd=2) +
+  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
+  ylim(-0.06, 0.06)
+```
+
+---
+
+## Incorrect belief of bistable system
+
+```{r a29}
+f <- may(0.29)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + 
+  geom_line(col=pal[1], lwd=2) +
+  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
+  ylim(c(-0.04, 0.04))
+```
+
+---
+
+## Incorrect belief of stronger bistable system
+
+```{r a31}
+f <- may(believe_a)
+tibble(x = states[1:120],
+       f = f(x,0) - x) %>%
+  ggplot(aes(x, f)) + 
+  geom_line(col=pal[1], lwd=2) +
+  geom_hline(aes(yintercept = 0), lwd=2, lty=2) + 
+  ylim(-0.03, 0.03)
+```
+
+---
+
+```{r results="hide"}
+m_true <- fisheries_matrices(states, actions, observations, reward_fn, 
+                        may(true_a), sigma_g, sigma_m, noise = "lognormal")
+
+m <- fisheries_matrices(states, actions, observations, reward_fn, 
+                        may(believe_a), sigma_g, sigma_m, noise = "lognormal")
+soln <- mdp_value_iteration(m$transition, m$reward, discount)
+opt_soln <- mdp_value_iteration(m_true$transition, m_true$reward, discount)
+
+```
+
+## Simulation under *no intervention*
+
+```{r no_intervention}
+x0 <- which.min(abs(states - 0.1))
+no_policy <- numeric(length(states)) + 1
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = no_policy, x0 = x0, Tmax = 100)
+```
+
+```{r}
+df %>% mutate(state = states[state], action = 5*actions[action]) %>% 
+  select(state, action, time) %>% 
+  gather(series, state, -time) %>%
+  ggplot(aes(time, state, col=series)) +
+  geom_line(lwd=2) + 
+  ylim(0,1.3)
+```
+
+---
+
+## Mistaken ghost for alternate stable state
+
+```{r mistaken_ghost}
+x0 <- which.min(abs(states - 0.1))
+no_policy <- numeric(length(states)) + 1
+
+df <- mdp_planning(m_true$transition, m_true$reward, discount, model_prior = c(1), 
+                   policy = soln$policy, x0 = x0, Tmax = 100)
+
+df %>%
+  mutate(state = states[state], action = 5*actions[action]) %>% 
+  select(state, action, time) %>% 
+  gather(series, state, -time) %>%
+  ggplot(aes(time, state, col=series)) +
+  geom_line(lwd=2) + 
+  ylim(0,1.3)
+```
+
+## Prior beliefs
+
+```{r priors}
+prior <- dnorm(possible_a, 0.295, 0.005)
+prior <- prior / sum(prior)
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + 
+  geom_bar(stat="identity", fill=pal[1], alpha=0.6) +
+  geom_vline(aes(xintercept = true_a), lwd=3, lty=2, col=pal[1]) + 
+  ylim(0, 0.6)
+```
+
+---
+
+## Learning late
+
+```{r learning}
+sim <- readRDS("../notebook/sim-ghost-learning.Rds")
+states <- seq(0,2, length=140)
+actions <- states
+Tmax <- 100
+
+ sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% seq(1,Tmax, by = 10), time < 50) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) 
+```
+
+---
+
+## Outcome
+
+```{r}
+ sim$df %>% 
+   mutate(state = states[state], action = 5*actions[action]) %>% 
+  select(-value, -obs) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, state, color = series)) +
+  geom_line(lwd=2) +
+  ylim(0,1.3) +
+  ylab("state / action")
+```
+
+---
+
+## More agnostic prior beliefs
+
+```{r priors_wide}
+prior <- dnorm(possible_a, 0.3, 0.05)
+prior <- prior / sum(prior)
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity", fill=pal[1], alpha=0.5) +
+  geom_vline(aes(xintercept = true_a), lwd=3, lty=2, col=pal[2]) +
+  ylim(0,.3)
+```
+
+---
+
+```{r less-learning}
+sim2 <- readRDS("../notebook/sim-ghost-learning-2.Rds")
+states <- seq(0,2, length=200)
+actions <- states
+Tmax <- 100
+
+ sim2$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% seq(1,Tmax, by = 10), time < 50) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+   ylim(0,0.3)
+```
+
+---
+
+## Less learning, more caution
+
+```{r}
+ sim2$df %>% 
+   mutate(state = states[state], action = 5*actions[action]) %>% 
+  select(-value, -obs) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, state, color = series)) +
+  geom_line(lwd=2) + 
+  ylim(0,1.3) +
+  ylab("state / action")
+```
+
+```{r}
+```
 

### 2019-10-05 notebook/appendix.Rmd
a revised discrete-sims version includes both greta estimation and mdp_learning

---
output: github_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
library(tidyverse) 
library(MDPtoolbox)
library(mdplearning)
library(expm)
library(truncnorm)

```

# Outbreak model

- 'harvest' term corresponds to removal of pest, with associated cost
- also experience damage costs proportional to pest abundance

```{r}
damage <- 0.05
control <- 1
discount <- 0.98

n_s <- 200
states <- seq(0,120, length=n_s)
actions <- states

p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, a=2.3,
          N = 3e3, x0 = 20)
#eps <- states[2]/10

may <- function(a){  
  function(x, h = 0){ # May
    x + x * p$r * (1 - x / p$K)  - a * x ^ p$q / (x ^ p$q + p$b ^ p$q) - h
  }
}

reward_fn <- function(x,h) - damage * (x / max(states)) ^ 2 - control * h / max(actions)

```

Range of possible a that covers tipping in both directions:

```{r}

true_a <- 27.4      ### 21.5
believe_a <- 29 # 28.5   ### 22.5
possible_a <- seq(2*true_a - believe_a, 2*believe_a - true_a, by = 0.1)
true_i <- which.min(abs(possible_a - true_a))
```

```{r}

## True a: The Ghost Attractor
names(possible_a) <- possible_a
df <- map_dfr(possible_a, 
        function(a){
          f <- may(a)
          tibble(x = states,
                 f = f(x,0) - x) 
        },
        .id = "a") %>% 
  mutate(group = a, 
         a = as.numeric(a))

df %>%
  ggplot(aes(x, f)) + 
    geom_point(aes(col = a), alpha = 0.1) + 
    geom_line(aes(lty = group), data = df %>% filter(a == true_a | a == believe_a), 
              lwd = 1, show.legend = FALSE) +
    geom_hline(aes(yintercept = 0))

```

```{r}

transition_matrix <- function(states, actions, f, sigma){
    n_s <- length(states)
    n_a <- length(actions)
    transition <- array(0, dim = c(n_s, n_s, n_a))
    for(i in 1:n_a){
      for (k in 1:n_s) {
          nextpop <- f(states[k], actions[i])
          if (nextpop < 0) {
          #if(which.min(abs(states - nextpop)) > 1) {
             x  <- c(1, rep(0, n_s - 1))
          } else {
            x <- truncnorm::dtruncnorm(states, 0, max(states), nextpop, sigma * nextpop)
            if(sum(x) <= 0){ 
              x  <- c(1, rep(0, n_s - 1))
            } else {
              x <- x / sum(x)
            }
          }
          transition[k, , i] <- x
      }
  }
  if(any(is.na(transition))) stop("error creating transition matrix")
    transition
}
```

```{r}
m_true <- transition_matrix(states, actions, may(true_a), p$sigma)
m_belief <- transition_matrix(states, actions, may(believe_a), p$sigma)
```

"simulate" without any actions

```{r}
x0 <- which.min(abs(states - p$x0))
Tmax <- 200
```

```{r}
X <- numeric(length(states))
X[x0] <- 1

prob_dynamics <- function(M, X, Tmax){
  probability <- t(M) %^% Tmax %*% X 
  data.frame(state = states, probability)
}

prob_dynamics(m_true[,,1], X, Tmax) %>% 
  ggplot(aes(state,probability))  + geom_point() # + geom_bar(stat="identity")
```

```{r}

## `actions` is an argument of indices, not necessarily the action values themselves 
sim <- function (transition,  x0, Tmax, action = rep(1, Tmax)){
    n_states <- dim(transition)[2]
    state <- numeric(Tmax + 1)
    state[1] <- x0
    time <- 1:Tmax
    for (t in time) {
        state[t + 1] <- base::sample(1:n_states, 
                               1, 
                               prob = transition[state[t], , action[t]])
    }
    data.frame(time = 1:Tmax, state = state[time])
}
```

```{r}
set.seed(12346)
no_switches <- sim(m_true, x0, Tmax)  %>% 
  mutate(state = states[state])

no_switches %>% 
  ggplot(aes(time, state)) + geom_point() + geom_path() 
```

```{r}
set.seed(12345)
switches <- sim(m_true, x0, Tmax) %>% mutate(state = states[state])

switches  %>% 
  ggplot(aes(time, state)) + geom_point() + geom_path() 
```

```{r}
set.seed(1234)
stable <- sim(m_belief, x0, Tmax)  %>% 
  mutate(state = states[state])

stable %>% 
  ggplot(aes(time, state)) + geom_point() + geom_path() 
```

```{r}
## Collect the data sets.  Use names.
examples <- bind_rows(switches = switches, 
                      no_switches = no_switches, 
                      stable = stable, 
                      .id="series") 
```

-----

# Greta

```{r}
library(greta)
```

```{r}
estimate_posterior <- function(df){
  X <- df$state
  # Reshape time-series data into ordered pairs
  n <- length(X)
  x_t1 <- X[-1]
  x_t <- X[-n]

  # Prior
  a <- uniform(min(possible_a), max(possible_a))

  # Model definition
  mean <- x_t + p$r * x_t * (1 - x_t / p$K) - a * x_t ^ p$q / (x_t ^ p$q + p$b ^ p$q)
  distribution(x_t1) <- normal(mean, p$sigma * x_t)
  m <- model(a)

  # MCMC
  system.time({
    draws <- mcmc(m, n_samples = 1000, warmup = 3000, chains = 4, verbose = FALSE)
  })

  ## tidy
  samples <-  
    map_dfr(draws, function(x) data.frame(x, t = 1:dim(x)[1]), .id = "chain") %>% 
    gather(variable, value, -t, -chain)
}
```

```{r}
samples <- examples %>% group_by(series) %>% group_modify(~estimate_posterior(.x))
```

```{r, warning=FALSE, message=FALSE}
true <- tribble(~ value, ~ a,
                believe_a, "attractor",
                believe_a,  "attractor",
                true_a,     "ghost")

samples %>% 
  ggplot() + 
  geom_histogram(aes(value, fill=series), stat="density", alpha = 0.8) + 
  geom_vline(data = true, aes(xintercept = value, lty=a), col="black", lwd = 1) + 
  facet_wrap(~variable, scales = "free")
```

----

# Learning

```{r}
reward <- array(dim=c(length(states), length(actions)))
for(i in 1:length(states)){
  for(j in 1:length(actions)){
    reward[i,j] <- reward_fn(states[i], actions[j])
  }
}
```

```{r}
no_policy <- numeric(length(states)) + 1
x0 <- which.min(abs(states - 25))

set.seed(12345678)
df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
                   policy = no_policy, x0 = x0, Tmax = Tmax)

df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) + geom_point()+geom_path() + 
  geom_line(aes(time, action), col="blue")
```

```{r}
soln <- mdp_value_iteration(m_belief, reward, discount)
opt_soln <- mdp_value_iteration(m_true, reward, discount)
#soln <- mdplearning::mdp_compute_policy(list(m_belief), reward, discount)
#opt_soln <- mdplearning::mdp_compute_policy(list(m_true), reward, discount)
```

Policy based on the belief (i.e. that system is bi-stable)

```{r}
tibble(state = states,
       belief = actions[soln$policy],
       ideal =  actions[opt_soln$policy]) %>%
  pivot_longer(c(belief, ideal), 
               names_to="model",
               values_to = "action") %>%
  ggplot(aes(state,action, color = model)) + geom_point() 
```

Result experienced by incorrect belief:  initial in-action followed by need for continued maintenance to prevent high-level outbreak:

```{r}
set.seed(12345)
df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
                   policy = soln$policy, x0 = x0, Tmax = 200)

df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) + geom_point()+geom_path() + 
  geom_line(aes(time, action), col="blue")
```

Expected result based on the belief: stable low level is acceptable, so no action is required:

```{r}

df <- mdp_planning(m_belief, m$reward, discount, model_prior = c(1), 
                   policy = soln$policy, x0 = x0, Tmax = 100)

df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) + geom_point()+geom_path() + 
  geom_line(aes(time, action), col="blue")
```

Optimal strategy knowing this is just a transient (will depend on discount rate):

```{r}

df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
                   policy = opt_soln$policy, x0 = x0, Tmax = 200)

df %>% mutate(state = states[state], action = actions[action]) %>% 
  ggplot(aes(time, state)) + geom_point()+geom_path() + 
  geom_line(aes(time, action), col="blue")
```

-----

```{r}

models <- map(possible_a, function(a){
  f <- may(a)
 transition_matrix(states, actions, f, p$sigma)
})

```

```{r}
transition <- models
```

## a near ghost

```{r}
wd <- sd(possible_a)
prior <- dnorm(possible_a, believe_a, wd/2)
prior <- prior / sum(prior)
```

```{r}
data.frame(a = possible_a, probability = prior) %>%
  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
```

```{r}
sim <- mdp_learning(transition, reward, discount, 
                    x0 = x0, 
                    Tmax = Tmax, 
                    true_transition = transition[[true_i]],
                    model_prior = prior,
                    type = "value iteration", 
                    epsilon = 1e-2)
```

```{r}
 sim$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

```{r}
 sim$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% seq(1,Tmax, by = 5)) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = as.numeric(param)) %>% 
  ggplot(aes(param, probability, group = time, alpha = time)) + 
  geom_line()
```

```{r}
```

### 2019-10-08 notebook/appendix.Rmd
WIP

@@
 ```{r}
+#damage <- 0.3
+#control <- 100
+#discount <- 0.99
+
 damage <- 0.05
@@
 
-n_s <- 200
-states <- seq(0,120, length=n_s)
+n_s <- 110
+states <- seq(0,110, length=n_s)
 actions <- states
 
-p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, a=2.3,
-          N = 3e3, x0 = 20)
-#eps <- states[2]/10
+p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, a=2.3, x0 = 20)
+eps <- states[2] / 10
 
@@
   function(x, h = 0){ # May
-    x + x * p$r * (1 - x / p$K)  - a * x ^ p$q / (x ^ p$q + p$b ^ p$q) - h
+    pmax(
+      x + x * p$r * (1 - x / p$K)  - a * x ^ p$q / (x ^ p$q + p$b ^ p$q) + eps - h,  
+      0)
   }
@@
 
-reward_fn <- function(x,h) - damage * (x / max(states)) ^ 2 - control * h / max(actions)
+reward_fn <- function(x,h) - damage * (x / 100) - control * h / 100
+
+#reward_fn <- function(x,h) - damage * (x / max(states)) - control * h / max(actions) * x / max(states)
 
@@
 
-Range of possible a that covers tipping in both directions:
+Range of possible a that covers tipping in both directions.  Belief is in a stable system while the reality is a transient (smaller `a`).
 
@@
 believe_a <- 29 # 28.5   ### 22.5
-possible_a <- seq(2*true_a - believe_a, 2*believe_a - true_a, by = 0.1)
+
+possible_a <- seq(2*true_a - believe_a, 2*believe_a - true_a, by = 0.2)
 true_i <- which.min(abs(possible_a - true_a))
@@
 X <- numeric(length(states))
-X[x0] <- 1
+X[2] <- 1
 
@@
 
+----
+
+# Learning
+
+```{r}
+x0 <- which.min(abs(states - 23))
+```
+
+```{r}
+no_policy <- numeric(length(states)) + 1
+
+#set.seed(12345678)
+set.seed(12345)
+
+df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
+                   policy = no_policy, x0 = x0, Tmax = Tmax)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+```{r}
+reward <- array(dim=c(length(states), length(actions)))
+for(i in 1:length(states)){
+  for(j in 1:length(actions)){
+    reward[i,j] <- reward_fn(states[i], actions[j])
+  }
+}
+```
+
+```{r}
+belief_p <- mdp_value_iteration(m_belief, reward, discount, epsilon = 1e-4)
+true_p <- mdp_value_iteration(m_true, reward, discount, epsilon = 1e-5)
+#soln <- mdplearning::mdp_compute_policy(list(m_belief), reward, discount)
+#opt_soln <- mdplearning::mdp_compute_policy(list(m_true), reward, discount)
+```
+
+Policy based on the belief (i.e. that system is bi-stable)
+
+```{r}
+tibble(state = states,
+       belief = actions[belief_p$policy],
+       ideal =  actions[true_p$policy]) %>%
+  pivot_longer(c(belief, ideal), 
+               names_to="model",
+               values_to = "action") %>%
+  ggplot(aes(state,action, color = model)) + geom_point() 
+```
+
+Result experienced by incorrect belief:  initial in-action followed by need for continued maintenance to prevent high-level outbreak:
+
+```{r}
+set.seed(12345)
+df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
+                   policy = belief_p$policy, x0 = x0, Tmax = 1000)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Expected result based on the belief: stable low level is acceptable, so no action is required:
+
+```{r}
+set.seed(12345)
+df <- mdp_planning(m_belief, reward, discount, model_prior = c(1), 
+                   policy = belief_p$policy, x0 = 26, Tmax = 1000)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Optimal strategy knowing this is just a transient (will depend on discount rate):
+
+```{r}
+
+df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
+                   policy = true_p$policy, x0 = x0, Tmax = 1000)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+----
+
+## Active Learning
+
+```{r}
+
+models <- map(possible_a, function(a){
+  f <- may(a)
+ transition_matrix(states, actions, f, p$sigma)
+})
+
+```
+
+```{r}
+transition <- models
+```
+
+## a near ghost
+
+```{r}
+wd <- sd(possible_a)
+prior <- dnorm(possible_a, believe_a, wd/2)
+prior <- prior / sum(prior)
+```
+
+```{r}
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
+  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
+  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
+```
+
+```{r}
+if(!file.exists("learning_sim.tsv.gz")){ # Cache the slow step
+
+  learning_sim <- mdp_learning(transition, reward, discount, 
+                      x0 = x0, 
+                      Tmax = Tmax, 
+                      true_transition = transition[[true_i]],
+                      model_prior = prior,
+                      type = "value iteration", 
+                      epsilon = 1e-2)
+  readr::write_tsv(learning_sim, "learning_sim.tsv.gz")
+
+} else {
+  learning_sim <- readr::read_tsv("learning_sim.tsv.gz")
+}
+```
+
+```{r}
+ learning_sim$df %>% 
+  select(-value) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, states[state], color = series)) + geom_line()
+```
+
+```{r}
+ learning_sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% seq(1,Tmax, by = 5)) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(param, probability, group = time, alpha = time)) + 
+  geom_line()
+```
+
+```{r}
+```
+
 -----
@@
 
-```{r}
-samples <- examples %>% group_by(series) %>% group_modify(~estimate_posterior(.x))
+```{r mcmc, cache=TRUE}
+if(!file.exists("samples.tsv.gz")){ # Manually cache the slow step
+
+samples <- examples %>% 
+  group_by(series) %>% 
+  group_modify(~estimate_posterior(.x))
+
+readr::write_tsv(samples, "samples.tsv.gz")
+
+} else {
+  samples <- readr::read_tsv("samples.tsv.gz")
+}
 ```
@@
 
-----
-
-# Learning
-
-```{r}
-reward <- array(dim=c(length(states), length(actions)))
-for(i in 1:length(states)){
-  for(j in 1:length(actions)){
-    reward[i,j] <- reward_fn(states[i], actions[j])
-  }
-}
-```
-
-```{r}
-no_policy <- numeric(length(states)) + 1
-x0 <- which.min(abs(states - 25))
-
-set.seed(12345678)
-df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = no_policy, x0 = x0, Tmax = Tmax)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-```{r}
-soln <- mdp_value_iteration(m_belief, reward, discount)
-opt_soln <- mdp_value_iteration(m_true, reward, discount)
-#soln <- mdplearning::mdp_compute_policy(list(m_belief), reward, discount)
-#opt_soln <- mdplearning::mdp_compute_policy(list(m_true), reward, discount)
-```
-
-Policy based on the belief (i.e. that system is bi-stable)
-
-```{r}
-tibble(state = states,
-       belief = actions[soln$policy],
-       ideal =  actions[opt_soln$policy]) %>%
-  pivot_longer(c(belief, ideal), 
-               names_to="model",
-               values_to = "action") %>%
-  ggplot(aes(state,action, color = model)) + geom_point() 
-```
-
-Result experienced by incorrect belief:  initial in-action followed by need for continued maintenance to prevent high-level outbreak:
-
-```{r}
-set.seed(12345)
-df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = soln$policy, x0 = x0, Tmax = 200)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-Expected result based on the belief: stable low level is acceptable, so no action is required:
-
-```{r}
-
-df <- mdp_planning(m_belief, m$reward, discount, model_prior = c(1), 
-                   policy = soln$policy, x0 = x0, Tmax = 100)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-Optimal strategy knowing this is just a transient (will depend on discount rate):
-
-```{r}
-
-df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = opt_soln$policy, x0 = x0, Tmax = 200)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
------
-
-```{r}
-
-models <- map(possible_a, function(a){
-  f <- may(a)
- transition_matrix(states, actions, f, p$sigma)
-})
-
-```
-
-```{r}
-transition <- models
-```
-
-## a near ghost
-
-```{r}
-wd <- sd(possible_a)
-prior <- dnorm(possible_a, believe_a, wd/2)
-prior <- prior / sum(prior)
-```
-
-```{r}
-data.frame(a = possible_a, probability = prior) %>%
-  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
-  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
-  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
-```
-
-```{r}
-sim <- mdp_learning(transition, reward, discount, 
-                    x0 = x0, 
-                    Tmax = Tmax, 
-                    true_transition = transition[[true_i]],
-                    model_prior = prior,
-                    type = "value iteration", 
-                    epsilon = 1e-2)
-```
-
-```{r}
- sim$df %>% 
-  select(-value) %>% 
-  gather(series, state, -time) %>% 
-  ggplot(aes(time, states[state], color = series)) + geom_line()
-```
-
-```{r}
- sim$posterior %>% 
-  data.frame(time = 1:Tmax) %>%
-  filter(time %in% seq(1,Tmax, by = 5)) %>%
-  gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = as.numeric(param)) %>% 
-  ggplot(aes(param, probability, group = time, alpha = time)) + 
-  geom_line()
-```
-
 ```{r}

### 2019-10-09 notebook/appendix.Rmd
apply h first much cleaner

@@
 control <- 1
-discount <- 0.98
+discount <- 0.999
 
-n_s <- 110
-states <- seq(0,110, length=n_s)
-actions <- states
+n_s <- 121
+states <- seq(0, 120, length = n_s)
+actions <- seq(0, 120, length = n_s)
 
 p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, a=2.3, x0 = 20)
-eps <- states[2] / 10
 
@@
   function(x, h = 0){ # May
+    y <- x - h
     pmax(
-      x + x * p$r * (1 - x / p$K)  - a * x ^ p$q / (x ^ p$q + p$b ^ p$q) + eps - h,  
+      ## controlling h is controlling the bifurcation point directly...
+      y + y * p$r * (1 - y / p$K)  - a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
       0)
@@
 
-reward_fn <- function(x,h) - damage * (x / 100) - control * h / 100
+reward_fn <- function(x,h) - damage * (x / 100) ^ 2 - control * (h / 100)
 
@@
 
-true_a <- 27.4      ### 21.5
+true_a <- 27.2      ### 21.5
 believe_a <- 29 # 28.5   ### 22.5
@@
           nextpop <- f(states[k], actions[i])
-          if (nextpop < 0) {
+          if (nextpop <= 0) {
           #if(which.min(abs(states - nextpop)) > 1) {
@@
 
-"simulate" without any actions
-
-```{r}
-x0 <- which.min(abs(states - p$x0))
-Tmax <- 200
-```
-
 ```{r}
@@
 
-prob_dynamics(m_true[,,1], X, Tmax) %>% 
-  ggplot(aes(state,probability))  + geom_point() # + geom_bar(stat="identity")
+Tmax <- 500
+true_distrib <- prob_dynamics(m_true[,,2], X, Tmax)
+believe_distrib <- prob_dynamics(m_belief[,,2], X, Tmax)
+bind_rows(true = true_distrib, believe = believe_distrib, .id = "model") %>% 
+  ggplot(aes(state,probability, col=model))  + geom_line(alpha=0.5) # + geom_bar(stat="identity")
+```
+
+----
+
+# Learning
+
+```{r}
+x0 <- which.min(abs(states - 25))
+```
+
+```{r}
+reward <- array(dim=c(length(states), length(actions)))
+for(i in 1:length(states)){
+  for(j in 1:length(actions)){
+    reward[i,j] <- reward_fn(states[i], actions[j])
+  }
+}
+```
+
+```{r}
+belief_p <- mdp_value_iteration(m_belief, reward, discount, epsilon = 1e-4)
+true_p <- mdp_value_iteration(m_true, reward, discount, epsilon = 1e-5)
+#soln <- mdplearning::mdp_compute_policy(list(m_belief), reward, discount)
+#opt_soln <- mdplearning::mdp_compute_policy(list(m_true), reward, discount)
+```
+
+Policy based on the belief (i.e. that system is bi-stable)
+
+```{r}
+tibble(state = states,
+       belief = actions[belief_p$policy],
+       ideal =  actions[true_p$policy]) %>%
+  pivot_longer(c(belief, ideal), 
+               names_to="model",
+               values_to = "action") %>%
+  ggplot(aes(state,action, color = model)) + geom_point(alpha=0.5) 
+```
+
+Result experienced by incorrect belief:  initial in-action followed by need for continued maintenance to prevent high-level outbreak:
+
+```{r}
+set.seed(12345)
+df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
+                   policy = belief_p$policy, x0 = x0, Tmax = 100)
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Expected result based on the belief: stable low level is acceptable, so no action is required:
+
+```{r}
+set.seed(12345)
+df <- mdp_planning(m_belief, reward, discount, model_prior = c(1), 
+                   policy = belief_p$policy, x0 = x0, Tmax = 1000)
+
+df %>% mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + geom_point()+geom_path() + 
+  geom_line(aes(time, action), col="blue")
+```
+
+Optimal strategy knowing this is just a transient (will depend on discount rate):
+
+```{r}
+df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
+                   policy = true_p$policy, x0 = x0, Tmax = 1000)
+
+df %>% 
+  mutate(state = states[state], action = actions[action]) %>% 
+  ggplot(aes(time, state)) + 
+  geom_point(alpha=0.5)  + 
+  geom_line(aes(time, action), col="blue")
+```
+
+----
+
+## Active Learning
+
+```{r}
+
+models <- map(possible_a, function(a){
+  f <- may(a)
+ transition_matrix(states, actions, f, p$sigma)
+})
+
+```
+
+```{r}
+transition <- models
+```
+
+## a near ghost
+
+```{r}
+wd <- sd(possible_a)
+prior <- dnorm(possible_a, believe_a, wd/2)
+prior <- prior / sum(prior)
+```
+
+```{r}
+data.frame(a = possible_a, probability = prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
+  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
+  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
+```
+
+```{r}
+mdp_learning_ <- memoise::memoise(mdp_learning)
+  learning_sim <- mdp_learning_(transition, reward, discount, 
+                      x0 = x0, 
+                      Tmax = Tmax, 
+                      true_transition = transition[[true_i]],
+                      model_prior = prior,
+                      type = "value iteration", 
+                      epsilon = 1e-2)
+
+```
+
+```{r}
+ learning_sim$df %>% 
+  select(-value) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, states[state], color = series)) + geom_line()
+```
+
+```{r}
+ learning_sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% seq(1,Tmax, by = 5)) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(param, probability, group = time, alpha = time)) + 
+  geom_line()
+```
+
+```{r}
+```
+
+-----
+
+# Simulations for Gretas
+
+"simulate" without any actions
+
+```{r}
+x0 <- which.min(abs(states - p$x0))
+Tmax <- 200
 ```
@@
 ```
-
-----
-
-# Learning
-
-```{r}
-x0 <- which.min(abs(states - 23))
-```
-
-```{r}
-no_policy <- numeric(length(states)) + 1
-
-#set.seed(12345678)
-set.seed(12345)
-
-df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = no_policy, x0 = x0, Tmax = Tmax)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-```{r}
-reward <- array(dim=c(length(states), length(actions)))
-for(i in 1:length(states)){
-  for(j in 1:length(actions)){
-    reward[i,j] <- reward_fn(states[i], actions[j])
-  }
-}
-```
-
-```{r}
-belief_p <- mdp_value_iteration(m_belief, reward, discount, epsilon = 1e-4)
-true_p <- mdp_value_iteration(m_true, reward, discount, epsilon = 1e-5)
-#soln <- mdplearning::mdp_compute_policy(list(m_belief), reward, discount)
-#opt_soln <- mdplearning::mdp_compute_policy(list(m_true), reward, discount)
-```
-
-Policy based on the belief (i.e. that system is bi-stable)
-
-```{r}
-tibble(state = states,
-       belief = actions[belief_p$policy],
-       ideal =  actions[true_p$policy]) %>%
-  pivot_longer(c(belief, ideal), 
-               names_to="model",
-               values_to = "action") %>%
-  ggplot(aes(state,action, color = model)) + geom_point() 
-```
-
-Result experienced by incorrect belief:  initial in-action followed by need for continued maintenance to prevent high-level outbreak:
-
-```{r}
-set.seed(12345)
-df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = belief_p$policy, x0 = x0, Tmax = 1000)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-Expected result based on the belief: stable low level is acceptable, so no action is required:
-
-```{r}
-set.seed(12345)
-df <- mdp_planning(m_belief, reward, discount, model_prior = c(1), 
-                   policy = belief_p$policy, x0 = 26, Tmax = 1000)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-Optimal strategy knowing this is just a transient (will depend on discount rate):
-
-```{r}
-
-df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = true_p$policy, x0 = x0, Tmax = 1000)
-
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
-```
-
-----
-
-## Active Learning
-
-```{r}
-
-models <- map(possible_a, function(a){
-  f <- may(a)
- transition_matrix(states, actions, f, p$sigma)
-})
-
-```
-
-```{r}
-transition <- models
-```
-
-## a near ghost
-
-```{r}
-wd <- sd(possible_a)
-prior <- dnorm(possible_a, believe_a, wd/2)
-prior <- prior / sum(prior)
-```
-
-```{r}
-data.frame(a = possible_a, probability = prior) %>%
-  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
-  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
-  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
-```
-
-```{r}
-if(!file.exists("learning_sim.tsv.gz")){ # Cache the slow step
-
-  learning_sim <- mdp_learning(transition, reward, discount, 
-                      x0 = x0, 
-                      Tmax = Tmax, 
-                      true_transition = transition[[true_i]],
-                      model_prior = prior,
-                      type = "value iteration", 
-                      epsilon = 1e-2)
-  readr::write_tsv(learning_sim, "learning_sim.tsv.gz")
-
-} else {
-  learning_sim <- readr::read_tsv("learning_sim.tsv.gz")
-}
-```
-
-```{r}
- learning_sim$df %>% 
-  select(-value) %>% 
-  gather(series, state, -time) %>% 
-  ggplot(aes(time, states[state], color = series)) + geom_line()
-```
-
-```{r}
- learning_sim$posterior %>% 
-  data.frame(time = 1:Tmax) %>%
-  filter(time %in% seq(1,Tmax, by = 5)) %>%
-  gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = as.numeric(param)) %>% 
-  ggplot(aes(param, probability, group = time, alpha = time)) + 
-  geom_line()
-```
-
-```{r}
-```
-
------
 

### 2019-10-09 notebook/transition-matrix-pile-boundary.R
pile-boundary version

@@ -1,0 +1,34 @@
+
+transition_matrix <- function(states, actions, f, sigma){
+  n_s <- length(states)
+  n_a <- length(actions)
+  transition <- array(0, dim = c(n_s, n_s, n_a))
+  for(i in 1:n_a){
+    for (k in 1:n_s) {
+      nextpop <- f(states[k], actions[i])
+      if (nextpop <= 0) {
+        #if(which.min(abs(states - nextpop)) > 1) {
+        x  <- c(1, rep(0, n_s - 1))
+      } else {
+        n_s <- length(states)
+        mu <- nextpop
+        sig <- sigma * nextpop
+        x <- dnorm(states, mu, sig)
+        if(sum(x) <= 0){
+          x  <- c(1, rep(0, n_s - 1))
+        } else {
+          ## Pile negative density onto boundary
+          negs <- pnorm(states[1], mu, sig)
+          x[1] <- x[1] + negs
+          ## Pile density beyond last state onto other boundary
+          N <- pnorm(states[n_s], mu, sig)
+          x[n_s] <- x[n_s] + N
+          x <- x/sum(x)
+        }
+      }
+      transition[k, , i] <- x
+    }
+  }
+  if(any(is.nan(transition))) stop("error creating transition matrix")
+  transition
+}

### 2019-10-25 notebook/gp-manual-vs-greta.Rmd
WIP

---
layout: post
title: Basic regression in Gaussian processes 
category: ecology
tag: nonparametric-bayes
code: true
output: github_document
---

```{r}
library(tidyverse)
library(greta)
library(greta.gp)

require(MASS)
require(reshape2)
require(ggplot2)
set.seed(12345)
```

```{r}
x_predict <- seq(-5,5,len=50)
l <- 1
```

We will use the squared exponential (also called radial basis or Gaussian, though it is not this that gives Gaussian process it's name; any covariance function would do) as the covariance function,

$$\operatorname{cov}(X_i, X_j) = \exp\left(-\frac{(X_i - X_j)^2}{2 \ell ^ 2}\right)$$

```{r}
SE <- function(Xi,Xj, l) exp(-0.5 * (Xi - Xj) ^ 2 / l ^ 2)
cov <- function(X, Y) outer(X, Y, SE, l)
```

```{r}
COV <- cov(x_predict, x_predict)
```

Generate (draw) a number of functions from the process

```{r}
values <- mvrnorm(3, rep(0, length=length(x_predict)), COV)
```

Reshape the data into long (tidy) form, listing x value, y value, and sample number

```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")

fig2a <- ggplot(dat,aes(x=x,y=value)) +
  geom_rect(xmin=-Inf, xmax=Inf, ymin=-2, ymax=2, fill="grey80") +
  geom_line(aes(group=variable))  +
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x") +
  theme_bw()
fig2a
```

### Posterior distribution given the data

Imagine we have some data,

```{r}
obs <- data.frame(x = c(-4, -3, -1,  0,  2),
                  y = c(-2,  0,  1,  2, -1))
```

In general we aren't interested in drawing from the prior, but want to include information from the data as well.  We want the joint distribution of the observed values and the prior is:

$$\begin{pmatrix} y_{\textrm{obs}} \\ y_{\textrm{pred}} \end{pmatrix} \sim \mathcal{N}\left( \mathbf{0}, \begin{bmatrix} cov(X_o,X_o) & cov(X_o, X_p) \\ cov(X_p,X_o) & cov(X_p, X_p) \end{bmatrix} \right)$$

## No observation noise

Assuming the data are known without error and conditioning on the data, and given $x \sim \mathcal{N}(0, cov(X_o, X_o))$, then the conditional probability of observing our data is easily solved by exploiting the nice properties of Gaussians,

$$x|y \sim \mathcal{N}\left(E,C\right)$$
$$E = cov(X_p, X_o) cov(X_o,X_o)^{-1} y$$
$$C= cov(X_p, X_p) - cov(X_p, X_o) cov(X_o,X_o)^{-1} cov(X_o, X_p) $$

(We use `solve(M)` which with no second argument will simply invert the matrix `M`, but should use the Cholsky decomposition instead for better numerical stability)

```{r}
cov_xx_inv <- solve(cov(obs$x, obs$x))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
```

Now lets take 3 random samples from the posterior distribution,

```{r}
values <- mvrnorm(3, Ef, Cf)
```

and plot our solution (mean, 2 standard deviations, and the ranom samples.)

```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")

gp <- data.frame(x = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )

ggplot(dat,aes(x=x,y=value)) + 
  geom_ribbon(data = gp, 
              aes(x, 
                  y = Ef, 
                  ymin = Ef - sigma, 
                  ymax = Ef + sigma
              ),
              fill="grey80") +
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(dat = gp, aes(x=x,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
```

![plot of draws from the posterior, with no proces noise](http://farm9.staticflickr.com/8237/8591045344_918138a53a_o.png)

Additive noise
--------------

In general the model may have process error, and rather than observe the deterministic map $f(x)$ we only observe $y = f(x) + \varepsilon$.  Let us assume for the moment that $\varepsilon$ are independent, normally distributed random variables with variance $\sigma_n^2$.

```{r}
sigma.n <- 0.25
cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
```

Now lets take 3 random samples from the posterior distribution,

```{r}
values <- mvrnorm(3, Ef, Cf)
```

and plot

```{r}
dat <- data.frame(x=x_predict, t(values))
dat <- melt(dat, id="x")

gp <- data.frame(x_predict = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )

fig2c <- ggplot(dat,aes(x=x,y=value)) +
  geom_ribbon(data=gp, 
              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
              fill="grey80") + # Var
  geom_line(aes(color=variable)) + #REPLICATES
  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
  xlab("input, x")
fig2c
```

Note that unlike the previous case, the posterior no longer collapses completely around the neighborhood of the test points.

### 2019-10-25 notebook/greta-gp.Rmd
WIP

---
output: github_document

---

```{r setup}
library(tidyverse)
library(greta)
library(greta.gp)
#tensorflow::use_session_with_seed(1234)

```

```{r}
arima_forecast <- readr::read_csv("../data/arima_forecast.csv.gz")
train <- arima_forecast  %>% filter(t %in% seq(0, 2000, by = 5))

wide <- select(train, x) %>% as.matrix() 
n <- dim(wide)[1]
x_t1 <- wide[-1,]
x_t <- wide[-n,]

y <- c(0,x_t1 - x_t)
x <- c(0, x_t)
x_plot <- seq(0, 2, length.out = 200)
```

```{r}

```

```{r}
# simulate toy data
#x <- runif(20, 0, 10)
#y <- sin(x) + rnorm(20, 0, 0.5)
#x_plot <- seq(-1, 11, length.out = 200)
```

```{r}

# hyperparameters
rbf_var <- lognormal(0, 1)
rbf_len <- lognormal(0, .01)
obs_sd <- lognormal(0, .01)

# kernel & GP
kernel <- rbf(rbf_len, rbf_var) + bias(1)
f <- gp(x, kernel)

# likelihood
distribution(y) <- normal(f, obs_sd)

# prediction
f_plot <- project(f, x_plot)
```

```{r}
# fit the model by Hamiltonian Monte Carlo
m <- model(f_plot)
draws <- mcmc(m, one_by_one = TRUE)
```

```{r}
# plot 200 posterior samples
#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 10), ylim = c(-2.5, 2.5))

plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
for (i in 1:200) {
  lines(draws[[1]][i, ] ~ x_plot,
        lwd = 2,
        col = rgb(0.7, 0.1, 0.4, 0.1))  
}
```

```{r}
```

### 2019-10-25 manuscript/debug.R

@@ -1,0 +1,46 @@
+type <- "policy iteration"
+n_models <- length(transition)
+n_states <- dim(transition[[1]])[1]
+n_actions <- dim(transition[[1]])[3]
+next_value <- numeric(n_states)
+next_policy <- numeric(n_states)
+V_model <- array(dim = c(n_states, n_models))
+converged <- FALSE
+t <- 1
+if (is.null(model_prior)) {
+  model_prior <- rep(1, n_models)/n_models
+}
+if (type == "finite time")
+  max_iter <- Tmax
+while (t < max_iter && converged == FALSE) {
+  Q <- array(0, dim = c(n_states, n_actions))
+  for (i in 1:n_actions) {
+    for (j in 1:n_models) {
+      V_model[, j] <- transition[[j]][, , i] %*% next_value
+    }
+    Q[, i] <- reward[, i] + discount * V_model %*% model_prior
+    if(any(is.nan(Q))){
+      stop(paste(V_model[, j]))
+      #stop(sprintf("t=%i, i=%i,j=%i", t, i, j))
+      }
+    if(any(is.na(Q))) stop("NAs found")
+
+  }
+  value <- apply(Q, 1, max)
+  policy <- apply(Q, 1, which.max)
+  if (type == "value iteration") {
+    if (sum(abs(value - next_value)) < epsilon) {
+      converged <- TRUE
+    }
+  }
+  else if (type == "policy iteration") {
+    if (sum(abs(policy - next_policy)) < epsilon) {
+      converged <- TRUE
+    }
+  }
+  next_value <- value
+  next_policy <- policy
+  t <- t + 1
+  if (t == max_iter)
+    message("Note: max number of iterations reached")
+}

### 2019-10-25 notebook/nvblas.R

@@ -1,0 +1,14 @@
+N <- 2^14
+system.time(
+  replicate(10,{
+  matrix(rnorm(N*N), nrow=N, ncol=N) %*%
+  matrix(rnorm(N*N), nrow=N, ncol=N)})
+  )
+
+  rx <- callr::r(
+  function(){
+    N <- 2^14
+    M <- matrix(rnorm(N*N), nrow=N, ncol=N)
+    M %*% M
+  },
+  env=c(LD_PRELOAD="libnvblas.so"))

### 2019-10-26 notebook/gp-manual-vs-greta.Rmd
greta gp exploration

@@
 code: true
-output: github_document
+output: html_document
 ---
 
-```{r}
-library(tidyverse)
-library(greta)
-library(greta.gp)
+First I take a look at doing some simple tasks involving Gaussian processes in base R and sample from them with basic multi-variate normal random draw, `mvrnorm`.
+Then, I speculate (unsuccessfully) on how to do this in greta.
+
+```{r setup, message = FALSE, warning = FALSE}
 
@@
 require(ggplot2)
+
+knitr::opts_chunk$set(message=FALSE, warning=FALSE)
+
 set.seed(12345)
+
 ```
@@
 
-Generate (draw) a number of functions from the process
+Generate (draw) a number of functions from the process (note that this is not conditioned on any data, and we are leaving the hyperparameter `l` fixed)
 
 ```{r}
-values <- mvrnorm(3, rep(0, length=length(x_predict)), COV)
+values <- mvrnorm(200, rep(0, length=length(x_predict)), COV)
 ```
@@
 ```{r}
-dat <- data.frame(x=x_predict, t(values))
-dat <- melt(dat, id="x")
+dat <- data.frame(x=x_predict, t(values)) %>%
+  tidyr::pivot_longer(-x, names_to = "rep", values_to = "value") %>% 
+  mutate(rep = as.numeric(as.factor(rep)))
 
-fig2a <- ggplot(dat,aes(x=x,y=value)) +
-  geom_rect(xmin=-Inf, xmax=Inf, ymin=-2, ymax=2, fill="grey80") +
-  geom_line(aes(group=variable))  +
-  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
-  xlab("input, x") +
-  theme_bw()
-fig2a
+ggplot(dat,aes(x=x,y=value)) +
+  geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.4) 
 ```
 
-### Posterior distribution given the data
+### Posterior distribution conditional on data
 
@@
 
-## No observation noise
+No observation noise
+--------------------
 
@@
 ```{r}
-values <- mvrnorm(3, Ef, Cf)
+values <- mvrnorm(200, Ef, Cf)
 ```
@@
 ```{r}
-dat <- data.frame(x=x_predict, t(values))
-dat <- melt(dat, id="x")
+dat <- data.frame(x=x_predict, t(values)) %>%
+  tidyr::pivot_longer(-x, names_to = "rep", values_to = "value") %>% 
+  mutate(rep = as.numeric(as.factor(rep)))
 
@@
 ggplot(dat,aes(x=x,y=value)) + 
+  #geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.2) + #REPLICATES
   geom_ribbon(data = gp, 
@@
                   ymin = Ef - sigma, 
-                  ymax = Ef + sigma
-              ),
-              fill="grey80") +
-  geom_line(aes(color=variable)) + #REPLICATES
+                  ymax = Ef + sigma),
+              fill="blue", alpha = 0.1) +
   geom_line(dat = gp, aes(x=x,y=Ef), size=1) + #MEAN
@@
 ```
-
-![plot of draws from the posterior, with no proces noise](http://farm9.staticflickr.com/8237/8591045344_918138a53a_o.png)
 
@@
 
-fig2c <- ggplot(dat,aes(x=x,y=value)) +
+ggplot(dat,aes(x=x,y=value)) +
   geom_ribbon(data=gp, 
@@
   xlab("input, x")
-fig2c
 ```
 
-Note that unlike the previous case, the posterior no longer collapses completely around the neighborhood of the test points.
+------
+
+# greta
+
+Can we express the same operations in greta?
+
+```{r, message = FALSE, warning = FALSE}
+library(tidyverse)
+library(greta)
+library(greta.gp)
+```
+
+```{r}
+# kernel & GP
+y <- obs$y
+x <- obs$x
+
+# hyperparameters
+rbf_var <- 10 # lognormal(0, 1)
+rbf_len <- 0.5 #lognormal(0, .01)
+obs_sd <- .01 #lognormal(0, .01)
+
+# kernel & GP
+kernel <- rbf(rbf_len, rbf_var)
+f <- gp(x, kernel)
+
+# likelihood -- not clear how this influences f_plot?
+distribution(y) <- normal(f, obs_sd)
+
+# prediction
+f_plot <- project(f, x_plot)
+```
+
+```{r}
+# fit the model by Hamiltonian Monte Carlo
+m <- model(f_plot)
+draws <- mcmc(m, warmup = 0)
+```
+```

### 2019-10-26 notebook/greta-gp.Rmd
greta gp exploration

@@
 # simulate toy data
-#x <- runif(20, 0, 10)
-#y <- sin(x) + rnorm(20, 0, 0.5)
-#x_plot <- seq(-1, 11, length.out = 200)
+n<- 10
+x <- runif(n, 0, 10)
+y <- sin(x) + rnorm(n, 0, 0.5)
+x_plot <- seq(-1, 14, length.out = 200)
 ```
@@
 # hyperparameters
-rbf_var <- lognormal(0, 1)
-rbf_len <- lognormal(0, .01)
-obs_sd <- lognormal(0, .01)
+rbf_var <- 10 # lognormal(0, 1)
+rbf_len <- 0.5 #lognormal(0, .01)
+obs_sd <- .01 #lognormal(0, .01)
 
 # kernel & GP
-kernel <- rbf(rbf_len, rbf_var) + bias(1)
+kernel <- rbf(rbf_len, rbf_var)
 f <- gp(x, kernel)
 
-# likelihood
+# likelihood -- not clear how this influences f_plot?
 distribution(y) <- normal(f, obs_sd)
@@
 m <- model(f_plot)
-draws <- mcmc(m, one_by_one = TRUE)
+draws <- mcmc(m, warmup = 0)
 ```
@@
 # plot 200 posterior samples
-#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 10), ylim = c(-2.5, 2.5))
+plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))
 
-plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
+#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
 for (i in 1:200) {
@@
 }
+points(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))
+
 ```

### 2019-10-26 notebook/python-reticulate-gp.Rmd
greta gp exploration

---
output: github_markdown
---

```{python}
import numpy as np
import tensorflow as tf
import tensorflow_probability as tfp
```

```{python}
tfd = tfp.distributions
psd_kernels = tfp.positive_semidefinite_kernels
```

```{python}
num_points = 100
# Index points should be a collection (100, here) of feature vectors. In this
# example, we're using 1-d vectors, so we just need to reshape the output from
# np.linspace, to give a shape of (100, 1).
index_points = np.expand_dims(np.linspace(-1., 1., num_points), -1)

# Define a kernel with default parameters.
kernel = psd_kernels.ExponentiatedQuadratic()
```

```{python}
gp = tfd.GaussianProcess(kernel, index_points)

samples = gp.sample(10)
# ==> 10 independently drawn, joint samples at `index_points`

noisy_gp = tfd.GaussianProcess(
    kernel=kernel,
    index_points=index_points,
    observation_noise_variance=.05)
noisy_samples = noisy_gp.sample(10)

noisy_samples
```

### 2019-10-28 manuscript/appendix/appendix.Rmd
greta, tensorflow GP, python, etc

@@
 ```{r}
+write_csv(sim_data, "../../data/ghost_sims.csv.gz")
 sim_data %>% group_by(reps) %>%
@@
   bind_cols(y)
+
+write_csv(arima_forecast, "../../data/arima_forecast.csv.gz")
 ```
@@
 ```{r arima_figure, message=FALSE, warning=FALSE}
+
 train <- arima_forecast %>% filter(t < 2000) 
@@
   geom_line(aes(y=pred), lwd = 1, lty = 3, col = "red", data = test) + 
-  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), alpha = 0.4, fill = "red", data = test)  + 
+  geom_ribbon(aes(y=pred, ymin=pred-se, ymax=pred+se), 
+              alpha = 0.4, fill = "red", data = test)  + 
   geom_line(aes(y=x), data = test, lty = 2, col="blue") +
@@
 
-# a <- 2.3
-r <- p$r
-Q <- p$Q
-K <- p$K
-H <- p$H
-sigma <- p$sigma
-
 a <- uniform(0, 10)
-r <- uniform(0, 2 * p$r)
-#Q <- uniform(0, 10)
-K <- uniform(0, 2*p$K)
-H <- uniform(0,  2*p$H)
-sigma <- uniform(0, 2 * p$sigma)
+r <- uniform(0, 4 * p$r)
+Q <- uniform(0, 4 * p$Q)
+K <- uniform(0, 4 * p$K)
+H <- uniform(0,  4 * p$H)
+sigma <- uniform(0, 4 * p$sigma)
 
@@
   mutate(set = "true")
+```
 
+```{r}
 predicted_forecast <- 
@@
 ```{r}
-bind_rows(predicted_forecast, true_forecast) %>% 
+model_forecast <- bind_rows(predicted_forecast, true_forecast)
+write_csv(model_forecast, "../../data/model_forecast.csv.gz")
+
+model_forecast %>% 
 #predicted_forecast %>%
   ggplot(aes(t,x)) +
-  geom_line(aes(group = interaction(reps, set), col = set), alpha = .05) + 
+  geom_line(aes(group = interaction(reps, set), col = set), alpha = .1) + 
   geom_line(data = training_data) + 

### 2019-10-28 notebook/gp-manual-vs-greta.Rmd
greta, tensorflow GP, python, etc

@@
 ```{r setup, message = FALSE, warning = FALSE}
-
-require(MASS)
-require(reshape2)
-require(ggplot2)
+library(MASS)
+library(tidyverse)
 
@@
               fill="grey80") + # Var
-  geom_line(aes(color=variable)) + #REPLICATES
+  geom_line(aes(group=variable), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) + #REPLICATES
   geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
@@
 # hyperparameters
-rbf_var <- 10 # lognormal(0, 1)
-rbf_len <- 0.5 #lognormal(0, .01)
-obs_sd <- .01 #lognormal(0, .01)
+rbf_var <- lognormal(0, 1)
+rbf_len <- lognormal(0, 1)
+obs_sd <- lognormal(0, 1)
 
@@
 # likelihood -- not clear how this influences f_plot?
-distribution(y) <- normal(f, obs_sd)
+distribution(y)<- normal(f, obs_sd)
 
 # prediction
-f_plot <- project(f, x_plot)
+f_plot <- project(f, x_predict)
 ```
@@
 # fit the model by Hamiltonian Monte Carlo
-m <- model(f_plot)
-draws <- mcmc(m, warmup = 0)
+m <- model(f_plot, rbf_len, rbf_var, obs_sd)
+draws <- mcmc(m, chains = 1, one_by_one = TRUE)
 ```
+
+```{r}
+df <- draws[[1]] %>% as_tibble(.name_repair = "universal") 
+hyperpars <- df %>% dplyr::select(rbf_len, obs_sd, rbf_var)  
+posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)
+
+hyperpars %>% colMeans()
+#hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
 ```
+
+```{r}
+posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
+  ggplot(aes(x,value)) + 
+  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
+  geom_point(data = obs, aes(x,y))
+```
+
+```{r}
+```

### 2019-10-28 notebook/greta-gp.Rmd
greta, tensorflow GP, python, etc

@@
 
+## Model Definition
+
+```{r}
+p <- list(r = .05, K = 200, Q = 5, H = 38, sigma = .002, a=2.3, N = 3e3, x0 = 6)
+```
+
+```{r}
+growth <- function(x, p) x * p$r * (1 - x / p$K)
+consumption <- function(x, p) p$a * x ^ p$Q / (x^p$Q + p$H^p$Q)
+may <- function(x, p) x + growth(x,p) - consumption(x,p)
+```
+
+## Model simulations
+
+```{r}
+# Generic simulator routine
+sim <- function(f, p){
+  x <- numeric(p$N)
+  x[1] <- p$x0
+  dBt <- numeric(p$N)
+  if(p$sigma > 0) dBt <- rnorm(p$N, 0, p$sigma)
+  for(t in 1:(p$N-1)){
+    x[t+1] <- max(f(x[t], p) + x[t] * dBt[t], 0)
+  }
+  data.frame(t = 1:p$N, x)
+}
+```
+
+Deterministic simulations:
+
+```{r}
+p_det <- p
+p_det$sigma <- 0
+det <- sim(may,p_det) 
+det_diffs <- det %>% mutate(y = lead(x)-x, case = "det")
+det_diffs <- det %>% mutate(y = c(diff(x),NA), case = "det")
+
+ex <- sim(may,p) 
+ex %>% mutate(y = lead(x)-x, case = "stoch")  %>% bind_rows(det_diffs) %>%
+  ggplot(aes(x,y, col=case)) + geom_point()
+```
+
+```{r}
+ghost <- readr::read_csv("../data/ghost_sims.csv.gz")
+wide <- ghost %>% pivot_wider(names_from = reps, values_from = x) %>% select(-t)
+n <- dim(wide)[1]
+x_t1 <- wide[-1,]
+x_t <- wide[-n,] 
+y <- x_t1 - x_t
+Y <- y %>% apply(1, mean, na.rm=TRUE)
+X <- x_t %>% apply(1, mean, na.rm=TRUE)
+
+tibble(x=X,y=Y, case="mean") %>% bind_rows(det_diffs) %>% ggplot(aes(x,y, col=case)) + geom_point()
+
+ghost %>% group_by(reps) %>%  mutate(y = lead(x)-x) %>% 
+  ggplot(aes(x,y)) + geom_point(size=.1, alpha=0.005)
+```
+
 ```{r}
 arima_forecast <- readr::read_csv("../data/arima_forecast.csv.gz")
-train <- arima_forecast  %>% filter(t %in% seq(0, 2000, by = 5))
+train <- arima_forecast  #%>% filter(t %in% seq(0, 2000, by = 5))
 
@@
 x <- c(0, x_t)
-x_plot <- seq(0, 2, length.out = 200)
+x_predict <- seq(0, 2, length.out = 200)
+
+tibble(x,y) %>% ggplot(aes(x,y)) + geom_point()
 ```
 
-```{r}
-
-```
+----
 
 ```{r}
-# simulate toy data
-n<- 10
-x <- runif(n, 0, 10)
-y <- sin(x) + rnorm(n, 0, 0.5)
-x_plot <- seq(-1, 14, length.out = 200)
-```
-
-```{r}
+# kernel & GP
+y <- diff(ex$x)
+x <- ex$x[-n]
+ggplot(data.frame(x,y), aes(x,y)) + geom_point()
 
 # hyperparameters
-rbf_var <- 10 # lognormal(0, 1)
-rbf_len <- 0.5 #lognormal(0, .01)
-obs_sd <- .01 #lognormal(0, .01)
+rbf_var <- lognormal(0, 1)
+rbf_len <- lognormal(0, 10)
+obs_sd <- lognormal(0, .1)
 
@@
 # fit the model by Hamiltonian Monte Carlo
-m <- model(f_plot)
-draws <- mcmc(m, warmup = 0)
+m <- model(f_plot, rbf_len, rbf_var, obs_sd)
+draws <- mcmc(m, chains = 1, one_by_one = TRUE)
 ```
@@
 ```{r}
-# plot 200 posterior samples
-plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))
+df <- draws[[1]] %>% as_tibble(.name_repair = "universal") 
+hyperpars <- df %>% dplyr::select(rbf_len, obs_sd, rbf_var)  
+posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)
 
-#plot(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 2), ylim = c(-0.1, 0.15))
-for (i in 1:200) {
-  lines(draws[[1]][i, ] ~ x_plot,
-        lwd = 2,
-        col = rgb(0.7, 0.1, 0.4, 0.1))  
-}
-points(y ~ x, pch = 16, col = grey(0.4), xlim = c(0, 14), ylim = c(-3, 3))
+hyperpars %>% colMeans()
+#hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
+```
 
+```{r}
+posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
+  ggplot(aes(x,value)) + 
+  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
+  geom_point(data = obs, aes(x,y))
 ```

### 2019-10-28 notebook/python-reticulate-gp.Rmd
greta, tensorflow GP, python, etc

@@
 
-```{python}
-import numpy as np
-import tensorflow as tf
-import tensorflow_probability as tfp
+```{r message=FALSE}
+library(tidyverse)
+library(reticulate)
+library(tensorflow)
+np <- import("numpy")
+tf <- import("tensorflow")
+tfp <- import("tensorflow_probability")
+sess <- tf$Session()
+
+```
+
+```{r}
+tfd <- tfp$distributions
+psd_kernels <- tfp$positive_semidefinite_kernels
+```
+
+## Draws from Prior
+
+```{r}
+num_points = 100L
+# Index points should be a collection (100, here) of feature vectors. In this
+# example, we're using 1-d vectors, so we just need to reshape the output from
+# np.linspace, to give a shape of (100, 1).
+index_points = np$expand_dims(np$linspace(-1., 1., num_points), -1L)
+
+var <- tf$constant(1.0, dtype = tf$float64)
+len <-  tf$constant(0.1, dtype = tf$float64)
+obs <-  tf$constant(0.05, dtype = tf$float64)
+
+# Define a kernel with default parameters.
+kernel = psd_kernels$ExponentiatedQuadratic(var, len)## Radial basis function
+  #psd_kernels$ExponentiatedQuadratic()
+```
+
+```{r}
+gp = tfd$GaussianProcess(kernel, index_points)
+
+samples = gp$sample(10L)
+# ==> 10 independently drawn, joint samples at `index_points`
+```
+
+Or with observation error:
+
+```{r}
+noisy_gp = tfd$GaussianProcess(
+    kernel=kernel,
+    index_points=index_points,
+    observation_noise_variance=obs)
+
+noisy_samples = noisy_gp$sample(10L)
+```
+
+Evaluate and extract results
+
+```{r}
+out <- sess$run(samples) 
+```
+
+```{r}
+out %>% 
+  t() %>% 
+  as_tibble %>% 
+  mutate(x = seq(-1, 1, length.out = 100) ) %>%
+  pivot_longer(-x) %>%
+  ggplot(aes(x, value, group = name)) + geom_line(col = "purple", alpha=0.4)
+```
+
+## Gaussian Process Regression
+
+```{r}
+# Generate noisy observations from a known function at some random points.
+observation_noise_variance <- 0.01 #<-  tf$constant(obs, dtype = tf$float64)
+len <-  tf$constant(0.1, dtype = tf$float64)
+var <-  tf$constant(1, dtype = tf$float64)
+
+f = function(x) sin(10 * x) * exp(-x ^ 2)
+## Index points and observation_index_points is an array, with 1 col per feature! ## (N-dimensional X)
+observation_index_points = runif(5L, -1, 1) %>% as.matrix(ncol = 1)
+## Observations must be a numeric, not an array. (vector of values Y, not matrix of features)
+observations = f(as.numeric(observation_index_points)) +
+  runif(2L, 0., sqrt(observation_noise_variance))
+
+index_points = seq(-1, 1, len=100L) %>% as.matrix(ncol = 1)
+
+## This works too but no need to it in python, the above is fine!!
+#index_points = np$linspace(-1., 1., 100L)  %>%  np$expand_dims(-1L)
+#x = np$random$uniform(-1., 1., 5L) 
+#observations = (np$sin(10*x) * np$exp(-x**2) + 
+#                  np$random$normal(0., np$sqrt(observation_noise_variance)))
+#observation_index_points <- x %>%  np$expand_dims(-1L)
+
+kernel = psd_kernels$ExponentiatedQuadratic(var, len) # MaternFiveHalves()
+
+gprm = tfd$GaussianProcessRegressionModel(
+    kernel = kernel,
+    index_points = index_points,
+    observation_index_points=observation_index_points,
+    observations = observations,
+    observation_noise_variance = observation_noise_variance)
+
+samples = gprm$sample(10L)
+```
+
+```{r}
+out <- sess$run(samples) 
+```
+
+```{r}
+ob <- data.frame(x = observation_index_points, y = observations, name = "data")
+
+out %>% 
+  t() %>% 
+  as_tibble %>% 
+  mutate(x = index_points) %>%
+  pivot_longer(-x) %>%
+  ggplot(aes(x, value, group = name)) + 
+  geom_line(col = "purple", alpha=0.4) + 
+  geom_point(data = ob, aes(x,y))
+```
+
+## ML
+
+```{r}
+# Define a kernel with trainable parameters. Note we transform the trainable
+# variables to apply a positivity constraint.
+amplitude = tf$exp(tf$Variable(1, dtype=np$float64), name='amplitude')
+length_scale = tf$exp(tf$Variable(2, dtype=np$float64), name='length_scale')
+kernel = psd_kernels$ExponentiatedQuadratic(amplitude, length_scale)
+
+observation_noise_variance = tf$exp(
+    tf$Variable(0.5, dtype=np$float64), name='observation_noise_variance')
+
+# We'll use an unconditioned GP to train the kernel parameters.
+gp = tfd$GaussianProcess(
+    kernel=kernel,
+    index_points= observation_index_points,
+    observation_noise_variance=observation_noise_variance)
+neg_log_likelihood = -gp$log_prob(observations)
+
+optimizer = tf$train$AdamOptimizer(learning_rate=.05, beta1=.5, beta2=.99)
+optimize = optimizer$minimize(neg_log_likelihood)
+
+gprm = tfd$GaussianProcessRegressionModel(
+    kernel=kernel,
+    index_points=index_points,
+    observation_index_points=observation_index_points,
+    observations=observations,
+    observation_noise_variance=observation_noise_variance)
+
+ml_samples = gprm$sample(10L)
+
+```
+
+Note that we need to initialize variables!  
+```{r}
+init_op <- tf$global_variables_initializer()
+with(tf$Session() %as% sess, {
+   # Run the 'init' op
+  sess$run(init_op)
+  out <- sess$run(ml_samples)
+
+  param_est <- sess$run(c(amp = amplitude, 
+                  len = length_scale, 
+                  obs = observation_noise_variance))
+})
+```
+
+```{r}
+x
+ob <- data.frame(x = observation_index_points, y = observations, name = "data")
+
+out %>% 
+  t() %>% 
+  as_tibble %>% 
+  mutate(x = index_points) %>%
+  pivot_longer(-x) %>%
+  ggplot(aes(x, value, group = name)) + 
+  geom_line(col = "purple", alpha=0.4) + 
+  geom_point(data = ob, aes(x,y))
+```
+
+## MCMC
+
+```{r}
+
+joint_log_prob <- function(
+    index_points, observations, amplitude, length_scale, noise_variance){
+  # Hyperparameter Distributions
+  rv_amplitude = tfd$LogNormal(tf$constant(0, dtype=np$float64),
+                               tf$constant(1, dtype=np$float64))
+
+  rv_length_scale = tfd$LogNormal(tf$constant(0, dtype=np$float64), 
+                                  tf$constant(1, dtype=np$float64))
+
+  rv_noise_variance = tfd$LogNormal(tf$constant(0, dtype=np$float64),
+                                    tf$constant(1, dtype=np$float64))
+
+  gp = tfd$GaussianProcess(
+      kernel=psd_kernels$ExponentiatedQuadratic(amplitude, length_scale),
+      index_points=index_points,
+      observation_noise_variance=noise_variance)
+
+  return (
+      rv_amplitude$log_prob(amplitude) +
+      rv_length_scale$log_prob(length_scale) +
+      rv_noise_variance$log_prob(noise_variance) +
+      gp$log_prob(observations)
+  )
+}
+```
+
+```{r}
+scale <- tf$constant(1e-1, dtype=np$float64)
+initial_chain_states = list(
+    scale * tf$ones(1, dtype=np$float64, name='init_amplitude'),
+    scale * tf$ones(1, dtype=np$float64, name='init_length_scale'),
+    scale * tf$ones(1, dtype=np$float64, name='init_obs_noise_variance')
+)
+
+# Since HMC operates over unconstrained space, we need to transform the
+# samples so they live in real-space.
+unconstraining_bijectors = list(
+    tfp$bijectors$Softplus(),
+    tfp$bijectors$Softplus(),
+    tfp$bijectors$Softplus()
+)
+
+unnormalized_log_posterior <- function(amplitude, length_scale, noise_variance){
+
+   joint_log_prob(observation_index_points, 
+                  observations, 
+                  amplitude, 
+                  length_scale,
+                  noise_variance)
+
+}
+
+num_results = 200L
+
+# list(
+  #   list(
+  #     amplitudes,
+  #     length_scales,
+  #     observation_noise_variances
+  #     ),
+  #   kernel_results
+  # )
+```
+
+```{r}
+out = tfp$mcmc$sample_chain(
+  num_results = num_results,
+  num_burnin_steps = 500L,
+  num_steps_between_results = 3L,
+  current_state = initial_chain_states,
+  kernel = tfp$mcmc$TransformedTransitionKernel(
+    inner_kernel = tfp$mcmc$HamiltonianMonteCarlo(
+      target_log_prob_fn = unnormalized_log_posterior,
+      step_size= tf$constant(.15, dtype = np$float64),
+      num_leapfrog_steps = 3L
+    ),
+    bijector = unconstraining_bijectors
+  )
+)
+
+# Now we can sample from the posterior predictive distribution at a new set
+# of index points.
+```
+```{r}
+#with(tf$Session() %as% sess){
+#  out <- sess$run(out)
+#}
+```
+
+```{r}
+gprm = tfd$GaussianProcessRegressionModel(
+    # Batch of `num_results` kernels parameterized by the MCMC samples.
+    kernel=psd_kernels$ExponentiatedQuadratic(out[[0]][[1]], out[[0]][[2]]),
+    index_points=index_points,
+    observation_index_points=observation_index_points,
+    observations=observations,
+    observation_noise_variance=out[[0]][[3]])
+
+## sample!
+samples = gprm$sample()
+```
+
+```{r}
+samples_ = sess$run(samples)
+
+res = sess$run(c(kernel_results = out, samples = samples))
+
 ```
@@
 ```{python}
-tfd = tfp.distributions
-psd_kernels = tfp.positive_semidefinite_kernels
+  print("Acceptance rate: {}".format(
+      np$mean(kernel_results_.inner_results.is_accepted)))
+
+  # Plot posterior samples and their mean, target function, and observations.
+  plt.plot(np.stack([index_points[:, 0]]*num_results).T,
+           samples_.T,
+           c='r',
+           alpha=.01)
+  plt.plot(index_points[:, 0], np.mean(samples_, axis=0), c='k')
+  plt.plot(index_points[:, 0], f(index_points))
+  plt.scatter(observation_index_points[:, 0], observations)
+}
 ```
-
-```{python}
-num_points = 100
-# Index points should be a collection (100, here) of feature vectors. In this
-# example, we're using 1-d vectors, so we just need to reshape the output from
-# np.linspace, to give a shape of (100, 1).
-index_points = np.expand_dims(np.linspace(-1., 1., num_points), -1)
-
-# Define a kernel with default parameters.
-kernel = psd_kernels.ExponentiatedQuadratic()
-```
-
-```{python}
-gp = tfd.GaussianProcess(kernel, index_points)
-
-samples = gp.sample(10)
-# ==> 10 independently drawn, joint samples at `index_points`
-
-noisy_gp = tfd.GaussianProcess(
-    kernel=kernel,
-    index_points=index_points,
-    observation_noise_variance=.05)
-noisy_samples = noisy_gp.sample(10)
-
-noisy_samples
-```

### 2019-10-29 notebook/gp-manual-vs-greta.Rmd
gp greta exploration

@@
 code: true
-output: html_document
+output: github_document
 ---
@@
 
-and plot our solution (mean, 2 standard deviations, and the ranom samples.)
+and plot our solution (mean in black, 2 standard deviations in grey riboon, and the 200 random samples in purple.)
 
@@
 ggplot(dat,aes(x=x,y=value)) + 
-  #geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.2) + #REPLICATES
+  geom_line(aes(group=rep), color =  rgb(0.7, 0.1, 0.4), alpha = 0.2) + #REPLICATES
   geom_ribbon(data = gp, 
@@
                   ymax = Ef + sigma),
-              fill="blue", alpha = 0.1) +
+              fill="grey", alpha = 0.4) +
   geom_line(dat = gp, aes(x=x,y=Ef), size=1) + #MEAN
@@
 
-and plot
+and plot (both sample curves, in purple, and the theoretical expected variance as a grey ribbon)
 
@@
 dat <- data.frame(x=x_predict, t(values))
-dat <- melt(dat, id="x")
+dat <- reshape2::melt(dat, id="x")
 
@@
 ```
+
+# Try greta with fixed parameters
+
+Here I try to replicate the above results under 'no additive noise' by setting the noise term to (almost) zero.  (trying `obs_sd = 0` causes the MCMC to fail to find initial values)
+
+```{r}
+# kernel & GP
+y <- obs$y
+x <- obs$x
+
+# hyperparameters
+rbf_var <- 1
+rbf_len <- 1
+obs_sd <- 0.001
+
+# kernel & GP
+kernel <- rbf(rbf_len, rbf_var)
+f <- gp(x, kernel)
+
+# likelihood -- not clear how this influences f_plot?
+distribution(y)<- normal(f, obs_sd)
+
+# prediction
+f_plot <- project(f, x_predict)
+```
+
+```{r}
+# fit the model by Hamiltonian Monte Carlo
+m <- model(f_plot)
+draws <- mcmc(m, chains = 1, one_by_one = TRUE)
+```
+
+Unfortunately, instead of causing the GP to collapse down only at the data points, like in the example above, this cause the GP samples to collapse entirely:
+
+```{r}
+draws[[1]] %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% 
+  mutate(x = x_predict) %>% pivot_longer(-x)  %>%
+  ggplot(aes(x,value)) + 
+  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
+  geom_point(data = obs, aes(x,y))
+```
+
+# MCMC with greta
 

### 2019-10-29 notebook/greta-gp.Rmd
gp greta exploration

@@
 ```{r setup}
-library(tidyverse)
 library(greta)
@@
 #tensorflow::use_session_with_seed(1234)
+library(tidyverse)
 
@@
 
-```{r}
+```{r eval=FALSE}
 ghost <- readr::read_csv("../data/ghost_sims.csv.gz")
-wide <- ghost %>% pivot_wider(names_from = reps, values_from = x) %>% select(-t)
+wide <- ghost %>% dplyr::select(-t) %>% pivot_wider(names_from = reps, values_from = x) 
 n <- dim(wide)[1]
@@
 # fit the model by Hamiltonian Monte Carlo
-m <- model(f_plot, rbf_len, rbf_var, obs_sd)
+m <- model(f, rbf_len, rbf_var, obs_sd)
 draws <- mcmc(m, chains = 1, one_by_one = TRUE)

### 2019-10-30 notebook/gp-manual-vs-greta.Rmd
Exporing the "two fold path" in GPs
Now that Nick has set me straight here on the differences!

@@
 # kernel & GP
-y <- obs$y
+y <- t(obs$y)
 x <- obs$x
@@
 
-# likelihood -- not clear how this influences f_plot?
-distribution(y)<- normal(f, obs_sd)
+n <- length(x)
+## More explicit
+distribution(y) <- multivariate_normal(zeros(1, n), kernel(x, x) + diag(n) * obs_sd)
+
+## This is more generic, we could have any function here.  Most methods assume additive Gaussian
+## noise to be able to exploit analytical sampling of the GP
+#distribution(y)<- normal(f, obs_sd)
 
@@
 ```{r}
-# fit the model by Hamiltonian Monte Carlo
-m <- model(f_plot, rbf_len, rbf_var, obs_sd)
+# fit the model by Hamiltonian Monte Carlo over hyperparameters only, not sampling from the GP
+m <- model(rbf_len, rbf_var, obs_sd)
 draws <- mcmc(m, chains = 1, one_by_one = TRUE)
@@
 hyperpars <- df %>% dplyr::select(rbf_len, obs_sd, rbf_var)  
-posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)
+#posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)
 
-hyperpars %>% colMeans()
-#hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
+posterior_means <- hyperpars %>% colMeans()
+posterior_means
+hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
+```
+
+And wow, we can do this math in greta...
+
+```{r}
+cov <- kernel
+sigma.n <- posterior_means["obs_sd"]
+cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
+Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% greta_array(obs$y)
+Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
 ```
@@
 ```{r}
-posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
-  ggplot(aes(x,value)) + 
-  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
-  geom_point(data = obs, aes(x,y))
+#posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
+#  ggplot(aes(x,value)) + 
+#  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
+#  geom_point(data = obs, aes(x,y))
 ```

### 2019-11-01 notebook/complete-discrete-version.Rmd
rename and clean up

--- 2019-10-09 notebook/appendix.Rmd
@@
 
-damage <- 0.05
+damage <- .5
 control <- 1
-discount <- 0.999
-
+endemic <- 50
+discount <- 0.999 # Note that dynamics are not on scale of years but ~ days
+e <- .4
 n_s <- 121
@@
   function(x, h = 0){ # May
-    y <- x - h
+    y <- x - e * h
     pmax(
@@
 
-reward_fn <- function(x,h) - damage * (x / 100) ^ 2 - control * (h / 100)
+## safe state
+##
+
+reward_fn <- function(x,h) - (damage * pmax(x - endemic, 0)) ^ 2  - (control * h)
+# reward_fn <- function(x,h) - damage * (x / 100) ^ 2 - control * (h / 100) ^ 2
+#
 
@@
 
-true_a <- 27.2      ### 21.5
-believe_a <- 29 # 28.5   ### 22.5
+true_a <- 27.4      ### 21.5
+believe_a <- 28.5 # 28.5   ### 22.5
 
@@
 
-# Learning
+# Planning
 
 ```{r}
-x0 <- which.min(abs(states - 25))
+x0 <- which.min(abs(states - 5))
+Tmax <- 700
 ```
@@
 df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = belief_p$policy, x0 = x0, Tmax = 100)
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
+                   policy = belief_p$policy, x0 = x0, Tmax = Tmax)
+```
+
+```{r}
+df %>% 
+  mutate(state = states[state], action = actions[action] * 5) %>% 
+  select(-obs,-value) %>%
+  pivot_longer(-time) %>%
+  ggplot(aes(time, value, col=name)) + 
+  geom_point() + geom_path() #+ 
+  #facet_wrap(~name, scales="free_y", ncol=1)
+
 ```
@@
 df <- mdp_planning(m_belief, reward, discount, model_prior = c(1), 
-                   policy = belief_p$policy, x0 = x0, Tmax = 1000)
+                   policy = belief_p$policy, x0 = x0, Tmax = Tmax)
 
-df %>% mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + geom_point()+geom_path() + 
-  geom_line(aes(time, action), col="blue")
+df %>% 
+  mutate(state = states[state], action = actions[action] * 5) %>% 
+  select(-obs,-value) %>%
+  pivot_longer(-time) %>%
+  ggplot(aes(time, value, col=name)) + 
+  geom_point() + geom_path()
 ```
@@
 df <- mdp_planning(m_true, reward, discount, model_prior = c(1), 
-                   policy = true_p$policy, x0 = x0, Tmax = 1000)
+                   policy = true_p$policy, x0 = x0, Tmax = Tmax)
 
 df %>% 
-  mutate(state = states[state], action = actions[action]) %>% 
-  ggplot(aes(time, state)) + 
-  geom_point(alpha=0.5)  + 
-  geom_line(aes(time, action), col="blue")
+  mutate(state = states[state], action = actions[action] * 5) %>% 
+  select(-obs,-value) %>%
+  pivot_longer(-time) %>%
+  ggplot(aes(time, value, col=name)) + 
+  geom_point() + geom_path()
 ```
@@
 wd <- sd(possible_a)
-prior <- dnorm(possible_a, believe_a, wd/2)
+prior <- dnorm(possible_a, believe_a, wd/3)
 prior <- prior / sum(prior)
@@
                       x0 = x0, 
-                      Tmax = Tmax, 
+                      Tmax = 20, 
                       true_transition = transition[[true_i]],
@@
   gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = as.numeric(param)) %>% 
+  mutate(param = possible_a[as.integer(param)]) %>% 
   ggplot(aes(param, probability, group = time, alpha = time)) + 

### 2019-11-01 notebook/gp-manual-vs-greta.Rmd
rename and clean up

@@
 sigma.n <- posterior_means["obs_sd"]
+
 cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
@@
 Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
+
+Ef <- calculate(Ef, 
+                values = list(rbf_len = posterior_means["rbf_len"], 
+                              rbf_var = posterior_means["rbf_var"])) %>% as.numeric()
+Cf <- calculate(Cf, 
+                values = list(rbf_len = posterior_means["rbf_len"], 
+                              rbf_var = posterior_means["rbf_var"]))
+```
+
+```{r}
+## Theory/Analytic
+gp <- data.frame(x_predict = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )
+## Draws
+values <- mvrnorm(200, Ef, Cf)
+dat <- data.frame(x=x_predict, t(values)) %>%  pivot_longer(-x)
+```
+
+```{r}
+ggplot(dat,aes(x=x,y=value)) +
+  geom_ribbon(data=gp, 
+              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
+              fill="grey80") + # Var
+  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) + #REPLICATES
+  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
+  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
+  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
+  xlab("input, x")
 ```

### 2019-11-01 notebook/greta-gp.Rmd
rename and clean up

@@
 x <- ex$x[-n]
-ggplot(data.frame(x,y), aes(x,y)) + geom_point()
+obs <- data.frame(x,y)
+ggplot(obs, aes(x,y)) + geom_point()
+```
 
+```{r}
 # hyperparameters
@@
 # prediction
+x_predict <- seq(0, 150, len=200)
 f_plot <- project(f, x_predict)
@@
 ```{r}
-# fit the model by Hamiltonian Monte Carlo
-m <- model(f, rbf_len, rbf_var, obs_sd)
+# fit the model by Hamiltonian Monte Carlo over hyperparameters only, not sampling from the GP
+m <- model(rbf_len, rbf_var, obs_sd)
 draws <- mcmc(m, chains = 1, one_by_one = TRUE)
@@
 df <- draws[[1]] %>% as_tibble(.name_repair = "universal") 
+readr::write_csv(df, "draws.csv.gz")
 hyperpars <- df %>% dplyr::select(rbf_len, obs_sd, rbf_var)
-posterior <- df %>% dplyr::select(-rbf_len, -obs_sd, -rbf_var)
-
-hyperpars %>% colMeans()
-#hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
+posterior_means <- hyperpars %>% colMeans()
+posterior_means
+hyperpars %>% gather() %>% ggplot(aes(x = value, fill=key)) + stat_density() + facet_wrap(~key)
 ```
@@
 ```{r}
-posterior %>% t() %>% as_tibble() %>% dplyr::select(1:200) %>% mutate(x = x_predict) %>% pivot_longer(-x)  %>%
-  ggplot(aes(x,value)) + 
-  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) +
-  geom_point(data = obs, aes(x,y))
+cov <- kernel
+sigma.n <- posterior_means["obs_sd"]
+
+cov_xx_inv <- solve(cov(obs$x, obs$x) + sigma.n^2 * diag(1, length(obs$x)))
+Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% greta_array(obs$y)
+Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
+
+Ef <- calculate(Ef, 
+                values = list(rbf_len = posterior_means["rbf_len"], 
+                              rbf_var = posterior_means["rbf_var"])) %>% as.numeric()
+Cf <- calculate(Cf, 
+                values = list(rbf_len = posterior_means["rbf_len"], 
+                              rbf_var = posterior_means["rbf_var"]))
 ```
@@
 ```{r}
+## Theory/Analytic
+gp <- data.frame(x_predict = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )
+## Draws
+
+values <- MASS::mvrnorm(200, Ef, Cf)
+dat <- data.frame(x=x_predict, t(values)) %>%  pivot_longer(-x)
 ```
+
+```{r}
+ggplot(dat,aes(x=x,y=value)) +
+  geom_ribbon(data=gp, 
+              aes(x=x_predict, y=Ef, ymin=Ef-sigma, ymax=Ef+sigma),
+              fill="grey80") + # Var
+  geom_line(aes(group=name), alpha=0.3, col=rgb(0.7, 0.1, 0.4, 0.1)) + #REPLICATES
+  geom_line(data=gp,aes(x=x_predict,y=Ef), size=1) + #MEAN
+  geom_point(data=obs,aes(x=x,y=y)) +  #OBSERVED DATA
+  scale_y_continuous(lim=c(-3,3), name="output, f(x)") +
+  xlab("input, x")
+```
+
+## Forecast from the estimated process
+
+```{r}
+
+## recompute
+gp_forecast <- function(kernel, obs, x_t, steps){
+
+  cov_xx_inv <- solve(cov(obs$x, obs$x))
+Ef <- cov(x_predict, obs$x) %*% cov_xx_inv %*% obs$y
+Cf <- cov(x_predict, x_predict) - cov(x_predict, obs$x)  %*% cov_xx_inv %*% cov(obs$x, x_predict)
+
+}
+
+## grid forecast: points in x_predict
+gp <- data.frame(x = x_predict, Ef = Ef, sigma = 2*sqrt(diag(Cf)) )
+
+```
+
+## Compute decisions from the estimated process

### 2019-11-01 notebook/make.R

@@ -1,19 +1,24 @@
 library(callr)
 
 
-p1 <- callr::r_bg(function() rmarkdown::render("notebook/ghost-sims.Rmd") )
+p1 <- callr::r_bg(function() rmarkdown::render("notebook/greta-gp.Rmd") )
 
-#  needs to wait for p1 to complete
-p2 <- callr::r_bg(function() rmarkdown::render("notebook/ghost-fit.Rmd") )
+p3 <- callr::r_bg(function() rmarkdown::render(here::here("manuscript/appendix/appendix.Rmd")) )
+
+
+
+p4 <- callr::r_bg(function() rmarkdown::render(here::here("notebook/complete-discrete-version.Rmd")),
+                  #env = c(LD_PRELOAD="libnvblas.so")
+                  )
+
+
 
 # greta tasks probably cannot run in parallel given current memory settings
-p3 <- callr::r_bg(function() rmarkdown::render("notebook/discrete-sims.Rmd") )
 
 
 
 p4 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost.Rmd"),
                   env = c(LD_PRELOAD="libnvblas.so"))
-
 
 
 p5 <- callr::r_bg(function() rmarkdown::render("notebook/may-outbreak-ghost-2.Rmd") )

### 2019-11-05 notebook/complete-discrete-version.Rmd
longer run

@@
 ```{r}
+set.seed(12345)
+Tmax_learning <- 500
 mdp_learning_ <- memoise::memoise(mdp_learning)
@@
                       x0 = x0, 
-                      Tmax = 20, 
+                      Tmax = Tmax_learning, 
                       true_transition = transition[[true_i]],
@@
 
+```
+
+```{r}
+saveRDS(learning_sim, "learning_sim.rds")
 ```
@@
  learning_sim$posterior %>% 
-  data.frame(time = 1:Tmax) %>%
-  filter(time %in% seq(1,Tmax, by = 5)) %>%
+  data.frame(time = 1:Tmax_learning) %>%
+  filter(time %in% seq(1,Tmax_learning, by = 5)) %>%
   gather(param, probability, -time, factor_key =TRUE) %>% 

### 2019-11-05 notebook/make.R

@@ -1,5 +1,8 @@
 library(callr)
 
+p4 <- callr::r_bg(function() rmarkdown::render(here::here("notebook/complete-discrete-version.Rmd"))
+                  #env = c(LD_PRELOAD="libnvblas.so")
+)
 
 p1 <- callr::r_bg(function() rmarkdown::render("notebook/greta-gp.Rmd") )
 
@@ -7,9 +10,6 @@ p3 <- callr::r_bg(function() rmarkdown::render(here::here("manuscript/appendix/a
 
 
 
-p4 <- callr::r_bg(function() rmarkdown::render(here::here("notebook/complete-discrete-version.Rmd")),
-                  #env = c(LD_PRELOAD="libnvblas.so")
-                  )
 
 
 

### 2019-11-05 notebook/complete-discrete-version.Rmd
weaker ghost re-run, shorter runtime.

@@
 
-true_a <- 27.4      ### 21.5
+true_a <- 27      ### 21.5
 believe_a <- 28.5 # 28.5   ### 22.5
@@
   ggplot(aes(x, f)) + 
-    geom_point(aes(col = a), alpha = 0.1) + 
+    geom_point(aes(col = a), alpha = 0.8) + 
     geom_line(aes(lty = group), data = df %>% filter(a == true_a | a == believe_a), 
               lwd = 1, show.legend = FALSE) +
-    geom_hline(aes(yintercept = 0))
+    geom_hline(aes(yintercept = 0)) + scale_color_viridis_c()
 
@@
 x0 <- which.min(abs(states - 5))
-Tmax <- 700
+Tmax <- 200
 ```
@@
 set.seed(12345)
-Tmax_learning <- 500
+Tmax_learning <- 50
 mdp_learning_ <- memoise::memoise(mdp_learning)
@@
 ```{r mcmc, cache=TRUE}
-if(!file.exists("samples.tsv.gz")){ # Manually cache the slow step
 
@@
 
-readr::write_tsv(samples, "samples.tsv.gz")
-
-} else {
-  samples <- readr::read_tsv("samples.tsv.gz")
-}
 ```

### 2019-11-07 manuscript/manuscript.Rmd
more scenarios

--- 2019-10-02 manuscript/manuscript.Rmd
@@
 
-Transient dynamics have long been identified in the ecological systems
-
 Long transients may be quite common features of the dynamics of ecological systems [@Hastings1994; @Hastings2018].
@@
 One of the most intuitive implications for the existence of long transients in the risk in mistaking transient behavior for stable state behavior and the consequences of managing a system under this false conclusion.
-The potential for this mistake can is straightforward to demonstrate even under idealized conditions such as access to accurate, high frequency data and knowledge of the correct model structure.
+The potential for this mistake is straightforward to demonstrate even under idealized conditions such as access to accurate, high frequency data and knowledge of the correct model structure.
 Decision-theoretic approaches to ecological management provide a more nuianced approach for dealing with uncertainty by explicitly considering the set of all actions available to the manager and the likely consequences of those actions over time.

### 2019-11-07 notebook/complete-discrete-version.Rmd
more scenarios

@@
 library(truncnorm)
-
+library(ggthemes)
+pal <- solarized_pal()(3)
 ```
@@
 ```{r}
-#damage <- 0.3
-#control <- 100
-#discount <- 0.99
+## Discrete version of the state space and action space
+n_s <- 121
+states <- seq(0, 120, length = n_s)
+actions <- seq(0, 120, length = n_s)
 
+# Constants in the utility function
 damage <- .5
@@
 discount <- 0.999 # Note that dynamics are not on scale of years but ~ days
-e <- .4
-n_s <- 121
-states <- seq(0, 120, length = n_s)
-actions <- seq(0, 120, length = n_s)
 
-p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, a=2.3, x0 = 20)
+## Model constants -- used to compute transistion probabilities
+efficiency <- .4    
+p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, x0 = 20) # fixed parameters
+```
 
+The ecological dynamics are given by @May1977's consumer-resource model:
+
+```{r}
 may <- function(a){  
   function(x, h = 0){ # May
-    y <- x - e * h
+    y <- x - efficiency * h
     pmax(
@@
 
-## safe state
-##
+```
 
+```{r}
 reward_fn <- function(x,h) - (damage * pmax(x - endemic, 0)) ^ 2  - (control * h)
-# reward_fn <- function(x,h) - damage * (x / 100) ^ 2 - control * (h / 100) ^ 2
-#
-
-#reward_fn <- function(x,h) - damage * (x / max(states)) - control * h / max(actions) * x / max(states)
-
 ```
@@
 
-true_a <- 27      ### 21.5
-believe_a <- 28.5 # 28.5   ### 22.5
+true_a <- 27      ### Weak ghost
+believe_a <- 28.5 ### Attractor
+
+ghost_a <- 27.4     ### Stronger ghost
 
@@
 true_i <- which.min(abs(possible_a - true_a))
+ghost_i <- which.min(abs(possible_a - ghost_a))
+
 ```
@@
 m_belief <- transition_matrix(states, actions, may(believe_a), p$sigma)
+m_ghost <- transition_matrix(states, actions, may(ghost_a), p$sigma)
+
 ```
@@
 
-## a near ghost
+## a weak ghost
 
@@
 learning_sim$posterior %>% 
-  data.frame(time = 1:Tmax_learning) %>%
-  filter(time %in% seq(1,Tmax_learning, by = 5)) %>%
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% c(1,50)) %>%
   gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = possible_a[as.integer(param)]) %>% 
-  ggplot(aes(param, probability, group = time, alpha = time)) + 
-  geom_line()
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+   ylim(0,0.3)
+```
+
+-----
+
+True value is a stronger ghost:
+
+```{r}
+set.seed(12345)
+Tmax_learning <- 50
+mdp_learning_ <- memoise::memoise(mdp_learning)
+  learning_sim <- mdp_learning_(transition, reward, discount, 
+                      x0 = x0, 
+                      Tmax = Tmax_learning, 
+                      true_transition = transition[[ghost_i]],
+                      model_prior = prior,
+                      type = "value iteration", 
+                      epsilon = 1e-2)
+
 ```
@@
 ```{r}
+saveRDS(learning_sim, "learning_sim2.rds")
+```
+
+```{r}
+ learning_sim$df %>% 
+  select(-value) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, states[state], color = series)) + geom_line()
+```
+
+```{r}
+learning_sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% c(1,50)) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+   ylim(0,0.3)
+```
+
+---
+
+## a strong ghost, wide prior:
+
+```{r}
+wd <- sd(possible_a)
+wide_prior <- dnorm(possible_a, believe_a, wd)
+wide_prior <- wide_prior / sum(wide_prior)
+```
+
+```{r}
+data.frame(a = possible_a, probability = wide_prior) %>%
+  ggplot(aes(a,prior)) + geom_bar(stat="identity") +
+  geom_vline(aes(xintercept = true_a), col="red", lwd=1, lty=2) + 
+  geom_vline(aes(xintercept = believe_a), col="blue", lwd=1, lty=2) 
+```
+
+```{r}
+set.seed(12345)
+Tmax_learning <- 50
+mdp_learning_ <- memoise::memoise(mdp_learning)
+  learning_sim <- mdp_learning_(transition, reward, discount, 
+                      x0 = x0, 
+                      Tmax = Tmax_learning, 
+                      true_transition = transition[[ghost_i]],
+                      model_prior = wide_prior,
+                      type = "value iteration", 
+                      epsilon = 1e-2)
+
+```
+
+```{r}
+saveRDS(learning_sim, "learning_sim3.rds")
+```
+
+```{r}
+ learning_sim$df %>% 
+  select(-value) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, states[state], color = series)) + geom_line()
+```
+
+```{r}
+learning_sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% c(1,50)) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+   ylim(0,0.3)
 ```

### 2019-11-07 manuscript/manuscript.Rmd
beetles!

@@
 library(hrbrthemes)
-library(ggplot2)
-library(ggthemes)
 library(Cairo)
@@
 extrafont::loadfonts()
+
+library(ggplot2)
+library(ggthemes)
+library(ggtext) # devtools::install_github("clauswilke/ggtext")
+ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
 pal <- solarized_pal()(3)
@@
 This analysis also highlights the importance of stochasticity has for transient dynamics.
+
+## Model and Methods
+
+I consider a generic model of population dynamics which can exhibit phenomena such as bistability and a ghost attractor dynamics.
+This model is amenable to interpretation in many ecological contexts (sensu @Levins1966 sacrificing precision to realism and generality), such as the collapse of a fish population, recovery of an endagered species, or the outbreak a pest.
+To provide a more tangible context to frame our questions of ecological mangement, I will focus on the latter scenario of a pest outbreak.
+In this scenario, we imagine a species such as the Pine bark beetle, (e.g.
+*Dendroctonus ponderosae*, @Harvey2014) whose recent boom in population has devastated much of Northern America's pine forests.
+The model is a stochastic version of the consumer-resource model of @May1977,
+
+\begin{equation}
+X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{logistic growth}}  - 
+\underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{saturating consumption}} + 
+\underbrace{\xi_t X_t}_{\textrm{environmental noise}} \label{model}
+\end{equation}
+
+Where $X_t$ is the population density of the focal species (the beetle, in our case).
+The growth  rate of the species is determined by an intrinsic growth rate $r$ and carrying capacity $K$.
+The species is under mortality pressure from another species (the natural defences of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$.
+$Q$ determines the steepness of the saturating response.
+The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$ [@Boettiger2018].
+Model dynamics are illustrated in Figure 1.
+
+```{r}
+## Model constants -- used to compute transistion probabilities
+efficiency <- .4    
+p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, x0 = 20) # fixed parameters
+
+may <- function(a){  
+  function(x, h = 0){ # May
+    y <- x - efficiency * h
+    pmax(
+      ## controlling h is controlling the bifurcation point directly...
+      y + y * p$r * (1 - y / p$K)  - a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
+      0)
+  }
+}
+```
+
+```{r figure1, fig.cap="For both curves, r = 0.8, K = 153, Q = 2, H = 20."}
+c(true_a = 27,           ## Weak ghost
+  believe_a = 28.5) %>%  ## Attractor
+  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a)) %>% 
+  mutate(group = as.character(a)) %>%
+  ggplot(aes(x, f)) + 
+    geom_line(aes(lty = group, col = group), lwd = 1) +
+    geom_hline(aes(yintercept = 0)) + 
+    labs(lty = "a", col = "a", y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
+    scale_color_solarized()
+
+```
+
+Under appropriate choice of parameters, the model is bistable, with two stable equilibiria at non-zero population sizes -- a low, "endemic" state, and a much higher "outbreak" state.
+In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emergign from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses.
+This saturating response and the resulting threshhold dynamics in the response of affected trees exhibited by this model is quite well motivated   [e.g.@Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate.
+Initially this results in the strengthening (concave up) portion of the S curve as other defences are induced by the attack.
+However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defences, leading to a saturating response.
+Low beetle densities are thus stable, kept in check by the increasingly strong response of healthy trees.
+Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defences of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting avaliable tree habitat $K$ [@Harvey2014].
+
+Thus our simple dynamical model from \eqref{model} captures the qualitative processes and dynamics expected in the beetle outbreak system.
+Bear in mind that this model is not intended to be a precise, quantitive match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.
+
+We also observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g.
+the vertical axis, Figure 1) is still small in magnitude, and thus dynamics near this region move slowly.
+This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.
+This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.
+
+### Management model
+
+```{r}
+## Discrete version of the state space and action space
+n_s <- 121
+states <- seq(0, 120, length = n_s)
+actions <- seq(0, 120, length = n_s)
+# Constants in the utility function
+damage <- .5
+control <- 1
+endemic <- 50
+discount <- 0.999 # Note that dynamics are not on scale of years but ~ days
+reward_fn <- function(x,h) - (damage * pmax(x - endemic, 0)) ^ 2  - (control * h)
+```
 

### 2019-11-08 manuscript/manuscript.Rmd
WIP

@@
 The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$ [@Boettiger2018].
-Model dynamics are illustrated in Figure 1.
+Model dynamics are illustrated in Figure 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.
 
@@
 
-```{r figure1, fig.cap="For both curves, r = 0.8, K = 153, Q = 2, H = 20."}
-c(true_a = 27,           ## Weak ghost
-  believe_a = 28.5) %>%  ## Attractor
-  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a)) %>% 
-  mutate(group = as.character(a)) %>%
+```{r figure1, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$.
+Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.
+By changing parameter a, the system goes from bistable (red, a = 28.5) to a ghost attractor (blue, a = 27).
+For both curves, r = 0.8, K = 153, Q = 2, H = 20."}
+c(`ghost attractor` = 27,           
+  `bistable attractor` = 28.5) %>%
+  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a), .id = "scenario") %>% 
   ggplot(aes(x, f)) + 
-    geom_line(aes(lty = group, col = group), lwd = 1) +
+    geom_line(aes(lty = scenario, col = scenario), lwd = 1) +
     geom_hline(aes(yintercept = 0)) + 
-    labs(lty = "a", col = "a", y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
+    labs(y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
     scale_color_solarized()
@@
 
-### Management model
+Figure 2 illustrates several simulations from the model, starting from a low density, under both the scenario of a stable endemic state and a ghost attractor.
+Note that it can in general be quite difficult to distinguish dynamics of the ghost from the dynamics of the stable attractor.
 
@@
 actions <- seq(0, 120, length = n_s)
+```
+
+```{r}
+transition_matrix <- function(states, actions, f, sigma){
+    n_s <- length(states)
+    n_a <- length(actions)
+    transition <- array(0, dim = c(n_s, n_s, n_a))
+    for(i in 1:n_a){
+      for (k in 1:n_s) {
+          nextpop <- f(states[k], actions[i])
+          if (nextpop <= 0) {
+            x  <- c(1, rep(0, n_s - 1))
+          } else {
+            x <- truncnorm::dtruncnorm(states, 0, max(states), nextpop, sigma * nextpop)
+            if(sum(x) <= 0){ 
+              x  <- c(1, rep(0, n_s - 1))
+            } else {
+              x <- x / sum(x)
+            }
+          }
+          transition[k, , i] <- x
+      }
+  }
+  if(any(is.na(transition))) stop("error creating transition matrix")
+transition
+}
+```
+
+```{r}
+P_ghost <- transition_matrix(states, actions, may(27.4), p$sigma)
+P_bistable <- transition_matrix(states, actions, may(27.6), p$sigma)
+```
+
+```{r}
+## `actions` is an argument of indices, not necessarily the action values themselves 
+sim <- function (transition,  x0, Tmax, action = rep(1, Tmax)){
+    n_states <- dim(transition)[2]
+    state <- numeric(Tmax + 1)
+    state[1] <- x0
+    time <- 1:Tmax
+    for (t in time) {
+        state[t + 1] <- base::sample(1:n_states, 
+                               1, 
+                               prob = transition[state[t], , action[t]])
+    }
+    data.frame(time = 1:Tmax, state = state[time])
+}
+```
+
+```{r}
+x0 <- which.min(abs(states - p$x0))
+Tmax <- 400
+set.seed(1234)
+reps <- 50
+ghost_sim <- map_dfr(1:reps, function(i)
+  sim(P_ghost, x0, Tmax)  %>% mutate(state = states[state], scenario = "ghost"),
+  .id = "rep")
+bistable_sim <- map_dfr(1:reps, function(i)
+  sim(P_bistable, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
+  .id = "rep")
+
+bind_rows(ghost_sim, bistable_sim) %>%
+  ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
+  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario)
+```
+
+### Management model
+
+To properly understand the consequences transient dynamics have for management, we must first be more explicit about both the management objectives and potential management actions.
+For simplicity, I will imagine a manager has the option to apply a direct mortality to the beetle larvae prior to recruitment, e.g.
+by spraying at some given intensity.
+Thus the action space is defined by some measure of spraying effort between 0% and 100%.
+The management objectives are defined by a utility function, which trades off the costs of control with the costs of damage due to the outbreak.
+Here I assume linearly increasing cost of control (fixed price per unit spraying).
+Below an endemic population size $X_0$ I assume damage (e.g.
+no tree mortality) is minimal, and then increases quadratically with increasing beetle density; that is, the instantaneous utility of state with beetle density $X_t$ and taking action $A_t$ is
+
+\begin{equation}
+U(X_t, A_t) = \underbrace{ - d \max\left(X_t - X_0, 0\right)^2}_{\textrm{damage}} - \underbrace{c A_t}_{control} \label{utility}
+\end{equation}
+
+The manager's objective is to select the sequence of actions $A_t \in \mathbf{A}$ which maximize the net expected discounted ($\delta$) utility,
+
+$$\max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)$$
+where the evolution of $X_t$ is determined by Eq \eqref{model}.
+Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.
+By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Marescot2013].
+
+```{r}
 # Constants in the utility function
@@
 reward_fn <- function(x,h) - (damage * pmax(x - endemic, 0)) ^ 2  - (control * h)
+
+reward <- array(dim=c(length(states), length(actions)))
+for(i in 1:length(states)){
+  for(j in 1:length(actions)){
+    reward[i,j] <- reward_fn(states[i], actions[j])
+  }
+}
+```
+
+## Optimal Planning under transient dynamics vs stable state dynamics
+
+First, I will consider the case that in which the manager has been able to accurately infer the dynamics (i.e.
+correctly estimate the parameters of Eq~\eqref{1}), and assuming the model structure postulated \eqref{1} is correct.
+The optimal strategy determined by SDP can be represented as a "policy" function $p(X)$, indicating for each possible state $X$ what the optimal action $\hat A = p(X)$ would be.
+Figure 3 displays the numerically computed optimal policy for the system with only a transient.
+Though the models differ by only a small amount in a single parameter, because one model corresponds to a system with only a ghost attractor while the other has a stable state attractor, the resulting optimal policy solutions differ significantly.
+Intuitively, in the system with stable state dynamics, the manager does not implement any control at low levels, where the presence of the lower stable state will constrain the beetle population to acceptable levels.
+Once that lower stable state becomes only a ghost attractor, the optimal solution targets elimination of the beetle even at low densities.
+Consequently, the optimal response becomes tightly dependent on our knowledge of the dynamics.
+
+```{r}
+soln_ghost <- mdplearning::mdp_compute_policy(list(P_ghost), reward, discount, max_iter = 1e4, epsilon = 1e-2)
+soln_bistable <-mdplearning::mdp_compute_policy(list(P_bistable), reward, discount, max_iter = 1e4, epsilon = 1e-2)
+```
+
+```{r}
+tibble(state = states,
+       ghost = actions[soln_ghost$policy],
+       bistable =  actions[soln_bistable$policy]
+       ) %>%
+  pivot_longer(c(ghost, bistable), 
+               names_to="model",
+               values_to = "action") %>%
+  ggplot(aes(state,action, color = model)) + 
+  geom_point(alpha=0.6) + 
+  scale_color_solarized()
+```
+
+## Planning over uncertainty
+
+The transition matrix is an expression of conditional probability $P(X_{t+1} | X_t,  A_t)$, the probability that a system in state $X_t$ will transition to state $X_{t+1}$ in the next time interval if the manager takes action $A_t$.
+If we have $n$ possible states and $m$ possible actions, we have $m$ $n \by n$ matrices for the possible transitions.
+The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
+We can represent uncertainty over the models and their parameters as uncertainties as conditional on the model, $M_i$, so that the transistion matrix must then be marganlized over the set of possible models: $$P(X_{t+1} | X_t,  A_t) = \sum_{i} P(X_{t+1} | X_t,  A_t, M_i) P(M_i)$$
+
+```{r}
+soln_dist <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
+                                             model_prior = c(0.5, 0.5),
+                                             max_iter = 1e4, epsilon = 1e-2)
+
+tibble(state = states,
+       action = actions[soln_dist$policy])  %>% 
+  ggplot(aes(state, action)) + geom_point()
+```
+
+```{r}
+set.seed(123456)
+Tmax <- 50
+df <- mdp_planning(P_ghost, reward, discount, 
+                   policy = soln_dist$policy,
+                   x0 = x0, Tmax = Tmax)
+
+df %>% 
+  mutate(state = states[state], action = actions[action] * 5) %>% 
+  select(-obs,-value) %>%
+  pivot_longer(-time) %>%
+  ggplot(aes(time, value, col=name)) + 
+  geom_point() + geom_path()
+```
+
+## Adaptive management
+
+This probability thus depends on our prior belief that model $i$ is true, $P(M_i)$.
+Not only does this allow our manager to make her decision in to reflect uncertainty over models, but it also suggests that having made that decision and observed the outcomes, the manager can update her belief over possible models according to Bayes Rule,
+
+$$P(M_i | X_{t+1} ) = \frac{P(X_{t+1} | M_i ) P(M_i)}{P(X_{t+1})}$$
+that is, the optimal manager not only plans over the uncertainty, but also learns which model is most likely to be correct over the course of management.
+
+```{r}
+set.seed(123456)
+Tmax <- 50
+learning <- mdp_learning(list(P_ghost, P_bistable), reward, discount, 
+                   model_prior = c(0.5, 0.5),
+                   x0 = x0, Tmax = Tmax,
+                  true_transition = P_ghost,
+                  max_iter = 1e4, epsilon = 1e-2)
+
+```
+
+```{r}
+ learning_sim$df %>% 
+  select(-value) %>% 
+  gather(series, state, -time) %>% 
+  ggplot(aes(time, states[state], color = series)) + geom_line()
+```
+
+```{r}
+learning_sim$posterior %>% 
+  data.frame(time = 1:Tmax) %>%
+  filter(time %in% c(1,50)) %>%
+  gather(param, probability, -time, factor_key =TRUE) %>% 
+  mutate(param = as.numeric(param)) %>% 
+  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+   ylim(0,0.3)
 ```

### 2019-11-14 manuscript/manuscript.Rmd
a nice example

@@
 journal: Theoretical ecology
-layout: 5p
+layout: 3p
 bibliography: refs.bib
@@
 library(tidyverse)   # plotting and data manipulation
+library(mdplearning)
 ```
@@
 ```{r}
+## Discrete version of the state space and action space
+n_s <- 121
+states <- seq(0, 120, length = n_s)
+actions <- seq(0, 120, length = n_s)
+```
+
+```{r}
 ## Model constants -- used to compute transistion probabilities
 efficiency <- .4    
-p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .02, x0 = 20) # fixed parameters
+p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .05, x0 = 20) # fixed parameters
 
@@
 ```{r}
-## Discrete version of the state space and action space
-n_s <- 121
-states <- seq(0, 120, length = n_s)
-actions <- seq(0, 120, length = n_s)
-```
-
-```{r}
 transition_matrix <- function(states, actions, f, sigma){
@@
 ```{r}
-P_ghost <- transition_matrix(states, actions, may(27.4), p$sigma)
-P_bistable <- transition_matrix(states, actions, may(27.6), p$sigma)
+P_ghost <- transition_matrix(states, actions, may(27.2), p$sigma)
+P_weak_attractor <- transition_matrix(states, actions, may(28), p$sigma)
+P_bistable <- transition_matrix(states, actions, may(28), p$sigma)
 ```
@@
 x0 <- which.min(abs(states - p$x0))
-Tmax <- 400
-set.seed(1234)
+Tmax <- 200
+set.seed(12345)
 reps <- 50
@@
 bistable_sim <- map_dfr(1:reps, function(i)
-  sim(P_bistable, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
+  sim(P_weak_attractor, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
   .id = "rep")
@@
 soln_ghost <- mdplearning::mdp_compute_policy(list(P_ghost), reward, discount, max_iter = 1e4, epsilon = 1e-2)
-soln_bistable <-mdplearning::mdp_compute_policy(list(P_bistable), reward, discount, max_iter = 1e4, epsilon = 1e-2)
+soln_bistable <-mdplearning::mdp_compute_policy(list(P_bistable), reward, discount, max_iter = 1e4, epsilon = 1e-4)
+
+#soln_ghost <-MDPtoolbox::mdp_value_iteration(P_ghost, reward, discount, max_iter = 5e3, epsilon = 1e-5)
+
 ```
@@
 ```{r}
+## With more noise, these are more similar.  With smaller noise, the ghost policy remains aggressive over a larger range
 tibble(state = states,
@@
 The transition matrix is an expression of conditional probability $P(X_{t+1} | X_t,  A_t)$, the probability that a system in state $X_t$ will transition to state $X_{t+1}$ in the next time interval if the manager takes action $A_t$.
-If we have $n$ possible states and $m$ possible actions, we have $m$ $n \by n$ matrices for the possible transitions.
+If we have $n$ possible states and $m$ possible actions, we have $m \phantom{\cdot} n \times n$ matrices for the possible transitions.
 The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
@@
 ```{r}
-soln_dist <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
+fifty_fifty <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
                                              model_prior = c(0.5, 0.5),
@@
 
+seventy_thirty <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
+                                             model_prior = c(0.7, 0.3),
+                                             max_iter = 1e4, epsilon = 1e-2)
+
+```
+
+```{r eval=FALSE}
+## Figure not shown -- intermediate prior gives an intermediate policy
 tibble(state = states,
-       action = actions[soln_dist$policy])  %>% 
-  ggplot(aes(state, action)) + geom_point()
+       action = actions[fifty_fifty$policy])  %>% 
+  ggplot(aes(state, action)) + geom_point() + 
+  geom_vline(aes(xintercept = states[x0]))
 ```
@@
 ```{r}
+
+## Rather stochastic in realization space, should consider ensembles instead...
+reps <- 10
+Tmax <- 60
 set.seed(123456)
-Tmax <- 50
-df <- mdp_planning(P_ghost, reward, discount, 
-                   policy = soln_dist$policy,
-                   x0 = x0, Tmax = Tmax)
+sim_ghost <- map_dfr(1:reps, function(i)
+                   mdp_planning(P_ghost, reward, discount, 
+                   policy = soln_ghost$policy,
+                   x0 = x0, Tmax = Tmax), .id = "reps")
+sim_bistable <- map_dfr(1:reps, function(i)
+                   mdp_planning(P_ghost, reward, discount, 
+                   policy = soln_bistable$policy,
+                   x0 = x0, Tmax = Tmax), .id = "reps")
+sim_fiftyfifty <-  map_dfr(1:reps, function(i)
+                   mdp_planning(P_ghost, reward, discount, 
+                   policy = fifty_fifty$policy,
+                   x0 = x0, Tmax = Tmax), .id = "reps")
+sim_seventythirty <- map_dfr(1:reps, function(i)
+                   mdp_planning(P_ghost, reward, discount, 
+                   policy = seventy_thirty$policy,
+                   x0 = x0, Tmax = Tmax), .id = "reps")
 
-df %>% 
-  mutate(state = states[state], action = actions[action] * 5) %>% 
+```
+
+```{r}
+ bind_rows(ghost = sim_ghost, 
+           bistable = sim_bistable, 
+           fifty_fifty = sim_fiftyfifty, 
+           seventy_thirty = sim_seventythirty, .id = "prior")  %>% 
+  mutate(state = states[state], action = actions[action]) %>% 
   select(-obs,-value) %>%
-  pivot_longer(-time) %>%
-  ggplot(aes(time, value, col=name)) + 
-  geom_point() + geom_path()
+  pivot_longer(c(-time, -prior, -reps)) %>%
+  ggplot(aes(time, value, group = interaction(name, reps), col=name)) + 
+  geom_path(alpha=0.5) + facet_wrap(~prior)
 ```
@@
 ```{r}
-set.seed(123456)
+set.seed(12345)
 Tmax <- 50
@@
 ```{r}
- learning_sim$df %>% 
+ learning$df %>% 
   select(-value) %>% 
@@
 ```{r}
-learning_sim$posterior %>% 
+model_names <- c("Ghost", "Bistable")
+learning$posterior %>% 
   data.frame(time = 1:Tmax) %>%
-  filter(time %in% c(1,50)) %>%
+  filter(time %in% floor(seq(1,10, length.out=5))) %>%
   gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = as.numeric(param)) %>% 
+  mutate(param = model_names[as.integer(param)]) %>% 
   ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
-  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
-   ylim(0,0.3)
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1])
+```
+
+```{r}
 ```

### 2019-12-30 manuscript/manuscript.Rmd
tidy up and WIP on manuscript

@@
 ---
-title: "Consequences of long transients in model inference and optimal control"
+title: "Ecological management of stochastic systems with long transients"
 date: "`r Sys.Date()`"
@@
     affiliation: ucb
-    footnote: Corresponding Author
+#    footnote: Corresponding Author
 address:
@@
 abstract: |
-  I explore the impact of long transient dynamics for model inference and optimal
+  Transient I explore the impact of long transient dynamics for model inference and optimal
 journal: Theoretical ecology
@@
 
-```{r figure1, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$.
+```{r fig_model, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$.
 Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.
-By changing parameter a, the system goes from bistable (red, a = 28.5) to a ghost attractor (blue, a = 27).
-For both curves, r = 0.8, K = 153, Q = 2, H = 20."}
-c(`ghost attractor` = 27,           
-  `bistable attractor` = 28.5) %>%
+By changing parameter $a$, the system goes from bistable (red, $a = 28$) to a ghost attractor (blue, $a = 27.2$).
+For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}
+c(`ghost attractor` = 27.2,           
+  `bistable attractor` = 28) %>%
   map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a), .id = "scenario") %>% 
@@
 
-Under appropriate choice of parameters, the model is bistable, with two stable equilibiria at non-zero population sizes -- a low, "endemic" state, and a much higher "outbreak" state.
+Under appropriate choice of parameters, the model is bistable, with a low, "endemic" stable state, and a much higher stable "outbreak" state.
 In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emergign from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses.
@@
 Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defences of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting avaliable tree habitat $K$ [@Harvey2014].
-
 Thus our simple dynamical model from \eqref{model} captures the qualitative processes and dynamics expected in the beetle outbreak system.
@@
 
-We also observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g.
+Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.
+Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g.
 the vertical axis, Figure 1) is still small in magnitude, and thus dynamics near this region move slowly.
@@
 
-Figure 2 illustrates several simulations from the model, starting from a low density, under both the scenario of a stable endemic state and a ghost attractor.
-Note that it can in general be quite difficult to distinguish dynamics of the ghost from the dynamics of the stable attractor.
+This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Figure 2.
+Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B).
+Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.
+Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastically escaping stable system.
 
@@
 P_ghost <- transition_matrix(states, actions, may(27.2), p$sigma)
-P_weak_attractor <- transition_matrix(states, actions, may(28), p$sigma)
 P_bistable <- transition_matrix(states, actions, may(28), p$sigma)
@@
 bistable_sim <- map_dfr(1:reps, function(i)
-  sim(P_weak_attractor, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
+  sim(P_bistable, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
   .id = "rep")
 
-bind_rows(ghost_sim, bistable_sim) %>%
+fig1_sims <- bind_rows(ghost_sim, bistable_sim)
+write_csv(fig1_sims, "manuscript_cache/fig1_sims.csv")
+```
+
+```{r fig_sims}
+fig1_sims %>%
   ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
@@
 ```
+
+This ambiguity is accentuated considerably if we are tasked with forecasting the future behavior having only observed the system during the early part of a trajectory (i.e.
+in the vicinity of the ghost attractor or lower stable point).
+Long transients under the influence of the ghost attractor can easily be mistaken for stable node dynamics, but will have considerably greater probability of suddenly transitioning to the outbreak state than trajectories under alternate stable state dynamics in the short term.
+
+While it is clear this situation -- in which a ghost attractor could be mistaken for a stable state -- can have important consequences for management, there has been little detailed investigation into precisely what those consequences are or how this uncertainty could be resolved.
+In the context of our bark beetle model, this corresponds to uncertainty as to whether the forest is healthy enough to contain the currently observed level of beetle activity to an endemic state, or whether the forest has been stressed sufficiently to be headed for a full-scale outbreak and thus warrant a more aggressive pest control response.
+
+In the remaining sections, I explore two different approaches to addressing the problem of management under a potential ghost attractor.
+The first considers the decision-theoretic optimal policy given uncertainty
 
@@
 Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.
-By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Marescot2013].
+By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985, @Marescot2013].
 
@@
 
-## Optimal Planning under transient dynamics vs stable state dynamics
+## Optimal planning under transient dynamics vs stable state dynamics
 
@@
 
-```{r}
+```{r mdp_solns}
 soln_ghost <- mdplearning::mdp_compute_policy(list(P_ghost), reward, discount, max_iter = 1e4, epsilon = 1e-2)
@@
 
+## Identical calculation using MDPtoolbox implementation
 #soln_ghost <-MDPtoolbox::mdp_value_iteration(P_ghost, reward, discount, max_iter = 5e3, epsilon = 1e-5)
@@
 ## With more noise, these are more similar.  With smaller noise, the ghost policy remains aggressive over a larger range
+policy_plot <- 
   tibble(state = states,
@@
                names_to="model",
-               values_to = "action") %>%
-  ggplot(aes(state,action, color = model)) + 
+               values_to = "action")
+
+write_csv(policy_plot, "manuscript_cache/policy_plot.csv")
+```
+
+```{r fig_policy}
+
+policy_plot %>%
+  ggplot(aes(state, action, color = model)) + 
   geom_point(alpha=0.6) + 
   scale_color_solarized()
+```
+
+```{r}
+### Consider other forecast models -- i.e. other transition functions
+
+## How does the identity matrix perform as a decision tool?  
+## Or, identity in f, but responsive in a?
 ```
@@
 
-```{r}
+```{r mdp_solns_mixed_prior}
 fifty_fifty <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
@@
 
-```{r}
+```{r planning_sims}
 
@@
 
-```
-
-```{r}
+sims_by_prior <- 
  bind_rows(ghost = sim_ghost, 
@@
            fifty_fifty = sim_fiftyfifty, 
-           seventy_thirty = sim_seventythirty, .id = "prior")  %>% 
+           seventy_thirty = sim_seventythirty, 
+           .id = "prior")
+
+sims_by_prior %>% write_csv("manuscript_cache/sims_by_prior.csv")
+
+```
+
+```{r fig_sims_by_prior}
+sims_by_prior <- read_csv("manuscript_cache/sims_by_prior.csv")
+
+sims_by_prior %>% 
   mutate(state = states[state], action = actions[action]) %>% 
@@
 
-```{r}
+```{r mdp_learning}
 set.seed(12345)
@@
 
+write_csv(learning$df, "manuscript_cache/learning_df.csv")
+write_csv(learning$posterior, "manuscript_cache/learning_posterior.csv")
 ```
 
-```{r}
+```{r fig_learning_sim}
  learning$df %>% 
@@
 
-```{r}
+```{r fig_learning_posterior}
 model_names <- c("Ghost", "Bistable")
@@
 
-```{r}
-```
-
-## Long transients when the underlying model is not known
-
-One of the more intuitive implications of the existence of long transients is the inherent risk in mistaking them for stable state dynamics.
-For example, Figure \ref{armina} illustrates how a transient
-
-### Stochastic dynamics and transients
-
-Stochastic dynamics can weaken the average influence of a ghost attractor, but long transients are still likely.
-Stochastic dynamics can also make weak attractors behave essentially as ghost attractors.
-
 # Methods
 
-Stochastic dynamic programming solutions were computed in R [@R] using value iteration using the package `MDPToolbox` [@Marescot2013; @MDPtoolbox].
-Code to reproduce this analysis is available at <https://github.com/cboettig/decisions-vs-transients>
+Stochastic dynamic programming solutions were computed in R [@R] using the value iteration stochastic dynamic programming algorithm [@Mangel1985; @Marescot2013].
+Generic R fucntions for the stochastic dynamic programming, including both planning over model uncertainty and adapative management for actively learning over model uncertainty, are supplied in the supplementary R package `mdplearning`, avaialable at <https://github.com/boettiger-lab/mdplearning>, v0.1.0 [@mdplearning].
+Code to reproduce the analysis, simulations and figures shown here can be found at <https://github.com/cboettig/decisions-vs-transients>.
 

### 2019-12-30 gh-pages.R

@@ -1,12 +1,0 @@
-library(fs)
-library(here)
-
-fs::dir_create(fs::path_rel("docs"))
-
-slides <- fs::path_rel("slides/solarized.html")
-#fs::link_create(slides, fs::path_rel("docs/index.html"))
-fs::file_copy(slides, fs::path_rel("docs/index.html"), overwrite = TRUE)
-fs::link_create(fs::path_rel("slides/img"), fs::path_rel("docs/img"))
-fs::link_create(fs::path_rel("slides/libs"), fs::path_rel("docs/libs"))
-fs::link_create(fs::path_rel("slides/solarized_files"), fs::path_rel("docs/solarized_files"))
-fs::link_create(fs::path_rel("slides/solarized-light.css"), fs::path_rel("docs/solarized-light.css"))

### 2019-12-30 notebook/nnetar.Rmd
neural net

---
author: "Carl Boettiger"
institute: "UC Berkeley"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(fable)
library(ggthemes)
pal <- solarized_pal()(6)

```

```{r}
n_s <- 121
states <- seq(0, 120, length = n_s)
actions <- seq(0, 120, length = n_s)

## Model constants -- used to compute transistion probabilities
efficiency <- .4    
p <- list(r = .8, K = 153, q = 2, b = 20, sigma = .05, x0 = 20) # fixed parameters

may <- function(a){  
  function(x, h = 0){ # May
    y <- x - efficiency * h
    pmax(
      ## controlling h is controlling the bifurcation point directly...
      y + y * p$r * (1 - y / p$K)  - a * y ^ p$q / (y ^ p$q + p$b ^ p$q),  
      0)
  }
}
```

```{r}
transition_matrix <- function(states, actions, f, sigma){
    n_s <- length(states)
    n_a <- length(actions)
    transition <- array(0, dim = c(n_s, n_s, n_a))
    for(i in 1:n_a){
      for (k in 1:n_s) {
          nextpop <- f(states[k], actions[i])
          if (nextpop <= 0) {
            x  <- c(1, rep(0, n_s - 1))
          } else {
            x <- truncnorm::dtruncnorm(states, 0, max(states), nextpop, sigma * nextpop)
            if(sum(x) <= 0){ 
              x  <- c(1, rep(0, n_s - 1))
            } else {
              x <- x / sum(x)
            }
          }
          transition[k, , i] <- x
      }
  }
  if(any(is.na(transition))) stop("error creating transition matrix")
transition
}
```

```{r}
## `actions` is an argument of indices, not necessarily the action values themselves 
sim <- function (transition,  x0, Tmax, action = rep(1, Tmax)){
    n_states <- dim(transition)[2]
    state <- numeric(Tmax + 1)
    state[1] <- x0
    time <- 1:Tmax
    for (t in time) {
        state[t + 1] <- base::sample(1:n_states, 
                               1, 
                               prob = transition[state[t], , action[t]])
    }
    data.frame(time = 1:Tmax, state = state[time])
}
```

```{r}
P_ghost <- transition_matrix(states, actions, may(27.2), p$sigma)
P_weak_attractor <- transition_matrix(states, actions, may(27.6), p$sigma)
P_bistable <- transition_matrix(states, actions, may(28), p$sigma)
```

```{r}
x0 <- which.min(abs(states - p$x0))
Tmax <- 200
set.seed(12345)
reps <- 50
ghost_sim <- map_dfr(1:reps, function(i)
  sim(P_ghost, x0, Tmax)  %>% mutate(state = states[state], scenario = "ghost"),
  .id = "rep")
weak <- map_dfr(1:reps, function(i)
  sim(P_weak_attractor, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
  .id = "rep")
bistable  <- map_dfr(1:reps, function(i)
  sim(P_bistable, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
  .id = "rep")
```

```{r}
#fig1_sims <- read_csv("../decisions-vs-transients/manuscript/manuscript_cache/fig1_sims.csv")
#bistable <- fig1_sims %>% filter(scenario == "bistable")
rep1 <- weak %>% filter(rep == 2, time <= 100) %>% select(time, state)
```

```{r}
rep1 %>%
  ggplot(aes(time, state)) + 
  geom_line(lwd=1.5, col = pal[1])

```

---

````{r}
horizon <- 200
marble <- rep1 %>%
    as_tsibble(index = time) %>%
  model(nn = NNETAR(state))
fable <- marble %>% 
  forecast(h = horizon, PI=TRUE) 
fc <- fable %>%
  mutate(sd = map_dbl(.distribution, function(x) sd(x[[1]][[1]])))

rep1 %>% filter(time <= 100) %>%
  ggplot(aes(time, state)) + 
  geom_line(lwd=1.5, col = pal[1]) + 
  geom_line(data = fc, aes(time, state), lwd=1.5) + 
  geom_ribbon(data = fc, aes(time, state, ymin = state - 2 *sd, ymax = state + 2 * sd ), 
              col=NA, alpha = 0.2) + 
  coord_cartesian(xlim= c(0,100+ horizon), ylim = c(0,60))

```

### 2019-12-30 manuscript/manuscript.Rmd
WIP writing writing

@@
 
+library(patchwork)
 library(ggplot2)
@@
 ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())
-pal <- solarized_pal()(3)
+pal <- solarized_pal()(7)
+scale_colour_discrete <- function(...) scale_colour_solarized()
+scale_fill_discrete <- function(...) scale_fill_solarized()
 ```
@@
 The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$ [@Boettiger2018].
-Model dynamics are illustrated in Figure 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.
+Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.
 
@@
 For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}
+
+## Fig 1
+
 c(`ghost attractor` = 27.2,           
@@
 Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g.
-the vertical axis, Figure 1) is still small in magnitude, and thus dynamics near this region move slowly.
+the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.
 This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.
@@
 
-This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Figure 2.
+This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Fig 2.
 Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B).
@@
 ```{r fig_sims}
+# Fig 2
 fig1_sims %>%
@@
 
-In the remaining sections, I explore two different approaches to addressing the problem of management under a potential ghost attractor.
-The first considers the decision-theoretic optimal policy given uncertainty
+In the remaining sections, we will explore two different approaches to addressing the problem of management under a potential ghost attractor.
+First, we will examine the decision-theoretic optimal policy for each stochastic model separately, illustrating that the transient system does indeed lead to a different management strategy than the alternate stable state system.
+We will then examine the optimal solution when the manager is uncertain which of the two models is the true.
+Finally, we will examine an active adaptive management approach, in which the manager updates the probability that each model is true following each iteration of decision and observation.
 
@@
 
+The transition matrix is an expression of conditional probability $P(X_{t+1} | X_t,  A_t)$, the probability that a system in state $X_t$ will transition to state $X_{t+1}$ in the next time interval if the manager takes action $A_t$.
+If we have $n$ possible states and $m$ possible actions, we have $m \phantom{\cdot} n \times n$ matrices for the possible transitions.
+The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
+For simplicity, the examples here have assumed a grid of state space values as he integers between 0 to 120, and an equivalently fine grid of 121 possible actions (control levels).
+This provides sufficient approximation to a continuous state space and action space, so that results are not sensitive to the details of this discritization.
+
 ```{r}
@@
 
-First, I will consider the case that in which the manager has been able to accurately infer the dynamics (i.e.
+First, let us consider the case that in which the manager has been able to accurately infer the dynamics (i.e.
 correctly estimate the parameters of Eq~\eqref{1}), and assuming the model structure postulated \eqref{1} is correct.
 The optimal strategy determined by SDP can be represented as a "policy" function $p(X)$, indicating for each possible state $X$ what the optimal action $\hat A = p(X)$ would be.
-Figure 3 displays the numerically computed optimal policy for the system with only a transient.
+Fig 3 displays the numerically computed optimal policy for the system with only a transient.
 Though the models differ by only a small amount in a single parameter, because one model corresponds to a system with only a ghost attractor while the other has a stable state attractor, the resulting optimal policy solutions differ significantly.
@@
 
-```{r fig_policy}
+```{r fig_policy, fig.cap="Optimal policy functions under each model.
+Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\delta = 0.999$, parameters for the transition models are as given in Fig 1.
+"}
+## Fig 3
 
@@
   ggplot(aes(state, action, color = model)) + 
-  geom_point(alpha=0.6) + 
-  scale_color_solarized()
+  geom_point(alpha=0.6)
+```
+
+```{r}
+reps <- 20
+Tmax <- 200
+set.seed(12345)
+sim_ghost <- map_dfr(1:reps, function(i)
+                   mdp_planning(P_ghost, reward, discount, 
+                   policy = soln_ghost$policy,
+                   x0 = x0, Tmax = Tmax), .id = "reps")
+sim_bistable <- map_dfr(1:reps, function(i)
+                   mdp_planning(P_ghost, reward, discount, 
+                   policy = soln_bistable$policy,
+                   x0 = x0, Tmax = Tmax), .id = "reps")
+
+fixed_sims <- 
+ bind_rows(ghost = sim_ghost, 
+           bistable = sim_bistable, 
+           .id = "prior") %>% 
+  mutate(state = states[state], 
+         action = actions[action])
+
+fixed_sims %>% write_csv("manuscript_cache/fixed_sims.csv")
+```
+
+We can get a better idea of the consequences of mistaking the long transient dynamics for stable state dynamics by examining the dynamics of a system when it is actively managed according to the policy function for the wrong model.
+It is important to note that even when the policy wrongly assumes the system has an endemic stable state (e.g.
+small beetle population density), this does not necessarily mean that the manager will never intervene and the dynamics will reach the high stable state unchecked.
+As we see in the policy function, at low densities, the alternative stable state policy does indeed choose to do nothing, since it expects the density to remain around an acceptable endemic level.
+However, once the densities get high enough (the system escapes the ghost attractor), the manager intervenes -- but this intervention is only enough to restore the system to the endemic state.
+This results in a pattern of perpetual interventions.
+Simulations of 10 independent realizations of a ghost-attractor system being managed with the optimal policy of the alternative stable state model (Fig 4, upper panel).
+In contrast, had the manager known the system no longer supported the endemic stable state but had only a ghost attractor, the optimal policy is a costly up-front intervention which eliminates the beetle entirely (Fig 4, lower panel).
+A comparison between the large up front cost of the optimally managed realizations to the ever rising costs of the mistaken case can be seen in Fig 5, shown in terms of net present value (i.e.
+cumulative costs, where future costs have been discounted by rate $\delta$.)  The examples shown here use a relatively small discount rate of 0.999 per unit time -- under sufficiently agressive discounting this large up-front investment would not be optimal.
+(Note that we use the term "small" meaning relative to the timescale shown, e.g.
+the value at time 200 is worth $0.999^{200} \approx 81.86\%$ of its starting value.
+As the units of time here correspond to the generation time of the beetle, this shouldn't be confused with an *annual* discount rate.)
+
+```{r fig_sims_by_prior}
+# Fig 4
+fixed_sims <- read_csv("manuscript_cache/fixed_sims.csv")
+
+fixed_sims %>% select(state, action, time, prior, reps) %>%
+  pivot_longer(c(-time, -prior, -reps), names_to = "line", values_to  = "state") %>%
+  filter(time < 100) %>%
+  ggplot(aes(time, state, group = interaction(line, reps), col=line)) + 
+  geom_line(alpha=0.5) + 
+  facet_wrap(~prior, ncol=1)
+```
+
+```{r}
+costs <- fixed_sims %>% group_by(time, prior)  %>% 
+  summarise(cost = mean(value)) %>%
+  group_by(prior) %>%
+  mutate(cost = abs(cumsum(cost * .999 ^ time))) %>%
+  rename(model = prior)
+
+costs %>%
+  ggplot(aes(time, cost, ymin = 0, ymax = cost, fill=model)) +
+  geom_ribbon(alpha = 0.5) +
+  ylab("Cost (net present value)") + xlab("time (days)")
+```
+
+```{r}
+# compare values
 ```
@@
 
-The transition matrix is an expression of conditional probability $P(X_{t+1} | X_t,  A_t)$, the probability that a system in state $X_t$ will transition to state $X_{t+1}$ in the next time interval if the manager takes action $A_t$.
-If we have $n$ possible states and $m$ possible actions, we have $m \phantom{\cdot} n \times n$ matrices for the possible transitions.
-The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
-We can represent uncertainty over the models and their parameters as uncertainties as conditional on the model, $M_i$, so that the transistion matrix must then be marganlized over the set of possible models: $$P(X_{t+1} | X_t,  A_t) = \sum_{i} P(X_{t+1} | X_t,  A_t, M_i) P(M_i)$$
+Rather than having to completely assume a given model, it is possible to reflect our uncertainty about which model is correct in our calculation of the optimal policy.
+We can represent uncertainty over both the models and their given parameters conditionally: so that the transistion matrix is marginalized over the set $M_i$ of possible models:
+
+$$P(X_{t+1} | X_t,  A_t) = \sum_{i} P(X_{t+1} | X_t,  A_t, M_i) P(M_i)$$
+
+This requires we can assign probabilities that each model is true, which I will refer to as 'prior probabilities' for reasons that will be clear in the next section.
+In this way, our uncertainty about the future state reflects both the inherent uncertainty of the stochastic process as well as our ignorance of the particular model.
+For simplicity I will consider only the two spcific models we have worked with so far, though this same approach could easily be extended to a suite of possible models.
+If the two models are considered equally likely, the optimal policy in the case of our examples looks like an only slightly more agressive version of the policy for the bistable system, as seen in the replicate simulations shown in Fig 6.
+This policy eliminates the beetles in a few of the replicate simulations where the population dips enough to make that decision economical, but otherwise settles for persistent management, as shown in the upper panel.
+However, we can also see that absolute certainty is not required for definitive action -- once we are sufficiently sure the system is a ghost attractor, e.g.
+70% probability as shown in the lower panel of Fig 6, the optimal solution eliminates the pest immediately on all replicates.
+Not surprisingly, the 50-50 case achieves a long run net present value that is intermediate to the optimal solution and the mistaken solution considered in Fig 5.
 
@@
 ```{r eval=FALSE}
-## Figure not shown -- intermediate prior gives an intermediate policy
+## Fig not shown -- intermediate prior gives an intermediate policy
 tibble(state = states,
@@
 
-## Rather stochastic in realization space, should consider ensembles instead...
-reps <- 10
-Tmax <- 60
-set.seed(123456)
-sim_ghost <- map_dfr(1:reps, function(i)
-                   mdp_planning(P_ghost, reward, discount, 
-                   policy = soln_ghost$policy,
-                   x0 = x0, Tmax = Tmax), .id = "reps")
-sim_bistable <- map_dfr(1:reps, function(i)
-                   mdp_planning(P_ghost, reward, discount, 
-                   policy = soln_bistable$policy,
-                   x0 = x0, Tmax = Tmax), .id = "reps")
 sim_fiftyfifty <-  map_dfr(1:reps, function(i)
@@
 sims_by_prior <- 
- bind_rows(ghost = sim_ghost, 
-           bistable = sim_bistable, 
-           fifty_fifty = sim_fiftyfifty, 
+ bind_rows(fifty_fifty = sim_fiftyfifty, 
            seventy_thirty = sim_seventythirty, 
-           .id = "prior")
+           .id = "prior")  %>% 
+  mutate(state = states[state], 
+         action = actions[action])
 
@@
 ```{r fig_sims_by_prior}
+## Fig 6
+
 sims_by_prior <- read_csv("manuscript_cache/sims_by_prior.csv")
 
-sims_by_prior %>% 
-  mutate(state = states[state], action = actions[action]) %>% 
-  select(-obs,-value) %>%
-  pivot_longer(c(-time, -prior, -reps)) %>%
-  ggplot(aes(time, value, group = interaction(name, reps), col=name)) + 
-  geom_path(alpha=0.5) + facet_wrap(~prior)
+sims_by_prior  %>% select(state, action, time, prior, reps) %>%
+  pivot_longer(c(-time, -prior, -reps), names_to = "line", values_to  = "state") %>%
+  filter(time < 100) %>%
+  ggplot(aes(time, state, group = interaction(line, reps), col=line)) + 
+  geom_line(alpha=0.5) + 
+  facet_wrap(~prior, ncol=1)
 ```
@@
 
-This probability thus depends on our prior belief that model $i$ is true, $P(M_i)$.
-Not only does this allow our manager to make her decision in to reflect uncertainty over models, but it also suggests that having made that decision and observed the outcomes, the manager can update her belief over possible models according to Bayes Rule,
+In the scenarios we have just considered, the manager continues to apply the same optimal decision rule over the entire management period.
+This does not, of course, mean that the management response is completely static -- in fact, the manager selects a different the intervention level at each timestep in response to updated information about the current beetle population size -- but the manager does not use that information to revisit the underlying assumptions about the appropriate model.
+However, each additional observation potentially provides further information about which model is correct.
+After enough attempts to restore the population to a stable endemic state only to have another outbreak follow, we might expect our manager to realize that the endemic state just isn't as stable as originally thought.
+More formally, we might assume the manager would update that initial prior belief over the possible models according to Bayes Rule,
 
 $$P(M_i | X_{t+1} ) = \frac{P(X_{t+1} | M_i ) P(M_i)}{P(X_{t+1})}$$
-that is, the optimal manager not only plans over the uncertainty, but also learns which model is most likely to be correct over the course of management.
+In other words, our optimal manager not only plans over the uncertainty, but also learns which model is most likely to be correct over the course of management.
+This approach corresponds to Carl Walters' original notion of "adaptive management" [@Walters1978].
+(Note that the term is now used much more loosely, and would commonly include the kind of decision rule shown previously where the decision is adjusted based on future observations.).
+This approach is considerably more computationally intensive, since the stochastic dynamic programming problem must be solved again at each time step after the model weights have been updated.
+Fig 7 illustrates a typical outcome of adaptive management starting with the uniform prior (equal weights on both models): a short period of modest intervention while the manager learns which model is more likely to be correct, followed by a more decisive intervention once the long transient becomes sufficiently likely.
+Of course, this defensive action also eliminates the potential for future learning, so the manager can never be completely sure this action was necessary, as shown by the model probabilities in Fig 8.
+This represents a form of precautionary principle already suggested by the purely planning approach considered in the previous section -- the manager need not be certain about the correct model, but merely reach a certain threshold of probability.
 
 ```{r mdp_learning}
+# Fig 7
 set.seed(12345)
@@
 ```{r fig_learning_sim}
+# Fig 8
  learning$df %>% 
@@
 
-# Methods
+# Discussion
+
+Long transients. Equilibrium.
+
+Note that this approach can equally well consider an arbitrary suite of models (i.e. varying the parameter $a$ over a range of values, as well as to models with completely different structure).
+
+Not considered:
+
+- partially observed optimal control
+- non-stationary dynamics e.g. in which
+
+Implications:
+
+- Long transients can be misleading -- e.g. mistaking a ghost attractor for a stable state.
+- These mistakes can be significant, since long transients require qualitatively different management strategies.
+- Nevertheless, thes
+
+# Numerical Methods
 

### 2019-12-31 manuscript/manuscript.Rmd
WIP

@@
 
-```{r fig_model, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$.
+```{r figure1, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$.
 Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.
@@
 
-```{r fig_sims}
+```{r figure2}
 # Fig 2
@@
 
-```{r fig_policy, fig.cap="Optimal policy functions under each model.
-Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\delta = 0.999$, parameters for the transition models are as given in Fig 1.
-"}
+```{r figure3, fig.cap="Optimal policy functions under each model.
+Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\\delta = 0.999$, parameters for the transition models are as given in Fig 1."}
 ## Fig 3
@@
 
-```{r fig_sims_by_prior}
+```{r figure4}
 # Fig 4
@@
 
-```{r}
+```{r figure5}
 costs <- fixed_sims %>% group_by(time, prior)  %>% 
@@
 
-```{r fig_sims_by_prior}
+```{r figure6}
 ## Fig 6
@@
 ```{r mdp_learning}
-# Fig 7
 set.seed(12345)
@@
 
-```{r fig_learning_sim}
-# Fig 8
+```{r figure7}
+# Fig 7
  learning$df %>% 
@@
 
-```{r fig_learning_posterior}
+```{r figure8}
 model_names <- c("Ghost", "Bistable")
@@
 - These mistakes can be significant, since long transients require qualitatively different management strategies.
-- Nevertheless, thes
+- Decision-theoretic approaches provide a way to manage this uncertainty
+
+Markov Decision Processes are an active area of research in robotics and artificial intelligence community, though terminology differs substantially.
 

### 2020-01-03 manuscript/manuscript.Rmd
WIP

@@
 
-Long transients. Equilibrium.
+While the importance of long transients to ecological dynamics is now being recognized, greater understanding of the consequences this realization has for ecological management is still needed.
+For instance, @Hastings2018 discusses the potential implications that long transients have the management of ecosystems, including the situation we have considered here:
 
-Note that this approach can equally well consider an arbitrary suite of models (i.e. varying the parameter $a$ over a range of values, as well as to models with completely different structure).
+> In other cases, when it is difficult to distinguish whether observed dynamics are transient or at equilibrium, models of both possibilities can be developed to test the sensitivity of proposed management strategies to the model assumptions,
 
-Not considered:
+while also calling for closer study of those consequences:
 
-- partially observed optimal control
-- non-stationary dynamics e.g. in which
+> A full treatment of the management consequences and opportunities presented by long transients re- quires further attention
 
-Implications:
+In this analysis, I have attempted to provide a more detailed treatment of the management consequences of just such a long transient by focusing on a minimal model capable of producing ghost-attractor dynamics.
+This model captures common ecological management scenarios, such as the outbreak of a bark beetle pest we have focused on here, and could easily be generalized to other typical ecological management problems which also involve tipping-point dynamics.
+As the name implies, ecological *management* is typically not a single decision-point, but rather, a *sequential decision process*, where a manager implements a policy over time and may adjust that policy in response to future observations.
+I have drawn on methods from optimal control theory, stochastic dynamic programming in particular, to explore how a manager responds to the possible outbreak of a bark beetle infestation when there is uncertainty as to whether the observed dynamics are a transient or equilbrium phase.
 
-- Long transients can be misleading -- e.g. mistaking a ghost attractor for a stable state.
-- These mistakes can be significant, since long transients require qualitatively different management strategies.
-- Decision-theoretic approaches provide a way to manage this uncertainty
+From this analysis, several conclusions emerge.
+First, we have a clear illustration of just how misleading long transient dynamics can appear -- without access to multiple replicate realizations (which is nearly always the case in real systems), distinguishing between transient dynamics and stable dynamics in the same system from purely observational data will always be difficult.
+(A more formal consideration of this statistical inference problem is treated in Reimer et al, intended for this same issue of Theoretical Ecology).
+Second, we see that this miss-identification leads to qualitatively different management outcomes, even when management is dynamic and following an optimal policy for a closely related model.
+In particular, in our example, a manager accumulats an ever-growing cost of coping with the outbreak while never fully eliminating the threat.
 
-Markov Decision Processes are an active area of research in robotics and artificial intelligence community, though terminology differs substantially.
+Despite these differences, we have also seen that dynamic and adaptive management approaches can nevertheless be an effective mechanism to cope with the uncertainty posed by long transients.
+First, it is important to note that all dynamic management scenarios succeed in preventing the outbreak even without the correct model.
+Because an outbreak is still possible under the stable state scenario due to stochastic tipping, the optimal manager still responds to high beetle densities.
+This highlights the importance of properly understanding the management context and the decision space in order to understand management implications -- management is a socio-ecological problem and can never be understood from the ecological dimension alone.
+
+Second, we observe that a manager can build the uncertainty over the two scenarios into the decision process.
+Once there is sufficient probability that the system is in a long transient sustained by a ghost attractor rather than in a true equilibrium, the manager reaches the identical optimal policy as would a manager with perfect knowledge of which model is correct.
+Perfect information is not required for perfect management.
+We see a second illustration of this fact when we consider an adaptive management approach, which attempts to learn the true model over the course of management.
+Adaptive management does indeed allow the manager to correctly adjust probabilities each model is true.
+However, learning does not continue indefinitely, since once the probability of a ghost attractor becomes high enough, the manager eliminates the beetles and with it, any potential for further learning.
+
+Several caveats about this analysis must be noted.
+First, I underscore that this is not intended to be a precise model of bark beetle dynamics, but rather a general model  [sensu @Levins1966] to explore the management consequences of confusing a ghost attractor for an alternative stable state.
+A precise model would depend on ecological details of forest and beetle dynamics (including spatially explicit dynamics), as well as the details of management interventions and associated costs.
+Second, for simplicity, we have assumed that while our manager does not know whether the system has multiple equilibria or a ghost attractor, the manager does know most other details about the system.
+Importantly, we have assumed the manager observes the current population size without error before each action.
+In reality, any such observation would be accompanied by some uncertainty, making the decision problem into a Partially Observed Markov Decision Process (POMDP), increasing the complexity dramatically [e.g.
+@pomdp-intro].
+I have also assumed the manager has knowledge of the general model structure.
+This assumption can be relaxed in much the same way in which introduced model uncertainty, allowing one to consider an arbitrary suite of models (i.e.
+varying the parameter $a$ over a range of values, as well as to models with completely different structure).
+However, the curse of dimensionality imposes obvious limitations on this.
+Moreover, the approach assumes that while the ecological dynamics can be transient, the environment in which they occur is stationary (i.e.
+model parameters are constant.)  It is worth noting that these all of these issues in Markov Decision Processes are active areas of research in the robotics and artificial intelligence community, and recent advances in reinforcement learning could provide a promising way forward for many ecological management problems as well [e.g.
+@pnas].
 

### 2020-01-03 manuscript/manuscript.Rmd
a few captions

@@
 knitr::opts_chunk$set(echo = FALSE, cache = 1,  
-                      dev = "cairo_pdf", fig.width=4.5, 
-                      fig.height=3.5, message = FALSE,
+                      dev = "cairo_pdf", 
+                      fig.width=4.5, 
+                      fig.height=3.5, 
+                      fig.align = "center",
+                      message = FALSE,
                       warning = FALSE)
@@
 
-```{r figure2}
+```{r figure2, fig.cap="50 replicate simulations with identical starting conditions ($X_0 = 21$) of each scenario with parameter values as given in Fig 1.
+While the ensemble of 50 replicates clearly appears more stable than the ghost attractor, individual replicates may be difficult to distingush."}
 # Fig 2
 fig1_sims %>%
-  ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
-  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario)
+  ggplot(aes(time, state, group = interaction(rep, scenario), col = scenario)) + 
+  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario, ncol = 1)
 ```
@@
 
-```{r figure4}
+```{r figure4, fig.cap = "Simulations of dynamic management"}
 # Fig 4
@@
 model parameters are constant.)  It is worth noting that these all of these issues in Markov Decision Processes are active areas of research in the robotics and artificial intelligence community, and recent advances in reinforcement learning could provide a promising way forward for many ecological management problems as well [e.g.
-@pnas].
+@pnas; or see @deep-rl for an overview].
 

### 2020-01-04 manuscript/manuscript.Rmd
abstract, fig caps

@@
 abstract: |
-  Transient I explore the impact of long transient dynamics for model inference and optimal
+  Long transients may be common in ecological dynamics, but the implications such dynamics have for ecological management have not been fully explored.
+Long transient periods can easily be mistaken for stable-state dynamics, but may require dramatically different management policies.
+Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: a recently recognized mechanism which can give rise to long transients.
+I consider three approaches of increasing sophistication: (1) dynamic management under a fixed model, (2) management that accounts for uncertainty over possible models, and (3) adaptive management that actively learns the correct model over the management process.
+This analysis confirms the prediction that long transients can create considerable uncertainty and give rise to very different optimal management policies, but also illustrates that dynamic management that can either plan for this uncertainty or actively learn to decrease the uncertainty can promote successful management of long transients.
 journal: Theoretical ecology
@@
 output: rticles::elsevier_article
+preamble: |
+  \journal{Theoretical Ecology}
+  \usepackage[left]{lineno}
 
@@
 The potential for this mistake is straightforward to demonstrate even under idealized conditions such as access to accurate, high frequency data and knowledge of the correct model structure.
-Decision-theoretic approaches to ecological management provide a more nuianced approach for dealing with uncertainty by explicitly considering the set of all actions available to the manager and the likely consequences of those actions over time.
+Decision-theoretic approaches to ecological management provide a more nuanced approach for dealing with uncertainty by explicitly considering the set of all actions available to the manager and the likely consequences of those actions over time.
 In this paper, I explore the ability of both passive and active adaptive management approaches to cope with the challenge of long transient dynamics.
@@
 I consider a generic model of population dynamics which can exhibit phenomena such as bistability and a ghost attractor dynamics.
-This model is amenable to interpretation in many ecological contexts (sensu @Levins1966 sacrificing precision to realism and generality), such as the collapse of a fish population, recovery of an endagered species, or the outbreak a pest.
-To provide a more tangible context to frame our questions of ecological mangement, I will focus on the latter scenario of a pest outbreak.
+This model is amenable to interpretation in many ecological contexts (sensu @Levins1966 sacrificing precision to realism and generality), such as the collapse of a fish population, recovery of an endangered species, or the outbreak a pest.
+To provide a more tangible context to frame our questions of ecological management, I will focus on the latter scenario of a pest outbreak.
 In this scenario, we imagine a species such as the Pine bark beetle, (e.g.
@@
 The growth  rate of the species is determined by an intrinsic growth rate $r$ and carrying capacity $K$.
-The species is under mortality pressure from another species (the natural defences of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$.
-$Q$ determines the steepness of the saturating response.
-The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$ [@Boettiger2018].
+The species is under mortality pressure from another species (the natural defenses of trees, in our context) which has an S curve saturation which reaches its half maximum at $H$ and maximum rate of $a$.
+Parameter $Q$ determines the steepness of the saturating response.
+The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$, uncorrelated Gaussian noise $\xi \sim \mathcal{N}(0,\sigma)$ with mean zero and standard deviation $\sigma$ [@Boettiger2018].
 Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.
@@
 Under appropriate choice of parameters, the model is bistable, with a low, "endemic" stable state, and a much higher stable "outbreak" state.
-In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emergign from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses.
-This saturating response and the resulting threshhold dynamics in the response of affected trees exhibited by this model is quite well motivated   [e.g.@Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate.
-Initially this results in the strengthening (concave up) portion of the S curve as other defences are induced by the attack.
-However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defences, leading to a saturating response.
+In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emerging from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses.
+This saturating response and the resulting threshold dynamics in the response of affected trees exhibited by this model is quite well motivated   [e.g.
+@Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate.
+Initially this results in the strengthening (concave up) portion of the S curve as other defenses are induced by the attack.
+However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defenses, leading to a saturating response.
 Low beetle densities are thus stable, kept in check by the increasingly strong response of healthy trees.
-Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defences of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting avaliable tree habitat $K$ [@Harvey2014].
-Thus our simple dynamical model from \eqref{model} captures the qualitative processes and dynamics expected in the beetle outbreak system.
-Bear in mind that this model is not intended to be a precise, quantitive match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.
+Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defenses of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting available tree habitat $K$ [@Harvey2014].
+Thus our simple dynamical model from Eq 1 captures the qualitative processes and dynamics expected in the beetle outbreak system.
+Bear in mind that this model is not intended to be a precise, quantitative match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.
 
@@
 Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.
-Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastically escaping stable system.
+Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastic escape from a stable node.
 
@@
 If we have $n$ possible states and $m$ possible actions, we have $m \phantom{\cdot} n \times n$ matrices for the possible transitions.
-The transistion matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
+The transition matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
 For simplicity, the examples here have assumed a grid of state space values as he integers between 0 to 120, and an equivalently fine grid of 121 possible actions (control levels).
@@
 First, let us consider the case that in which the manager has been able to accurately infer the dynamics (i.e.
-correctly estimate the parameters of Eq~\eqref{1}), and assuming the model structure postulated \eqref{1} is correct.
+correctly estimate the parameters of Eq \eqref{model}, and assuming the model structure postulated \eqref{model} is correct.
 The optimal strategy determined by SDP can be represented as a "policy" function $p(X)$, indicating for each possible state $X$ what the optimal action $\hat A = p(X)$ would be.
@@
 Simulations of 10 independent realizations of a ghost-attractor system being managed with the optimal policy of the alternative stable state model (Fig 4, upper panel).
-In contrast, had the manager known the system no longer supported the endemic stable state but had only a ghost attractor, the optimal policy is a costly up-front intervention which eliminates the beetle entirely (Fig 4, lower panel).
-A comparison between the large up front cost of the optimally managed realizations to the ever rising costs of the mistaken case can be seen in Fig 5, shown in terms of net present value (i.e.
-cumulative costs, where future costs have been discounted by rate $\delta$.)  The examples shown here use a relatively small discount rate of 0.999 per unit time -- under sufficiently agressive discounting this large up-front investment would not be optimal.
-(Note that we use the term "small" meaning relative to the timescale shown, e.g.
-the value at time 200 is worth $0.999^{200} \approx 81.86\%$ of its starting value.
-As the units of time here correspond to the generation time of the beetle, this shouldn't be confused with an *annual* discount rate.)
 
-```{r figure4, fig.cap = "Simulations of dynamic management"}
+```{r figure4, fig.cap = "Simulations of dynamic management under each of the policy functions shown in Fig 3, when the true model is in fact the ghost attractor.
+We have used a common scale for both the state (beetle density) and action (control intensity), which range from 0 to 120, allowing us to display these on a common y-axis to reduce the number of panels.
+In general states and actions will be in different units.
+The manager who has incorrectly assumed the system is bistable allows the beetle to persist, but must take constant management action to maintain the population at the acceptable endemic size.
+The manager who knows the dynamics are merely a long transient due to a ghost attractor immediately takes much more decisive action (compare costs, Fig 5) eliminate the beetle."}
 # Fig 4
@@
   geom_line(alpha=0.5) + 
-  facet_wrap(~prior, ncol=1)
+  facet_wrap(~prior, ncol=1) + ylab("state or action")
 ```
 
-```{r figure5}
+  In contrast, had the manager known the system no longer supported the endemic stable state but had only a ghost attractor, the optimal policy is a costly up-front intervention which eliminates the beetle entirely (Fig 4, lower panel).
+A comparison between the large up front cost of the optimally managed realizations to the ever rising costs of the mistaken case can be seen in Fig 5, shown in terms of net present value (i.e.
+cumulative costs, where future costs have been discounted by rate $\delta$.)  The examples shown here use a relatively small discount rate of 0.999 per unit time -- under sufficiently aggressive discounting this large up-front investment would not be optimal.
+(Note that we use the term "small" meaning relative to the timescale shown, e.g.
+the value at time 200 is worth $0.999^{200} \approx 81.86\%$ of its starting value.
+As the units of time here correspond to the generation time of the beetle.)
+
+```{r figure5, fig.cap="Average costs associated with control of the beetle over time under across the simulations shown in Fig 4.
+Costs are in net present value: cumulative cost of management (in arbitrary units), discounting future values by the discount rate $\\delta$.
+Observe that optimal management of the ghost attractor results in a large up-front investment in control, while incorrectly assuming the system is bistable results in control that is initially less costly but continues to accumulate over time."}
 costs <- fixed_sims %>% group_by(time, prior)  %>% 
@@
   geom_ribbon(alpha = 0.5) +
-  ylab("Cost (net present value)") + xlab("time (days)")
-```
-
-```{r}
-# compare values
-```
-
-```{r}
-### Consider other forecast models -- i.e. other transition functions
-
-## How does the identity matrix perform as a decision tool?  
-## Or, identity in f, but responsive in a?
+  ylab("Cost (net present value)") + xlab("time")
 ```
@@
 Rather than having to completely assume a given model, it is possible to reflect our uncertainty about which model is correct in our calculation of the optimal policy.
-We can represent uncertainty over both the models and their given parameters conditionally: so that the transistion matrix is marginalized over the set $M_i$ of possible models:
+We can represent uncertainty over both the models and their given parameters conditionally: so that the transition matrix is marginalized over the set $M_i$ of possible models:
 
@@
 In this way, our uncertainty about the future state reflects both the inherent uncertainty of the stochastic process as well as our ignorance of the particular model.
-For simplicity I will consider only the two spcific models we have worked with so far, though this same approach could easily be extended to a suite of possible models.
-If the two models are considered equally likely, the optimal policy in the case of our examples looks like an only slightly more agressive version of the policy for the bistable system, as seen in the replicate simulations shown in Fig 6.
+For simplicity I will consider only the two specific models we have worked with so far, though this same approach could easily be extended to a suite of possible models.
+If the two models are considered equally likely, the optimal policy in the case of our examples looks like an only slightly more aggressive version of the policy for the bistable system, as seen in the replicate simulations shown in Fig 6.
 This policy eliminates the beetles in a few of the replicate simulations where the population dips enough to make that decision economical, but otherwise settles for persistent management, as shown in the upper panel.
@@
 
-```{r figure6}
+```{r figure6, fig.cap="Simulations of scenarios which reflect manager's uncertainty about the true model.
+Upper panel: manager assumes both the ghost attractor and the bistable attractor models are equally probable.
+Some of the 20 replicate simulations result in large actions that eliminate the pest, while most resemble management under the assumption of bistability.
+Lower panel: manager assumes a 70\\% probability of a ghost attractor.
+All 20 replicate simulations immediately opt to eliminate the beetle immediately in this scenario.
+Note that as in Fig 4, y axis is used as a common scale for state and action."}
 ## Fig 6
@@
 In the scenarios we have just considered, the manager continues to apply the same optimal decision rule over the entire management period.
-This does not, of course, mean that the management response is completely static -- in fact, the manager selects a different the intervention level at each timestep in response to updated information about the current beetle population size -- but the manager does not use that information to revisit the underlying assumptions about the appropriate model.
+This does not, of course, mean that the management response is completely static -- in fact, the manager selects a different the intervention level at each time-step in response to updated information about the current beetle population size -- but the manager does not use that information to revisit the underlying assumptions about the appropriate model.
 However, each additional observation potentially provides further information about which model is correct.
@@
 
-```{r figure7}
+```{r figure7, fig.cap="Simulated adaptive management of the forest-beetle system.
+Initally each model is considered equally likely, and management aims at retaining a stable endemic population.
+The manager learns over time (Fig 8), and decides to take more definitive action to eliminate the beetle pest."}
 # Fig 7
  learning$df %>% 
-  select(-value) %>% 
+  select(-value, -obs) %>% 
   gather(series, state, -time) %>% 
-  ggplot(aes(time, states[state], color = series)) + geom_line()
+  ggplot(aes(time, states[state], color = series)) + geom_line() + ylab("state or action")
 ```
 
-```{r figure8}
+```{r figure8, fig.cap="Learning through Adaptive Management of the forest-beetle system.
+Bars show the probabilities that each model is correct over the first 10 steps of the simulated management process shown in Fig 7.
+Initial probabilities for each model are equal, and learning proceeds over the first 5 timesteps (shown in lighter shades in decreasing probability of the bistable model beign correct).
+By the fifth iteration, the manager has opted to eliminate the beetles and further learning is no longer possible in this system."}
 model_names <- c("Ghost", "Bistable")
@@
 As the name implies, ecological *management* is typically not a single decision-point, but rather, a *sequential decision process*, where a manager implements a policy over time and may adjust that policy in response to future observations.
-I have drawn on methods from optimal control theory, stochastic dynamic programming in particular, to explore how a manager responds to the possible outbreak of a bark beetle infestation when there is uncertainty as to whether the observed dynamics are a transient or equilbrium phase.
+I have drawn on methods from optimal control theory, stochastic dynamic programming in particular, to explore how a manager responds to the possible outbreak of a bark beetle infestation when there is uncertainty as to whether the observed dynamics are a transient or equilibrium phase.
 
@@
 First, we have a clear illustration of just how misleading long transient dynamics can appear -- without access to multiple replicate realizations (which is nearly always the case in real systems), distinguishing between transient dynamics and stable dynamics in the same system from purely observational data will always be difficult.
-(A more formal consideration of this statistical inference problem is treated in Reimer et al, intended for this same issue of Theoretical Ecology).
+(A more formal consideration of this statistical inference problem is treated in Reimer et al., intended for this same issue of Theoretical Ecology).
 Second, we see that this miss-identification leads to qualitatively different management outcomes, even when management is dynamic and following an optimal policy for a closely related model.
-In particular, in our example, a manager accumulats an ever-growing cost of coping with the outbreak while never fully eliminating the threat.
+In particular, in our example, a manager accumulates an ever-growing cost of coping with the outbreak while never fully eliminating the threat.
 
@@
 A precise model would depend on ecological details of forest and beetle dynamics (including spatially explicit dynamics), as well as the details of management interventions and associated costs.
-Second, for simplicity, we have assumed that while our manager does not know whether the system has multiple equilibria or a ghost attractor, the manager does know most other details about the system.
-Importantly, we have assumed the manager observes the current population size without error before each action.
+Parameter values have been chosen to be illustrative of the region of interest where a ghost attractor could be mistaken for a weakly bistable system.
+Extreme values of the parameters will alter the qualitative patterns seen here.
+For instance, very large environmental noise intensity ($\sigma$), erode any differences between the ghost and the weakly bistable system, since escapes from the low density, endemic state become commonplace.
+Likewise, extreme values of the costs of control or costs of tree damage can also make the decision problem trivial, where intervention is never worthwhile or always worthwhile.
+The examples here focus on the more ecologically typical region where management is not trivial.
+
+Another caveat regards the inevitable simplifying assumptions we have introduced to maintain a tractable system.
+We have assumed that while our manager does not know whether the system has multiple equilibria or a ghost attractor, the manager does know most other details about the system.
+In particular, we have assumed the manager observes the current population size without error before each action.
 In reality, any such observation would be accompanied by some uncertainty, making the decision problem into a Partially Observed Markov Decision Process (POMDP), increasing the complexity dramatically [e.g.
 @pomdp-intro].
-I have also assumed the manager has knowledge of the general model structure.
+We have also assumed the manager has knowledge of the general model structure.
 This assumption can be relaxed in much the same way in which introduced model uncertainty, allowing one to consider an arbitrary suite of models (i.e.
@@
 Stochastic dynamic programming solutions were computed in R [@R] using the value iteration stochastic dynamic programming algorithm [@Mangel1985; @Marescot2013].
-Generic R fucntions for the stochastic dynamic programming, including both planning over model uncertainty and adapative management for actively learning over model uncertainty, are supplied in the supplementary R package `mdplearning`, avaialable at <https://github.com/boettiger-lab/mdplearning>, v0.1.0 [@mdplearning].
+Generic R functions for the stochastic dynamic programming, including both planning over model uncertainty and adaptive management for actively learning over model uncertainty, are supplied in the supplementary R package `mdplearning`, available at <https://github.com/boettiger-lab/mdplearning>, v0.1.0 [@mdplearning].
 Code to reproduce the analysis, simulations and figures shown here can be found at <https://github.com/cboettig/decisions-vs-transients>.

### 2020-01-04 docs/pdf-slides.R
ESA slides now in separate repo

@@ -1,10 +1,0 @@
-#pagedown::chrome_print(here::here("slides/boettiger.Rmd"), timeout=3000)
-
-
-
-library(webshot)
-install_phantomjs()
-file_name <- paste0("file://", here::here("slides/boettiger.Rmd"))
-webshot(file_name, "slides.pdf")
-
-

### 2020-01-04 manuscript/manuscript.Rmd
add line numbers
repo organization, structure

@@
   \journal{Theoretical Ecology}
-  \usepackage[left]{lineno}
+  \usepackage{lineno}
+  \linenumbers
 
@@
   ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
-  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1])
+  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
+  xlab("Model")
 ```
@@
 
+# Acknowledgements
+
+This work was inspired and influenced many discussions at the 2019 NIMBioS workshop on transient dynamics in ecology, and the 2019 organized oral session on ecological transients.
+CB also acknowledges computational resources from NSF's XSEDE Jetstream (DEB160003) and Chameleon cloud platforms, as well as the support from UC Berkeley and the USDA Hatch project CA-B-INS-0162-H.
+
 \pagebreak

### 2020-01-04 manuscript/debug.R

@@ -1,46 +1,0 @@
-type <- "policy iteration"
-n_models <- length(transition)
-n_states <- dim(transition[[1]])[1]
-n_actions <- dim(transition[[1]])[3]
-next_value <- numeric(n_states)
-next_policy <- numeric(n_states)
-V_model <- array(dim = c(n_states, n_models))
-converged <- FALSE
-t <- 1
-if (is.null(model_prior)) {
-  model_prior <- rep(1, n_models)/n_models
-}
-if (type == "finite time")
-  max_iter <- Tmax
-while (t < max_iter && converged == FALSE) {
-  Q <- array(0, dim = c(n_states, n_actions))
-  for (i in 1:n_actions) {
-    for (j in 1:n_models) {
-      V_model[, j] <- transition[[j]][, , i] %*% next_value
-    }
-    Q[, i] <- reward[, i] + discount * V_model %*% model_prior
-    if(any(is.nan(Q))){
-      stop(paste(V_model[, j]))
-      #stop(sprintf("t=%i, i=%i,j=%i", t, i, j))
-      }
-    if(any(is.na(Q))) stop("NAs found")
-
-  }
-  value <- apply(Q, 1, max)
-  policy <- apply(Q, 1, which.max)
-  if (type == "value iteration") {
-    if (sum(abs(value - next_value)) < epsilon) {
-      converged <- TRUE
-    }
-  }
-  else if (type == "policy iteration") {
-    if (sum(abs(policy - next_policy)) < epsilon) {
-      converged <- TRUE
-    }
-  }
-  next_value <- value
-  next_policy <- policy
-  t <- t + 1
-  if (t == max_iter)
-    message("Note: max number of iterations reached")
-}

### 2020-01-04 manuscript/manuscript.Rmd
remote dependencies

@@
 output: rticles::elsevier_article
+keywords: 
+  - transients, 
+  - optimal control, 
+  - adaptive management, 
+  - stochasticity, 
+  - uncertainty, 
+  - ecological management
 preamble: |

### 2020-06-22 manuscript/manuscript.Rmd
EM adds line numbers already
revisions WIP

@@
 Long transient periods can easily be mistaken for stable-state dynamics, but may require dramatically different management policies.
-Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: a recently recognized mechanism which can give rise to long transients.
+Here, I explore the optimal management of stochastic ecological systems that may contain either a tipping point or a ghost attractor: an important mechanism which can give rise to long transients.
 I consider three approaches of increasing sophistication: (1) dynamic management under a fixed model, (2) management that accounts for uncertainty over possible models, and (3) adaptive management that actively learns the correct model over the management process.
@@
   \journal{Theoretical Ecology}
-  \usepackage{lineno}
-  \linenumbers
+
+## EditorialManager adds its own linenumbers to pdf already
+#\usepackage{lineno}
+# \linenumbers
 
@@
 *Dendroctonus ponderosae*, @Harvey2014) whose recent boom in population has devastated much of Northern America's pine forests.
-The model is a stochastic version of the consumer-resource model of @May1977,
+The model is a stochastic extension of the deterministic consumer-resource model of @May1977,
 
@@
 Parameter $Q$ determines the steepness of the saturating response.
-The dynamics are subject to environmental white noise proportional to the population size, $\xi_t X_t$, uncorrelated Gaussian noise $\xi \sim \mathcal{N}(0,\sigma)$ with mean zero and standard deviation $\sigma$ [@Boettiger2018].
+The stochastic formulation follows directly from interpreting May's model as the average dynamics of a process subject to to environmental white noise proportional to the population size, $\xi_t X_t$, uncorrelated Gaussian noise $\xi \sim \mathcal{N}(0,\sigma)$ with mean zero and standard deviation $\sigma$ [@Boettiger2018].
 Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.
@@
 Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.
-By changing parameter $a$, the system goes from bistable (red, $a = 28$) to a ghost attractor (blue, $a = 27.2$).
+By changing parameter $a$, the system goes from bistable (blue, $a = 28$) to a ghost attractor (red, $a = 27.2$).
 For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}
@@
 the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.
-This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.
+This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].
 This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

### 2020-06-22 manuscript/manuscript.Rmd
WIP

@@
 \begin{equation}
-X_{t+1} = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{logistic growth}}  - 
+X_{t+1} = g(X_t) = X_t + \underbrace{X_t r \left(1 -\frac{X_t}{K} \right)}_{\textrm{logistic growth}}  - 
 \underbrace{\frac{a X_t ^ Q}{X_t^ Q + H ^ Q}}_{\textrm{saturating consumption}} + 
@@
 Model dynamics are illustrated in Fig 1 under two different parameter choices for the saturation response $a$: one which corresponds to a ghost attractor and one to a bistable system.
+
+We can extend this model to include the application of a control action in year $t$ of $A_t$ that decreases the population by a fixed amount before reproduction occurs, such that:
+
+\begin{equation}
+X_{t+1} = f(X_t, A_t) = g(X_t - A_t)
+\end{equation}
+
+Thus with $A_t = 0$ we can model the dynamics without control, and can model any control policy over time as a set of actions $A_t$ describing the beetle mortality imposed by management response that year.
 
@@
 Under appropriate choice of parameters, the model is bistable, with a low, "endemic" stable state, and a much higher stable "outbreak" state.
-In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emerging from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses.
+Population growth is subject to both the intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses.
+$t+1$ denotes the species generation time, which may be considerably shorter than an annual cycle.
 This saturating response and the resulting threshold dynamics in the response of affected trees exhibited by this model is quite well motivated [e.g.
@@
 This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].
-This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.
+A decrease in the parameter $a$ could be thought to model the weakened ability of the trees to respond to attack, arising from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.
 
-This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Fig 2.
+This same model can thus produce two very different realizations depending on the timing of the outbreak.
+It is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastic escape from a stable node, as illustrated in the ensemble simulations shown in Fig 2.
 Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B).
 Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.
-Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastic escape from a stable node.
 
@@
 
-$$\max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)$$
+\begin{equation}
+V_t(X_t) = \max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)
+\end{equation}
+
 where the evolution of $X_t$ is determined by Eq \eqref{model}.
 Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.
-By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985, @Marescot2013].
+By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985; @Marescot2013].
 

### 2020-06-23 manuscript/manuscript.Rmd
wip

@@
 This analysis also highlights the importance of stochasticity has for transient dynamics.
+
+Systems with alternative stable states [@Scheffer2001; @Carpenter] which can exhibit tipping points [@Scheffer2009; @Carpenter] are of great theoretical interest and pose unique challenges to ecological management [].
 
@@
 The transition matrix also depends on our knowledge or estimates of the parameters of \eqref{model}.
-For simplicity, the examples here have assumed a grid of state space values as he integers between 0 to 120, and an equivalently fine grid of 121 possible actions (control levels).
+For simplicity, the examples here have assumed a grid of state space values as the integers between 0 to 120, and an equivalently fine grid of 121 possible actions (control levels).
 This provides sufficient approximation to a continuous state space and action space, so that results are not sensitive to the details of this discritization.
@@
 In the scenarios we have just considered, the manager continues to apply the same optimal decision rule over the entire management period.
-This does not, of course, mean that the management response is completely static -- in fact, the manager selects a different the intervention level at each time-step in response to updated information about the current beetle population size -- but the manager does not use that information to revisit the underlying assumptions about the appropriate model.
+This does not, of course, mean that the management response is completely static -- in fact, the manager selects a different intervention level at each time-step in response to updated information about the current beetle population size -- but the manager does not use that information to revisit the underlying assumptions about the appropriate model.
 However, each additional observation potentially provides further information about which model is correct.
@@
 This represents a form of precautionary principle already suggested by the purely planning approach considered in the previous section -- the manager need not be certain about the correct model, but merely reach a certain threshold of probability.
+
+Note that in the management scenario stipulated, we have not assigned an intrinsic value to preserving beetles (or equivalently, a penalty for eliminating the beetle population entirely).
+Consequently, once the manager concludes the endemic state is no longer stable but is merely a transient created by the ghost attractor, elimination of the beetles is the best way to maximize the particular utility function chosen for this example.
+Given a different utility function, the outcome of the adaptive management process could be quite different.
+While our utlity function has been chosen for illustrative purposes, in practice a utility function would be determined through a delibrate process of stakeholder engagement.
 
@@
 For instance, very large environmental noise intensity ($\sigma$), erode any differences between the ghost and the weakly bistable system, since escapes from the low density, endemic state become commonplace.
+If noise is made much smaller than considered here, the systems are essentially deterministic, and thus would be trivial to distinguish.
+Rarely is reality so clean.
+On the other hand, if noise is much greater, neither the weak attractor or ghost attractor can constrain the dynamics very long, so the models would generate much the same outcome and there would be little management value in distinguish them.
 Likewise, extreme values of the costs of control or costs of tree damage can also make the decision problem trivial, where intervention is never worthwhile or always worthwhile.
@@
 In particular, we have assumed the manager observes the current population size without error before each action.
-In reality, any such observation would be accompanied by some uncertainty, making the decision problem into a Partially Observed Markov Decision Process (POMDP), increasing the complexity dramatically [e.g.
+In reality, any such observation would be accompanied by some uncertainty, making the decision problem into a Partially Observed Markov Decision Process (POMDP, [see @Smallwood1973; @Sondik1978]), increasing the complexity dramatically [e.g.
 @pomdp-intro].

### 2020-07-01 manuscript/manuscript.Rmd
wip

@@
 The optimal strategy determined by SDP can be represented as a "policy" function $p(X)$, indicating for each possible state $X$ what the optimal action $\hat A = p(X)$ would be.
-Fig 3 displays the numerically computed optimal policy for the system with only a transient.
+Fig 3 displays the numerically computed optimal policy for the system with only a transient compared to the optimal policy for the bistable system.
 Though the models differ by only a small amount in a single parameter, because one model corresponds to a system with only a ghost attractor while the other has a stable state attractor, the resulting optimal policy solutions differ significantly.
@@
 fixed_sims <- 
- bind_rows(ghost = sim_ghost, 
-           bistable = sim_bistable, 
+ bind_rows("simulations under ghost policy function" = sim_ghost, 
+           "simulations under bistable policy function" = sim_bistable, 
            .id = "prior") %>% 
@@
   geom_line(alpha=0.5) + 
-  facet_wrap(~prior, ncol=1)
+  facet_wrap(~prior, ncol=1) +
+  ylab("state or action")
 ```
@@
 # Fig 7
+learning <- list(df = read_csv("manuscript_cache/learning_df.csv"), 
+                 posterior = read_csv("manuscript_cache/learning_posterior.csv"))
  learning$df %>% 
@@
 ```{r figure8, fig.cap="Learning through Adaptive Management of the forest-beetle system.
-Bars show the probabilities that each model is correct over the first 10 steps of the simulated management process shown in Fig 7.
-Initial probabilities for each model are equal, and learning proceeds over the first 5 timesteps (shown in lighter shades in decreasing probability of the bistable model beign correct).
-By the fifth iteration, the manager has opted to eliminate the beetles and further learning is no longer possible in this system."}
+Initial probabilities for each model are equal, and learning proceeds over the next six timesteps (shown in lighter shades in decreasing probability of the bistable model beign correct).
+By the sixth iteration, the manager has opted to eliminate the beetles and further learning is no longer possible in this system."}
 model_names <- c("Ghost", "Bistable")
@@
   data.frame(time = 1:Tmax) %>%
-  filter(time %in% floor(seq(1,10, length.out=5))) %>%
-  gather(param, probability, -time, factor_key =TRUE) %>% 
-  mutate(param = model_names[as.integer(param)]) %>% 
-  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
-  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) + 
-  xlab("Model")
+  filter(time < 12) %>%
+  gather(model, probability, -time, factor_key =TRUE) %>% 
+  mutate(model = model_names[as.integer(model)]) %>% 
+  ggplot(aes(x = time, y = probability, col = model)) + geom_point() + 
+  geom_line()
+#  geom_bar(aes(x = param, y = probability, group = time, alpha = time), stat="identity", position = "identity", show.legend = FALSE, fill=pal[1]) +  xlab("Model")
 ```

### 2020-07-07 manuscript/manuscript.Rmd
first pass revisions done

@@
 
-Systems with alternative stable states [@Scheffer2001; @Carpenter] which can exhibit tipping points [@Scheffer2009; @Carpenter] are of great theoretical interest and pose unique challenges to ecological management [].
+Systems with alternative stable states which can exhibit critical transitions are of great theoretical interest [@Scheffer2001; @Ratajczak2018] and pose unique challenges to ecological management [@Folke2004; @Fischer2009; @Polasky2011].
+Most of this work has focused on critical transitions driven by saddle-node bifurcations [@Boettiger2013], which produce a blue-sky transition between alternative stable states and exhibit hysteresis which can frustrate efforts to restore the pre-transition state.
+Less attention has been given to the dynamics which immediately follow the saddle node bifurcation, in which the effects of the anhilated stable state may still be felt as a "Ghost Attractor," as introduced in the molecular and neuroscience literature [e.g.
+@Grossberg1980; @Deco2012], and more recently in the ecological literature [@Hastings2018].
+As such, despite the relative simplicity of the model, Ghost Attractor dynamics can provide important insights to our understanding of long transients, alternative stable state dynamics,and critical transitions, and the consequences such phenomena have for ecological management.
 
@@
 the vertical axis, Fig 1) is still small in magnitude, and thus dynamics near this region move slowly.
-This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossman1980; @Hastings2018].
+This is a simple example of a "Ghost Attractor" [@Deco2012] -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed [@Grossberg1980; @Hastings2018].
 A decrease in the parameter $a$ could be thought to model the weakened ability of the trees to respond to attack, arising from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.
@@
 Below an endemic population size $X_0$ I assume damage (e.g.
-no tree mortality) is minimal, and then increases quadratically with increasing beetle density; that is, the instantaneous utility of state with beetle density $X_t$ and taking action $A_t$ is
+no tree mortality) is minimal, and then increases quadratically with increasing beetle density; that is, the instantaneous utility of state with beetle density $X_t$ and taking action $A_t$ (the "utility function") is
 
@@
 
+Where `d` is the coefficient representing the damage (in terms of lost utility), and `c` is the coefficient representing the cost to providing control.
+Note that while it is common to express utility in strictly monetary terms (i.e.
+damage to the forest expressed in dollars lost, vs dollars spent to control the outbreak), it is by no means an assumption of the approach that utility must maximize monetary reward.
+In particular, different managers may assign different utility to either forest state or the control effort (i.e.
+differ in their values of `d` and `c`), and thus reach different management strategies, both of which are 'optimal' relative to their separate utilities.
+See @Walters1986, @Clark1990, @Fischer2009, and @Marescot2013 for further discussion of utility maximization in ecological management.
+
 The manager's objective is to select the sequence of actions $A_t \in \mathbf{A}$ which maximize the net expected discounted ($\delta$) utility,
@@
 \begin{equation}
-V_t(X_t) = \max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)
+V_t(X_t) = \max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t) \label{bellman}
 \end{equation}
@@
 Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.
-By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985; @Marescot2013].
+By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the Hamilton-Jacobi-Bellman recursion, Eq \eqref{bellman}, using Stochastic Dynamic Programming (SDP) [@Mangel1985; @Marescot2013].
 
@@
 ```{r figure3, fig.cap="Optimal policy functions under each model.
+A policy function indicates the optimal action (y-axis) that should be taken given each possible state (x-axis) of the system.
+The policy function arises as the solution to the Hamilton-Jacobi-Bellman recursion, Eq (4), e.g.
+as found by stochastic dynamic programming.
 Parameter values for utility function used here are $c = 1$, $d = 0.5$, $X_0 = 50$, $\\delta = 0.999$, parameters for the transition models are as given in Fig 1."}

### 2020-07-09 manuscript/manuscript.Rmd
resubmit

@@
 fixed_sims <- 
- bind_rows("simulations under ghost policy function" = sim_ghost, 
-           "simulations under bistable policy function" = sim_bistable, 
+ bind_rows("sims under ghost policy" = sim_ghost, 
+           "sims under bistable policy" = sim_bistable, 
            .id = "prior") %>% 
@@
   geom_ribbon(alpha = 0.5) +
-  ylab("Cost (net present value)") + xlab("time")
+  ylab("Cost (net present value)") +
+  xlab("time") +
+  theme(legend.position="bottom")
 ```
@@
 sims_by_prior <- 
- bind_rows(fifty_fifty = sim_fiftyfifty, 
-           seventy_thirty = sim_seventythirty, 
+ bind_rows("50-50 model prior" = sim_fiftyfifty, 
+           "70-30 model prior" = sim_seventythirty, 
            .id = "prior")  %>% 
